Cell crowding induces TRPV4 inhibition and its relocation to plasma membranes, implicating pro-invasive cell volume reduction mechanotransduction pathway

Version 1

Abstract

Abstract

Cell crowding is a common microenvironmental factor that affects various disease processes, but its impact on cell invasiveness into surrounding tissues is not well understood. In this study, we investigated the biomechanical changes induced by cell crowding, focusing on pro-invasive cell volume reduction. We discovered that cell crowding enhanced the invasiveness of high-grade ductal carcinoma in situ (DCIS) cells, which experienced significant cell volume reduction compared to hyperplasia-mimicking or normal cells. Mass spectrometry analyses revealed that cell crowding relocated ion channels, including TRPV4, a calcium-permeant ion channel, to the plasma membrane selectively in high-grade DCIS cells but not in less aggressive or normal cells. Cell crowding inhibited TRPV4 in high-grade DCIS cells, which led to decreased intracellular calcium levels and subsequent volume reduction. TRPV4 inhibition also prompted relocation of TRPV4 to the plasma membrane. This relocation primed inactive TRPV4 for activation, effectively counterbalancing the calcium loss from crowding-induced channel inhibition. Analyses of patient-derived breast cancer tissues validated that TRPV4 selectively associated with the plasma membrane in high-grade DCIS but not in lower-grade DCIS or less aggressive pathologies. The extent of plasma membrane TRPV4 association scaled with cell volume reduction and increased cell invasiveness and motility, suggesting its utility as an active pro-invasive mechanotransduction pathway indicator. Additionally, hyperosmotic conditions and pharmacologic TRPV4 inhibition mimicked the pro-invasive volume reduction observed under cell crowding, while TRPV4 activation reversed this effect by inducing cell volume increase. In summary, our study reveals a previously unrecognized pro-invasive mechanotransduction pathway triggered by cell crowding, which is selective in high-grade DCIS cells. This discovery offers new biophysical perspectives on cell invasiveness, highlighting the critical role of a selective mechanotransduction mechanism in the progression of breast cancer cells that are considered non-invasive but associated with high risk.

Introduction

The complex interplay between cellular mechanics and invasive behaviors is crucial to understanding various physiological and pathological processes, such as wound healing1 and disease progression2,3. Numerous studies show that these processes are mediated by mechanotransduction, whereby cells translate mechanical stimuli into cellular activity2,4–6. While mechanotransduction is traditionally studied concerning fluid stress7,8, matrix stiffness4,9, and other biomechanical changes such as osmotic stress10,11, the role of cell crowding, characterized by increased cell density and spatial constraints, is relatively less explored. A recent report shows that cell crowding plays a role in facilitating wound closure and repair by enhancing cell proliferation12. As tissues develop, repair, or undergo pathological transformation, cell crowding becomes common13,14. This challenges individual cells, forcing them to perceive and respond to the mechanical constraints of a crowded environment. Our study used human breast cell line model systems to describe a novel mechanotransduction pathway triggered by cell crowding that induces invasiveness into surrounding tissues. Interestingly, this pathway exhibited unique selectivity, as it specifically associated with a type of non-invasive cancer pathology and was not present in lower-grade or less aggressive pathologies. This suggests that not all cells possess the ability to translate mechanical stimuli, such as cell crowding, into cell invasiveness.

To assess the role of cell crowding in cell invasiveness, we chose in vitro cell lines linked to different pathological states that reflect cell crowding conditions in vivo, including atypical ductal hyperplasia (ADH)15 and ductal carcinoma in situ (DCIS)16–18. ADH is an intraductal clonal epithelial cell proliferative lesion18 and represents an intermediate step between normal breast tissue and in situ carcinomas19–21. ADH is associated with high risk as it is reported that an ADH diagnosis is associated with a five-fold increased risk of breast cancer15. DCIS is non-invasive form of cancer characterized by proliferating malignant epithelial cells16–18. Unlike ADH, DCIS is considered a precursor to invasive breast cancer22. However, the mechanism of how DCIS transitions to invasive cancer is not well understood and therefore, there is currently no reliable and robust method to differentiate which DCIS cells are at high risk of becoming invasive23,24.

Both ADH and DCIS conditions can potentially expose cells to crowding in vivo. However, how ADH and DCIS cells respond to such changes in cell density remains unknown. Our study revealed that cell crowding selectively triggered a pro-invasive mechanotransduction program in a specific type of DCIS cell line associated with high-grade comedo-type pathology25,26. The mechanotransduction program induced by cell crowding involved the inhibition of ion channels, such as transient receptor potential vanilloid 4 (TRPV4)27, a calcium-permeable ion channel, as identified from our mass spectrometry assay. This inhibition prompted the relocation of TRPV4 and other ion channels to the plasma membrane, leading to decreased intracellular calcium levels, subsequent induction of reduced cell volume and cortical stiffening, thereby promoting cell invasiveness.

We confirmed that inhibition of TRPV4 led to channel relocation to the plasma membrane in high-grade DCIS cells using TRPV4-specific pharmacologic inhibitors under normal cell density conditions without crowding. Pharmacologically activating TRPV4 in high-grade DCIS cells under cell crowding conditions resulted in a significant calcium spike, reflecting that the majority (>80%) of TRPV4 was located on the plasma membrane due to crowding. These data suggest that crowding-induced TRPV4 relocation to the plasma membrane via channel inhibition primed the inactive ion channel for activation, thereby counteracting decreased calcium levels. In contrast, activating TRPV4 under cell crowding conditions triggered channel uptake, indicating that TRPV4 trafficking depended on the channel’s activation status.

Remarkably, high-grade DCIS cells exhibited significant cell volume reduction under cell crowding conditions, indicating high cell volume plasticity. In contrast, ADH-mimicking cells showed largely unchanged cell volumes, reflecting low cell volume plasticity and a lack of mechanosensing capability. This high cell volume plasticity in high-grade DCIS cells was further evidenced by the dose-dependent cell volume changes observed with TRPV4 inhibitor and activator treatments, which correlated with the extent of plasma membrane relocation of TRPV4. Cell volume reduction induced by cell crowding or pharmacologic TRPV4 inhibition increased cell invasiveness and motility. This increase strongly correlated with the greater association of TRPV4 with the plasma membrane in high-grade DCIS cells, highlighting a marked selectivity for this cell type.

Further analysis of patient samples indeed confirmed that the selective relationship between plasma membrane TRPV4 association and pro-invasive cell volume reduction pathway was exclusive to high-grade DCIS cells, which was not present in lower-grade DCIS or less aggressive pathologies including ADH. This underscored the selectivity of this mechanotransduction pathway. Additionally, our findings revealed a dichotomy in the pro-invasive mechanosensing capabilities of invasive breast cancer cells, with some lacking these abilities. Notably, this selective pro-invasive mechanotransduction pathway was also activated by hyperosmotic stress, which led to ion channel relocation to the plasma membrane and osmotic cell volume reduction. The convergence of these endpoints resulted from distinct mechanical stimuli highlights the pivotal role of cell volume reduction in facilitating the invasion of high-grade DCIS cells, which are normally categorized as non-invasive but associated with high risk.

Results

Cell Crowding Enhances Invasiveness in High-Grade DCIS Cells

Cell crowding reflects a common condition in tumor microenvironments for ADH and DCIS, arising from aberrant cell proliferation within spatially constrained intraductal spaces. We examined the influence of this prevalent environmental factor on the invasiveness of these cells in vitro. To conduct this investigation, we assembled a panel of cell lines derived from the normal breast epithelial cell line MCF10A, including its H-RAS mutation-driven derivatives associated with various pathologies28: MCF10AT1, MCF10DCIS.com, and invasive cells MCF10CA1a. MCF10AT1 resembles ADH15, MCF10DCIS.com mimics high-grade DCIS26, and MCF10CA1a represents a malignant invasive cancer that was observed to form metastatic lesions in a mouse xenograft29. The current classification of DCIS relies on histological factors such as cell growth patterns and cytonuclear features30,31. Comedo-DCIS is a histologic subtype, which is characterized by apoptotic cell death, representing a high-grade DCIS with higher invasive potential than those of lower-grades25. MCF10DCIS.com forms comedo-type DCIS lesions that can spontaneously transition to invasive cancer when xenografted26.

To compare cell invasiveness under normal cell density and cell crowding conditions in vitro, we opted for a modified 2D matrix degradation assay. This approach allowed us to quantify overall cell invasiveness by normalizing it with the total cell number, thereby accounting for differences in cell densities. We chose this method over transwell, 3D Matrigel, or spheroid assays, where quantifying cell invasiveness as a function of cell density is challenging. By modifying an existing collagen-crosslinked polyacrylamide hydrogel matrix-based invasion assay32–34, we could determine the invasive cell fraction out of the total cell population by fluorescence imaging of the invasion gel bed and cell nuclei through a low magnification (4X) imaging (Fig. 1A; for a detailed procedure, see Suppl. Fig. 1) when cells were at normal density versus under cell crowding conditions.

Cell crowding selectively increases invasiveness in high-grade DCIS cells.

(A) We used a collagen-crosslinked polyacrylamide hydrogel matrix-based invasion assay to assess the effect of cell crowding on invasiveness. Representative images show gelatin-Alexa488 conjugates, where dark areas indicate cell invasion through degradation (green; Gelatin488), and DAPI staining marks cell locations (blue; DAPI) in a two-day invasion assay of MCF10DCIS.com cells under normal density (ND; upper panel) and overconfluent (OC; lower panel) conditions. “Masked” images are thresholded to produce positive masks applied to the “DAPI” images. Individual cell locations detected in “DAPI” images are marked with purple circles in “Detected points” images. The total number of cells within the field of view is counted from these points. By overlaying the mask and DAPI images, “Masked DAPI” images are obtained, and invaded cells are detected and represented by purple circles in “Masked points” images. The invasive cell fraction is calculated by the ratio of the number of invaded cells to the total number of cells (0.24 for ND and 0.59 for OC MCF10DCIS.com cells). These data show that cell invasiveness is enhanced by cell crowding. Scale bar = 100 μm. (B) Comparison of “Gelatin488” images of MCF10A (normal breast epithelial cells), MCF10AT1 (ADH-mimicking cells), MCF10DCIS.com (high-grade DCIS mimic), and MCF10CA1a (invasive breast cancer cells) between ND and OC conditions. MCF10DCIS.comcell invasion is significantly higher under cell crowding than under ND conditions. (C) Invasive cell fractions of these cells between ND and OC conditions are compared, showing that cell crowding-induced increases in invasiveness are notable only in MCF10DCIS.comcells. We used the two-tailed Mann-Whitney U test, a nonparametric and unpaired statistical method, to compare differences between groups. ****: p < 0.0001, ***: p < 0.001, **: p < 0.01, *: p < 0.05, ns: p > 0.05, throughout the manuscript.

We observed alterations in live cell morphology and invasiveness as they progressed from normal density (ND: 40-70%) to full confluence (100%) and beyond. Our goal was to determine a distinct time window, termed overconfluence (OC), where cell crowding-induced changes level off. The OC time window was achieved as cells cultured for an additional 5-10 days after reaching 100% confluence, with the growth medium replaced twice per day to prevent acidification. The schematic of the timeline to achieve the OC conditions, and low-magnification (4X) bright-field images at different time points from the ND to OC conditions are shown in Suppl. Fig. 2A-B. The time-dependent equilibration of cell invasiveness after the cells reached confluence is demonstrated in Suppl. Fig. 2C-D using our quantifiable collagen-crosslinked polyacrylamide hydrogel matrix-based invasion assay. Cell crowding mechanosensing significantly increased the invasive cell fraction of MCF10DCIS.comcells from ∼24% (ND cells) to 59% (OC cells). In contrast, other cell types such as normal breast epithelial cells MCF10A and MCF10AT1, which mimic ADH conditions in vivo, did not display invasiveness in ND conditions or enhanced invasiveness in OC conditions, suggesting insensitivity to cell crowding (Fig. 1B-C).

The invasive cell fraction of ND MCF10CA1a cells was already ∼80%, and the additional increase in invasiveness due to cell crowding was not readily discernible, with a slight increase to ∼82% under crowded conditions. These data suggest a striking and selective mechanosensing effect of cell crowding on cell invasiveness in MCF10DCIS.comcells. To ensure that acidification did not affect the invasiveness of MCF10DCIS.com cells despite the frequent replacement of the cell growth medium, we incubated ND cells for two days with acidified medium from cultures of OC MCF10DCIS.comcells. We observed that medium acidity did not alter cell invasiveness (Suppl. Fig. 2E), reinforcing that the increased invasiveness under OC conditions was induced by cell crowding.

Cell crowding reduces cell volume and stiffens MCF10DCIS.comcells

As cells became crowded, we observed a reduction in cell size. Previous reports indicate an inverse relationship between cell volume and cortical stiffness35,36, leading us to hypothesize that reduced cell volume would be accompanied by increased cortical stiffness. Research on glioma cell invasion underscores the critical role of hydrodynamic cell volume changes in penetration into surroundings37,38, suggesting that significant cell volume reduction facilitates cell invasion. Additionally, increased cortical stiffness is known to help cells overcome the physical barriers of the dense extracellular matrix39,40. We thus investigated whether cell crowding selectively induces pro-invasive cell volume reduction and cell stiffening in MCF10DCIS.comcells, thereby priming them for invasion.

To measure the cell volume of individual cells, we used confocal microscopy with a 60X oil immersion objective to obtain z-stack images of live or fixed cells stably expressing red fluorescent protein (RFP)41. Using a nanoindenter attached to the confocal microscope (Suppl. Fig. 3A), we extracted Young’s modulus of live individual cells to assess changes in cortical stiffness. Cell crowding significantly reduced cell volume (Fig. 2A, B) and increased cortical stiffness (Fig. 2C) in both MCF10DCIS.comand MCF10A cells. However, the changes in volume and stiffness between ND and OC conditions were more pronounced in MCF10DCIS.comcells (Suppl. Fig. 3B, C). The cell crowding-induced alterations in cell volume and stiffness were negligible in MCF10AT1 cells (Fig. 2A-C). In MCF10CA1a cells, cell crowding led to an increase in stiffness, but a decrease in volume was not evident. This was expected because the cell volume was already too small to be accurately measured with the available resolution (Fig. 2A-C). The notable plasticity in cell volume and stiffness changes observed in MCF10DCIS.comcells in response to cell crowding potentially underscores the critical link between this plasticity and mechanosensitive increases in cell invasiveness. We assessed cell volume changes only as an effector event of cell crowding, without measuring cell stiffness, because cell volume reflects the mechanical properties of the entire cell, while Young’s modulus can vary depending on the location of indentations in the plasma membrane42,43. The observed cell volume changes by cell crowding (Fig. 2A) showed the same trend in the ND cell volume, with final volumes under crowded conditions being comparable across all cells. Consequently, the ND cell volume and the volume changes from ND to OC conditions exhibited an approximately linear relationship, with an R² ∼ 0.97 (Suppl. Fig. 3D), indicating a highly linear correlation. This finding suggests that the cell volume at ND could serve as an indicator of relative cell volume plasticity.

Cell Crowding induces significant cell volume reduction and stiffening in High-Grade DCIS Cells.

(A) Cell volume (V; mean and SD) differences between ND and OC conditions of MCF10A, MCF10AT1, MCF10DCIS.co, and MCF10CA1a cells are plotted. The high-grade DCIS cell mimic, MCF10DCIS.co, shows a large volume reduction due to cell crowding. (B) Representative confocal microscopy images of RFP-coexpressing cells of the four cell types in ND and OC conditions. The images include x-y (left) and x-z (right) views, with scale bar = 10 μm. The large volume reduction of MCF10DCIS.comcells is evident. (C) Plots showing changes in cortical stiffness (mean and SD) measured by Young’s modulus (Y) using a nanoindenter, displaying significant cell stiffening of MCF10DCIS.comcells due to cell crowding. (D) Hyperosmotic conditions induced by PEG 300 treatment (2% = 74.4; 4% = 148.8 mOsm/kg) lead to dose-dependent cell volume reduction. (E) Treatment with 2% PEG 300 for two days significantly increased the invasiveness (mean and SD) of MCF10DCIS.comcells. ****: p < 0.0001, ***: p < 0.001, **: p < 0.01, *: p < 0.05, ns: p > 0.05.

To validate the pro-invasive nature of cell volume reduction in MCF10DCIS.comcells, we examined the effect of hyperosmotic cell volume reduction in ND cells using polyethylene glycol (PEG) 300 for 15 minutes. The cell volume reduction effects of this treatment were previously described36,44. PEG 300 effectively reduced ND MCF10DCIS.com cell volume in a dose-dependent manner, showing a greater reduction with 148.8 mOsmol/kg PEG 300 than with 74.4 mOsmol/kg PEG 300 (Fig. 2D). To assess the impact of PEG-induced cell volume reduction on cell invasiveness, we used collagen-crosslinked polyacrylamide hydrogel matrix-based invasion assay and exposed cells to 74.4 mOsmol/kg PEG 300 for 2 days. While treatment with 148.8 mOsmol/kg PEG 300 significantly reduced cell viability, cells treated with 74.4 mOsmol/kg PEG 300 remained viable for 2 days. Monitoring changes in cell invasiveness revealed that exposure of ND MCF10DCIS.com cells to 74.4 mOsmol/kg PEG 300 increased cell invasiveness (Fig. 2E), confirming the causal relationship between cell volume reduction and increased cell invasiveness.

Cell crowding induces TRPV4 relocation to plasma membrane in MCF10DCIS.com cells

Cell volume regulation typically depends on osmotic gradients that direct the movement of water across cell membranes45. This process is facilitated by ion flux modulation, which is controlled by ion channels and ion transporters located on the plasma membrane46–49. Given the observation that cells capable of attaining minimal cell volume can successfully penetrate neighboring tissues38, we speculated that the high cell volume plasticity observed in MCF10DCIS.com cells reflects this ability. For such efficient cell volume changes, we hypothesized that cell crowding mechanosensing in MCF10DCIS.com cells would alter the number of plasma membrane-associated ion channels or transporters.

To compare the plasma membrane associated proteins between ND and OC conditions, we employed mass spectrometry and profiled those proteins using the streptavidin-pulled surface-biotinylated cell lysates. Fig. 3A depicts the relative protein densities on the plasma membrane between the ND (blue bars) and OC (red bars) conditions for the top 25 proteins showing over 5-fold increases (triangle plots) in protein expression on the plasma membrane in the OC condition compared to the ND condition in MCF10DCIS.com cells. The gene names of these 25 proteins are presented in Fig. 3A (Suppl. Table 1 provides the corresponding protein names along with the corresponding fold increases). Notably, TRPV4, a member of the transient receptor potential family of ion channels known for its mechanosensitive properties50,51, showed a remarkable 153-fold increase in plasma membrane association in response to cell crowding, as illustrated in Fig. 3A. Additionally, we observed an increase in plasma membrane association of other ion channels, such as SCN11A (the alpha subunit of the voltage-gated sodium channel Nav1.9) with ∼42-fold enrichment, and KCNN4 (the small-conductance calcium-activated potassium channel 3 SK3) showing ∼6-fold increase, as shown in Fig. 3A. To evaluate the relative plasma membrane association of proteins under OC versus ND conditions, we performed a comparable mass spectrometry analysis of cell-surface biotinylated cell lysates for MCF10A, MCF10AT1, and MCF10CA1a cells. The gene/protein names showing increased plasma membrane associations in these cells are listed in Suppl. Table 2. Unlike in MCF10DCIS.com cases, there was no cell-crowding-induced relocation of ion channels and ion transporters to the plasma membrane in MCF10A and MCF10AT1 cells. However, in MCF10CA1a cells, ion transporters such as ATP2B4 showed a ∼37-fold greater plasma membrane association under crowding conditions. This suggests that plasma membrane relocations of these ion channels and ion transporters in response to cell crowding selectively occurred in MCF10DCIS.com cells and, to an extent, in invasive ductal cancer MCF10CA1a cells.

Cell crowding induces TRPV4 relocation to plasma membrane in MCF10DCIS.com cells.

(A) Mass spectrometry data showing proteins enriched in the plasma membrane (PM) >5-fold (fold changes represented using triangle plots; OC/ND ratio on the right axis) when cells are under OC (orange bars) relative to ND conditions (blue bars). Ion channels are marked with red boxes, where TRPV4 shows about a 160-fold increased association with the plasma membrane under OC conditions. (B) Proteins near and on the plasma membrane were pulled down after cell surface biotinylation with streptavidin beads and immunoblotted for TRPV4. TRPV4 is significantly associated with the plasma membrane in OC MCF10DCIS.comcells. In MCF10CA1a cells, TRPV4 appears to be associated with the plasma membrane under both ND and OC conditions, with a slight increase under OC conditions. (C) Immunoblots of whole-protein lysates demonstrate similar overall TRPV4 protein levels across MCF10A cell derivatives, regardless of cell density. This indicates that the differing plasma membrane association of TRPV4 is due to trafficking changes, not expression level changes. GAPDH is used as a loading control. (D) Representative IF images by confocal microscopy show TRPV4 (red) localization compared to the control protein transferrin receptor (TfR; green) in MCF10A, MCF10AT1, MCF10DCIS.co, and MCF10CA1a cells under ND and OC conditions. DAPI (blue) staining was used for visualizing the nuclei. As observed in the biochemical data in (B-C), cell crowding induces the relocation of TRPV4 to the plasma membrane in MCF10DCIS.comcells. TRPV4 is associated with the plasma membrane in ND MCF10CA1a cells, with a clear elevated association in OC cells. Scale bar = 10 μm. (E) Plasma membrane-associated TRPV4 (%) is quantified for the four cell lines under ND and OC conditions by line analysis, showing a significant increase in both MCF10DCIS.comcells and MCF10CA1a cells due to cell crowding. (F) IF images showing that hyperosmotic conditions induced by PEG 300 (74.4 mOsm/Kg) treatment also relocate TRPV4 (red) to the plasma membrane in MCF10DCIS.comcells. TfR localization remains intact under hyperosmotic conditions. Increased relocation is also observed in MCF10CA1a cells. Scale bar = 10 μm. (G) The increased plasma membrane association of TRPV4 due to hyperosmotic stress is quantified by line analysis. Scale bar = 10 μm. (H) Representative regions of interest (ROIs) of TRPV4-stained immunohistochemistry (IHC) images in different pathology phenotypes. High-grade DCIS and invasive ductal cancer (IDC) ROIs clearly exhibit plasma membrane association of TRPV4. Two high-grade DCIS IHC images were acquired by two different people and both show plasma membrane-associated TRPV4. Scale bar = 20 μm. (I) Statistical results from independent histological evaluations of pathologies and TRPV4 distributions of 97 ROIs from 39 patient specimens indicate a high correlation (>70%) of plasma membrane association of TRPV4 with high-grade DCIS or IDC pathologies. Y/N: Yes/no, indicating both pathologists agreed that PM ion channels were present/absent. E: Equivocal, indicating only one pathologist agreed. Significantly high proportions of high-grade DCIS (75%) and IDC (73%) ROIs exhibited plasma membrane TRPV4 association, which was not observed in lower-risk cases.

TRPV4 plays a pivotal role in facilitating the passage of calcium ions (Ca2+) and is integral in detecting various forms of mechanical stresses, including temperature fluctuations and osmotic pressure52,53. Activation of TRPV4 channels elevates intracellular calcium levels, leading to cell volume increase via osmotic water influx52,54. Nav1.9 primarily contributes to the generation and propagation of action potentials in sensory neurons and is associated with various pain disorders55–57. SK3 channels are activated by elevated intracellular calcium levels and serve diverse physiological functions58,59. However, unlike TRPV4, Nav1.9 and SK3 channels are not generally classified as mechanosensors. Consequently, we focused on TRPV4, the mechanosensor channel displaying the most significant increase in the plasma membrane association under cell crowding conditions, to further investigate its role in promoting pro-invasive cell volume reduction in MCF10DCIS.comcells.

To confirm the selective plasma membrane association of TRPV4 in OC MCF10DCIS.comcells suggested by mass spectrometry results, we conducted immunoblots on surface-biotinylated lysates from all four cell types under both conditions. Fig. 3B and Suppl. Fig. 4A showed a notable association of TRPV4 with the plasma membrane in OC MCF10DCIS.comcells, which was absent in ND cells. This association was not observed in MCF10A and MCF10AT1 cells under any conditions. Interestingly, in MCF10CA1a cells, TRPV4 associated with the plasma membrane under both conditions, with a slight increase in OC conditions. This increase suggests that these cells may also possess mechanosensing abilities, enabling them to respond to cell crowding. Additionally, immunoblots from whole-cell lysates showed consistent TRPV4 expression levels across different cell types and densities (Fig. 3C, Suppl. Fig. 4B), indicating that cell crowding influences ion channel trafficking without altering overall expression. This evidence strongly suggests that the relocation of TRPV4 to the plasma membrane is a mechanosensitive response to cell crowding.

To investigate the redistribution of TRPV4 in response to cell crowding, we performed IF imaging with confocal microscopy across all four cell types under both ND and OC conditions. The binding specificity of the TRPV4 antibody was confirmed through IF imaging (Suppl. Fig. 5A) and immunoblotting (Suppl. Fig. 5B) of MCF10DCIS.comcells treated with TRPV4 shRNA. As anticipated from the mass spectrometry results, TRPV4 (red) was significantly associated with the plasma membrane, selectively observed in OC MCF10DCIS.comcells in the IF images shown in Fig. 3D. In MCF10CA1a cells, TRPV4 was already associated with plasma membrane under ND conditions, but cell crowding increased the association significantly. Plasma membrane associations were verified using the plasma membrane marker DiIC18(3), which we previously used to visualize the plasma membrane60. IF images showed that TRPV4 successfully colocalized with DiIC18(3) in OC MCF10DCIS.comand OC MCF10CA1a cells (Suppl. Fig. 6A). However, the non-interacting protein transferrin receptor (TfR; green) remained similarly distributed in the OC cells compared to the ND cells, indicating that the protein trafficking changes induced by cell crowding are specific to TRPV4 (Fig. 3D).

To investigate whether other ion channels identified through mass spectrometry and a well-known mechanosensory channel PIEZO161,62, exhibit similar behavior, we conducted IF imaging on ND and OC MCF10DCIS.comcells targeting KCNN4 and PIEZO1. Unfortunately, due to the absence of specific antibodies, we were unable to examine SCN11. In ND MCF10DCIS.comcells, KCNN4 was observed to be mostly cytosolic, while PIEZO1 was mildly but slightly more associated with the plasma membrane than TRPV4 under ND conditions. However, under OC conditions, both PIEZO1 and KCNN4 also relocated to the plasma membrane, albeit to a lesser extent than TRPV4 (Suppl. Fig. 6B). The varying degrees of plasma membrane association of PIEZO1 and KCNN4 are quantified using line analysis (Supple. Fig. 6C).

As expected, in ND MCF10A, MCF10AT1, and MCF10DCIS.comcells, TRPV4 was predominantly intracellular, with notable enrichment in the nuclei (Fig. 3D). However, in OC MCF10DCIS.comcells, approximately 80% of TRPV4 was located in the plasma membrane based on line analysis (Fig. 3E). In contrast, in OC MCF10A and OC MCF10AT1 cells, TRPV4 remained cytosolic but migrated out of the nuclei (Fig. 3D). TRPV4 association with the plasma membrane was observed in both ND and OC MCF10CA1a cells, with an observable increase under OC conditions, further confirming the immunoprecipitation results shown in Fig. 3B. Plasma membrane-associated TRPV4 was quantified using line analysis and plotted in Fig. 3E. Refer to Suppl. Fig. 7A for plots of the relative TRPV4 associations with the plasma membrane, cytosol, and nucleus between ND and OC conditions in all four cell types. These IF imaging results underscore the mechanosensitive plasma membrane relocation of ion channels in MCF10DCIS.co, and to an extent, in MCF10CA1a cells, contrasting with the observed insensitivity of MCF10A and MCF10AT1 cells to cell crowding.

To assess the relationship between cell crowding-induced cell volume reduction and plasma membrane relocation of TRPV4, we investigated whether hyperosmotic conditions, which reduce cell volume without cell crowding, would also result in TRPV4 relocation. We subjected ND cells to the same PEG 300 condition (74.4 mOsm/kg PEG for 15 minutes) used in Fig. 2D to induce cell volume reduction. In ND MCF10DCIS.comcells, such hyperosmotic cell volume reduction prompted plasma membrane relocation of TRPV4, a response not observed in MCF10A and MCF10AT1 cells, but seen in MCF10CA1a cells (Fig. 3F). Quantitative line analysis confirmed these findings by assessing the relative TRPV4 association with the plasma membrane (Fig. 3G; Suppl. Fig. 7B for plots of the relative TRPV4 association with the plasma membrane, cytosol, or nucleus). This finding underscores the relationship between mechanosensitive cell volume reduction and plasma membrane relocation of TRPV4, highlighting that both cell crowding and hyperosmotic stress lead to the same effect, particularly pronounced in MCF10DCIS.comcells.

Patient tissue analysis shows selective plasma membrane association of TRPV4 in high-grade DCIS cells

MCF10DCIS.comcells represent a basal cell model for high-grade DCIS cells driven by HRAS mutation26,63. However, the majority of patient-derived DCIS cells originate from the luminal cell population and lack the HRAS mutation63. Moreover, there are limited options for patient-derived DCIS cell line models that have not linked to well-defined pathology64,65. Thus, to seek generality of our finding about the selective association between plasma membrane relocation and high-grade DCIS pathology, we examined 39 breast cancer patient tissue blocks to assess the selectivity in vivo. We designed an early-stage retrospective study by selecting various breast tissue pathologies that range from benign to invasive cancer. Those pathologies include usual ductal hyperplasia, papilloma, columnar changes, ADH, low-to high-grade DCIS, and invasive ductal carcinoma (IDC). We also incorporated normal regions for comparison. Hematoxylin and eosin (H&E) staining was used to visualize tissue sections from each patient, excluding samples with prior cancer diagnoses or drug treatments. In total, 97 regions of interest (ROIs) from H&E-stained sections of 39 patient tissue blocks were selected. Two pathologists independently assessed TRPV4 distribution patterns at the single-cell level in corresponding ROIs in the immunohistochemistry (IHC) images (Fig. 3H). A detailed methodology is described in the Methods section and more example images are shown in Suppl. Fig. 8. High-grade DCIS cells, as depicted in Fig. 3H, clearly demonstrated plasma membrane-associated TRPV4, a feature absent in lower-grade DCIS cells (intermediate- and low-grade). This distinction sets them apart not only from cells with ADH and benign pathologies but also from lower-grade DCIS cells. As demonstrated by the IF images in Fig. 3D for MCF10CA1a (invasive cell mimic), IDC cells exhibited notable plasma membrane TRPV4, suggesting that they possess a similar pro-invasive mechanotransduction capability to high-grade DCIS cells.

We found that the sensitivity and specificity for plasma membrane TRPV4 in high-grade DCIS cells were 0.75 ± 0.19 (15/20) and 0.98 ± 0.03 (61/62) respectively (95% confidence interval) (Fig. 3I). It is important to note that the specificity calculation excluded IDC cases. Notably, even in less aggressive pathologies, a significant amount of TRPV4 was localized in the nuclei, as shown for MCF10A and MCF10AT1 cells in the IF images in Fig. 3D. This underscores our interpretation from the in vitro results in Fig. 3A-D that increased plasma membrane association of TRPV4 in high-grade DCIS cells results from changes in protein localization through trafficking alterations, rather than differences in expression levels. These initial in vivo results clearly demonstrate the selective and specific association of TRPV4 with the plasma membrane in high-grade DCIS cells. Notably, this association is absent in lower-grade DCIS, ADH, benign, and normal cells, thereby confirming our in vitro findings using MCF10A cell derivatives. These results suggest a potentially critical role for TRPV4 in the progression of high-grade DCIS.

Cell crowding inhibits ion channels and triggers their plasma membrane relocations

A calcium-permeant ion channel like TRPV4 is known to influence intracellular calcium dynamics54. While activation of TRPV4 typically elevates calcium levels and this potentially increases cell volume66–68, the impact of its inhibition is less clear, given the multifaceted nature of calcium signaling in cell volume control69. This complexity is compounded by compensatory cellular mechanisms and the involvement of other ion channels in response to altered calcium homeostasis46. Considering these factors, we hypothesized that cell crowding might inhibit calcium-permeant ion channels, including TRPV4, which would then lower intracellular calcium levels and subsequently reduce cell volume via osmotic water movement. To test this, we employed the Fluo-4 calcium assay70 to compare relative intracellular calcium levels of MCF10DCIS.comcells between ND and confluent (Con) conditions through 4X confocal microscopy imaging with 488 nm excitation. For this calcium assay, we opted for confluent conditions instead of OC conditions to collect the Fluo-4 signal from confluent monolayers of MCF10DCIS cells, which demonstrates comparable cell-crowding induced TRPV4 relocation to the plasma membrane (Suppl. Fig. 9A) as observed in the OC condition. This choice was made because more than one layer of cells was occasionally observed in OC conditions, which would yield overestimation of crowding-dependent intracellular calcium levels. The intracellular Fluo-4 signal under Con conditions was significantly lower than that under ND conditions, as shown in the fluorescence images of ND versus Con cells in Fig. 4A. Time-dependent Fluo-4 images were acquired over 25 minutes. For ND cell images, the average Fluo-4 intensity values were calculated from 10-15 selected cells (highlighted in the white box in Fig. 4A). For Con cell images, intensity values were derived from all cells within the entire 50 μm by 50 μm field of view, as depicted in Fig. 4A. After background subtraction, the intensity values were plotted in Fig. 4B. The temporal Fluo-4 intensity profiles from both ND and Con cells remained largely constant over the measured 20-minute period. However, Con cells exhibited significantly lower Fluo-4 intensity compared to ND cells, indicating reduced intracellular calcium levels. Lower Fluo-4 intensity in Con cells than in ND cells was not due to limited Fluo-4 reagent in Con samples. Within cell clusters, cells experiencing crowding exhibited notably lower Fluo-4 levels compared to those at the periphery (Suppl. Fig. 9B), suggesting that cell crowding had resulted in a decrease in intracellular calcium levels, likely due to the mechanosensitive inhibition of calcium-permeable channels like TRPV4.

Cell crowding inhibits ion channels and triggers their plasma membrane relocations.

(A) To compare intracellular calcium (Ca2+) levels, we used a Fluo-4 AM assay, where green fluorescence intensity increases with higher Ca2+ levels. Calcium levels are significantly lower in Con MCF10DCIS.comcells, which form a monolayer immediately after reaching full confluence. (B) The temporal progression of averaged Fluo-4 intensity in ND MCF10DCIS.comcells (blue curve) in the box shown on the left image is compared with that of Con cells (red curve). Fluo-4 intensity is consistently lower in Con cells than in ND cells for approximately 25 minutes (Acquisition time: 5 s). (C) Time- and cell number-averaged Fluo-4 intensity reduction due to cell crowding is significant in MCF10DCIS.comcells. (D-H) Pharmacological inhibition of TRPV4 with 1 nM GSK219 generates dips in the Fluo-4 signal. Fluo-4 images at the baseline (t1) and the dip (t2) post 1 nM GSK219 are compared in MCF10DCIS.comcells between ND (D) and Con (F) conditions. The Fluo-4 intensity time traces are compared between ND (E) and Con (G) conditions, where the magnitude of the dip (marked with ΔCa) is significantly lower in Con cells, indicating cell crowding-induced TRPV4 inhibition. Using a lower GSK219 concentration (0.2 nM: 219L), we found that ΔCa increases with higher doses (1 nM: 219H), while the overall ΔCa magnitudes are significantly lower in Con MCF10DCIS.comcells. (I-M) TRPV4 activation with 0.2 pM GSK101 leads to a small spike in ND cells (I, J). However, in Con cells, the same GSK101 treatment leads to a notable spike in Fluo-4 intensity, indicating that TRPV4 inhibition and subsequent relocation to the plasma membrane by cell crowding primes the ion channels for activation. GSK101 treatment also leads to a dose-dependent increase in the spike magnitude with higher GSK101 concentrations (0.05 pM: 101L; 0.2 pM: 101H), which is strikingly high in Con MCF10DCIS.comcells. (N-Q) TRPV4 activation status-dependent intracellular localization changes. (N) IF images of TRPV4 (red) and TfR (green) in ND MCF10DCIS.comcells show that GSK101 does not increase plasma membrane association of TRPV4. However, GSK219 significantly relocates TRPV4 to the plasma membrane in a dose-dependent manner, similar to ND cells treated with 74.4 mOsm/Kg PEG 300. (O) In OC cells, while GSK219 does not significantly alter TRPV4 association with the plasma membrane, GSK101 depletes plasma membrane TRPV4 in a dose-dependent manner, suggesting that TRPV4 activation status affects its trafficking. Relative plasma membrane associations with different treatments are quantified for ND (P) and OC (Q) cells using line analysis. (R) The magnitudes of Fluo-4 spikes by GSK101 and dips by GSK219 show a linear relationship (R² ∼ 0.69), indicating a negative correlation between them. This reinforces the observation that TRPV4 inhibition increases its association with the plasma membrane, while activation shows the reverse effect.

To investigate whether TRPV4 inhibition was involved in decreased intracellular calcium levels and subsequent relocation of TRPV4 to the plasma membrane, we employed TRPV4-specific pharmacological agents to alter TRPV4 activity and modulate calcium concentrations in ND MCF10DCIS.comcells without applying cell crowding conditions. To inhibit TRPV4 activity, we used TRPV4 inhibitor (GSK2193874; GSK219)71 at 0.2 and 1 nM, which we confirmed to have insignificant effects on cell viability in both ND and OC conditions for two days (Suppl. Fig. 10A). This allowed us to use these conditions for the invasion assays described below.

GSK219 treatments immediately reduced the Fluo-4 signal, followed by a recovery, indicating cellular homeostatic activity. Fig. 4D shows Fluo-4 images before and at the signal dip after applying 1 nM GSK219. Fig. 4E presents the temporal profile of Fluo-4 intensity, marking two time points: t1 (baseline) and t2 (dip). Although the Fluo-4 dip at t2 was not visibly apparent in the image of ND MCF10DCIS.comcells (Fig. 4D), the temporal profile (Fig. 4E) clearly showed TRPV4 inhibition by GSK219. The extent of inhibition was reflected in the calcium depletion, ΔCa (vertical dashed line between the dip and baseline), followed by homeostatic recovery.

In Con cells, 1 nM GSK219 led to a very small dip in Fluo-4 intensity with no changes afterward, likely due to the already low intracellular calcium levels caused by cell crowding. Fig. 4F shows the baseline and dip Fluo-4 images, while Fig. 4G presents the temporal Fluo-4 intensity profile, highlighting the two time points. The reduction in Fluo-4 intensity induced by GSK219 was smaller in Con MCF10DCIS.comcells (mean ΔCa ∼ −39 ± 9) compared to ND cells (mean ΔCa ∼ −49 ± 18), as illustrated in Fig. 4H. This smaller reduction in calcium in Con cells may also be due to less active TRPV4 channels, resulting from cell crowding-induced TRPV4 inhibition. The dose-dependent (0.2 nM and 1 nM GSK219, labeled 219L and 219H) and cell density-dependent (ND and Con) Fluo-4 signal reductions were reflected in the ΔCa values plotted in Fig. 4H. The results show the most significant TRPV4 inhibition in ND MCF10DCIS.comcells at 1 nM GSK219. These data suggest that a significant portion of TRPV4 that relocated to the plasma membrane in response to cell crowding was in an inactive state.

Administering a TRPV4 activator (GSK1016790A; GSK101)72,73 immediately caused a notably greater spike in Fluo-4 signal in Con cells than in ND cells, evidencing that a large number of inactive TRPV4 channels were indeed associated with the plasma membrane in Con cells. In Con cells, 0.2 pM GSK101 led to a Fluo-4 signal spike with ΔCa ∼ +2,348 ± 597, while in ND cells, ΔCa was ∼ +324 ± 130. This difference is clearly depicted in the Fluo-4 images (Fig. 4I for ND and Fig. 4K for Con) and the time-dependent Fluo-4 intensity plots (Fig. 4J for ND and Fig. 4L for Con). The dose (101L and 101H representing 0.05 pM and 0.2 pM GSK101) and cell-density (ND and Con) dependent Fluo-4 signal spikes are reflected in the ΔCa values, as shown in Fig. 4M. The results show the most significant TRPV4 activation in Con MCF10DCIS.comcells at 0.2 pM GSK101, confirming a large number of primed inactive TRPV4 channels at the plasma membrane in Con cell.

Based on the differing effects of GSK219 and GSK101 on immediate calcium responses, we hypothesized that cell crowding induced TRPV4 inhibition and subsequent calcium depletion, which triggered the plasma membrane relocation of TRPV4 to compensate for the loss of calcium. We examined GSK219 and GSK101 dose-dependent relocation in both ND and OC MCF10DCIS.comcells via IF imaging (Fig. 4N for higher GSK101/219 concentrations and Suppl. Fig. 10B for both concentrations). Treating ND cells with GSK219 for 1 hour confirmed plasma membrane relocation of TRPV4 (∼53 ± 7%) at 0.2 nM and more (∼62 ± 7%) at 1 nM, significantly higher than in control ND cells (∼15 ± 6%), as shown by the line analysis (Fig. 4P). These data indicate that TRPV4 inhibition triggered its strategic relocation to the plasma membrane, preparing it for rapid activation to offset the channel inhibition effects.

Following the trend of GSK101-induced small Fluo-4 spikes in ND MCF10DCIS.comcells (Fig. 4J), the plasma membrane associated TRPV4 remained similarly with GSK101 to the control condition (Fig. 4N, 4P). However, in Con cells, GSK101 diminished the plasma membrane associated TRPV4 channels in a dose-dependent manner through internalization (Fig. 4O, 4Q), consistent with previous findings65. These data suggest that plasma membrane TRPV4 levels were largely regulated by the channel activity status. Specifically, channel activation led to the internalization of TRPV4, while channel inhibition promoted the relocation of TRPV4 to the plasma membrane. However, we observed that GSK219 treatment in Con cells results in a slight reduction of TRPV4 from the plasma membrane, rather than an expected increase via additional plasma membrane relocation (Fig. 4O, 4Q). Considering that nuclear TRPV4 levels increase with higher doses of GSK219 in Con cells (Suppl. Fig. 10B), as opposed to the cytoplasmic TRPV4 levels, we postulate that a distinct trafficking alteration occurs when TRPV4 is further inhibited under crowding conditions, where the channels are largely inactive.

Given the role of TRPV4 in osmoregulation53,74 and its significant relocation to the plasma membrane under hyperosmotic conditions (∼55% plasma membrane association by line analysis; Fig. 3E, 4P) in ND MCF10DCIS.comcells, we explored the potential involvement of TRPV4 inhibition during this relocation. Using the Fluo-4 assay, we observed an initial calcium spike following 74.4 mOsmol/kg PEG 300 in ND MCF10DCIS.comcells due to osmotic water outflow, which was followed by a homeostatic relaxation aimed at restoring calcium levels (Suppl. Fig. 11A). This relaxation likely involved the inhibition of ion channels like TRPV4 and other ion transporters, followed by their relocations to the plasma membrane. Indeed, we confirmed the relocation of other ion channels that responded to cell crowding, including KCNN4 and PIEZO1, to the plasma membrane following the same PEG 300 treatment (Suppl. Fig. 11B for IF images and line analysis results). The data suggest that the increased association of TRPV4 with the plasma membrane under hyperosmotic stress may reflect an adaptive inhibition response to water outflow.

Cell crowding-induced inactive TRPV4 relocation to the plasma membrane indicates inhibition-mediated volume reduction and enhanced invasiveness and motility

We next investigated whether the inhibition-driven plasma membrane relocation of inactive TRPV4 precedes the reduction in cell volume and the enhancement of cell invasiveness observed in response to cell crowding (as shown in Fig. 2A, 2B). To explore these relationships, we examined the effects of 2-day treatments with GSK101 and GSK219 on the cell volumes of ND and OC MCF10DCIS.comcells. This duration was selected to correlate the treatments with changes in cell volume and invasiveness.

Under ND conditions, activating TRPV4 with GSK101 (0.05 or 0.2 pM for 2 days), which caused minor calcium spikes and left TRPV4 distribution largely unaltered (Fig. 3A; Suppl. Fig. 6A), did not significantly alter cell volume (Fig. 5A). Conversely, TRPV4 inhibition by GSK219 treatment (1 nM for 2 days) that induced greater calcium changes (dips rather than spikes) and a significant plasma membrane relocation of TRPV4, led to a noticeable cell volume reduction (Fig. 5A). 0.2 nM GSK219 treatment had a negligible impact on ND cell volume (Fig. 5A). The GSK219 effect in ND MCF10DCIS.comcells was similar to that observed under hyperosmotic conditions by 74.7 mOsm/Kg PEG 300 (Fig. 5B). In contrast, under OC conditions, GSK101, which triggered dose-dependent calcium spikes, led to corresponding increases in cell volume. In contrast, GSK219, which had a minimal impact on intracellular calcium levels and did not significantly alter TRPV4 distribution at the plasma membrane, resulted in no significant change in cell volume. The effect of GSK219 on cell volume reduction in OC conditions was similar to that of PEG 300 (Fig. 5B). As expected, a significant linear correlation was observed between the plasma membrane association of TRPV4 and changes in cell volume (R² ∼ 0.59) (Fig. 5C). This finding confirms that the channel inhibition-driven relocation of TRPV4 to the plasma membrane in response to cell crowding, and the subsequent decrease in intracellular calcium, are associated with a reduction in cell volume in MCF10DCIS.comcells.

Cell crowding-induced plasma membrane TRPV4 association scales with cell volume reduction and increases in invasiveness and motility.

A-C. MCF10DCIS.comcell volume changes with TRPV4 inhibition and activation. (A) In ND MCF10DCIS.comcells, GSK101, which did not alter plasma membrane association of TRPV4, did not affect cell volume. Conversely, GSK219, which increased such association in a dose-dependent manner, reduced cell volume, with the effect of 1 nM GSK219 (219H) being similar to that of 74.4 mOsm/Kg PEG 300. (B) Under OC conditions, GSK101, which leads to significant Fluo-4 spikes, increased cell volume in a dose-dependent manner, while GSK219 and PEG only mildly reduced cell volume. (C) Cell volume changes in MCF10DCIS.comcells show an inverse relationship with plasma membrane association of TRPV4, reflecting the activation status of the channel (R² = 0.59). (D-F) Cell invasiveness increases with cell volume reduction and greater plasma membrane association of TRPV4. (D) Cell invasiveness significantly increases with GSK219 under ND conditions. (E) GSK101 under OC conditions shows a notable decrease in cell invasiveness in a dose-dependent manner. (F) Plasma membrane association of TRPV4 predictably reports cell invasiveness (R² ∼ 0.69), while cell invasiveness and cell volume are inversely related, reinforcing our observation that cell volume reduction promotes cell invasiveness. (H-M) To assess if cell motility also follows the trend of cell invasiveness, we performed a single-cell motility assay by tracking nuclear WGA in individual live cells every 5 s for 25 min. (H) Representative trajectories of individual cells are color-coded to reflect displacement at each time interval. Compared to untreated ND cells, 0.2 pM GSK101 treatment slowed overall cell diffusion, while 1 nM GSK219 and 74.4 mOsm/Kg PEG 300 treatments increased diffusion. Scale bar = 200 μm. Using single-cell analysis, we quantified cell diffusivity (D) and speed (v; movement directionality). (I) GSK101 treatment significantly reduced D. (J) GSK101 treatment increased v, while GSK219 treatment significantly increased D but did not affect directionality. (K) Like cell invasiveness, cell motility (D) positively scales with plasma membrane association of TRPV4 (R² ∼ 0.73). (L) Cell motility (D) inversely relates to cell volume (R² ∼ 0.89). (M) Cell motility (D) and cell invasiveness show a strong linear relationship (R² ∼ 0.85), enabling the use of cell motility measurements to assess overall cell invasiveness.

Next, we examined the overall connections between TRPV4 activity, its plasma membrane association, cell volume changes, and cell invasiveness. Thus, we evaluated changes in invasiveness following 2-day treatments with GSK101 or GSK219 using collagen-crosslinked polyacrylamide hydrogel matrix-based invasion assay used in Fig. 1. In ND MCF10DCIS.comcells, 0.2 pM GSK101 led to only a mild suppression of cell invasiveness, and 0.05 pM GSK101 did not alter cell invasiveness (Fig. 5D). In contrast, under ND conditions, GSK219 caused a dose-dependent increase in the invasive cell fraction, mirroring the patterns for Fluo-4 signal changes and plasma membrane-associated TRPV4 (Fig. 5D). Under OC conditions, GSK101 increased cell volume and suppressed cell invasiveness in a dose-dependent manner (Fig. 5E). Conversely, both GSK219 and PEG 300, which reduced cell volume, led to increased cell invasiveness (Fig. 5E). The treatment dependent invasion results are shown in Suppl. Fig. 12A. This solidified the relationship between TRPV4 activity, plasma membrane-associated TRPV4, cell volume, and cell invasiveness. There were significant linear correlations between plasma membrane-associated TRPV4 and cell invasiveness changes (Fig. 5F; R² ∼ 0.69), and between cell volume and cell invasiveness changes (Fig. 5G; R² ∼ 0.89).

Cell invasiveness refers to the ability of cells to penetrate and move within their surrounding tissues or environments. As expected from the increased cortical stiffness of OC MCF10DCIS.comcells (Fig. 2C), we observed that OC MCF10DCIS.comcells underwent cytoskeletal rearrangements, including the formation of stress fibers (Suppl. Fig. 12B), which likely enhanced cell motility75. To assess how TRPV4 inhibition and activation, which reduced and increased cell volume, respectively, affected the motility of ND MCF10DCIS.comcells and the role of these motility changes on cell invasiveness, we conducted single-cell tracking experiments similar to single molecule tracking methods60,76–80. We utilized wheat germ agglutinin (WGA) as a nuclear marker, which stained the cell nucleus within 1 hour of incubation with cells. We then analyzed the diffusivity (D) and directionality (v) of single cell movements. We tracked overall single-cell trajectories for 3 hr with 1 min intervals, marking cell displacement over each interval in different colors, ranging from navy for the slowest to red for the fastest (Fig. 5H). Consistent with the trend in cell invasiveness, TRPV4 activation with 1 pM GSK101 significantly reduced single-cell movements, while inhibition with 0.2 nM GSK219 and 74.4 mOsm/kg PEG 300 notably increased these movements. Interestingly, while GSK219 and hyperosmotic conditions promoted increased cell movement (D; Fig. 5I), they also reduced cell directionality (v; Fig. 5I). This reduction in directionality suggests that while the cells become more mobile, their movements are less targeted, potentially indicating a more random or exploratory behavior under mechanical stress. Remarkably, cell movement increased in proportion to the amount of TRPV4 at the plasma membrane (Fig. 5K; R² ∼ 0.73) and decreased with increased cell volume (Fig. 5L; R² ∼ 0.89). Cell movement and invasiveness showed a strong linear correlation (Fig. 5M; R² ∼ 0.85), suggesting that these behaviors are governed by the same underlying mechanisms.

Mechanosensitive TRPV4 relocation to plasma membrane indicates enhanced invasiveness

We investigated whether the mechanosensitive relocation of TRPV4 to the plasma membrane can reliably predict a cell’s ability to undergo pro-invasive mechanosensitive cell volume reduction. To achieve this, we used changes in cell motility as markers for overall shifts in cell invasiveness, given the strong correlation observed between cell motility and invasiveness (Fig. 5M). Our hypothesis is that cells with such mechanotransduction capability, like MCF10DCIS.comcells, should show increased plasma membrane association of TRPV4 when inhibited with GSK219 or under hyperosmotic conditions, evidenced by increased cell motility (D) while demonstrating the opposite effect with GSK101 treatment.

For comparison, we employed three uncharacterized cell types: MDA-MB-231, a triple-negative invasive breast cancer cell line, and two recently developed patient-derived DCIS cell lines, ETCC-06 and ETCC-1064,65, which have not yet been histologically classified by grade. These cells were compared with positive control groups, including MCF10DCIS.comand MCF10CA1a, which were observed to possess pro-invasive mechanotransduction capability (Fig. 5D, 6A-6C). MCF10AT1 served as a negative control, as it lacks this capability (Fig. 6D-6F).

Pro-invasive cell volume reduction mechanotransduction pathway is indicated by TRPV4 inhibition and hyperosmotic stress-driven increases in TRPV4 association with the plasma membrane and cell motility, as well as TRPV4 activation-induced decreases in cell motility.

(A) MCF10CA1a cells respond to 15 minutes of PEG 300 (74.4 mOsm/kg) and OC conditions by relocating TRPV4 to the plasma membrane, increasing the channel’s association with the membrane. IF images show a predominantly intracellular TRPV4 (red) distribution in the ND control MCF10CA1a cells, whereas plasma membrane association of TRPV4 occurs in response to PEG 300 and OC. DAPI (blue) staining indicates the nucleus. Scale bars throughout Fig. 6 = 10 μm. (B) Line analysis results of plasma membrane-associated TRPV4 (%) show significant increases in plasma membrane TRPV4 (PM TRPV4) with 1 hour of GSK219 (1 nM), 15 minutes of 74.4 mOsm/kg PEG 300, and OC conditions. (C) Cell movement diffusivity (D) increases with GSK219 and PEG treatments, while movement directionality (v) increases with GSK101 (0.2 pM) but shows no other effects. (D-O) MCF10AT1 (D, E), MDA-MB-231 (G, H), ETCC-06 (J, K), and ETCC-10 (M, N) do not relocate TRPV4 to the plasma membrane in response to TRPV4 inhibition by GSK219, hyperosmotic conditions by PEG-300, or OC conditions, as evidenced by IF images (TRPV4: red; DAPI: blue) and line analysis results for plasma membrane associations of TRPV4. (F) None of these cells’ motility responded to PEG 300. However, their responses to TRPV4 activation (GSK101) and inhibition (GSK219) varied, suggesting different roles of TRPV4 in their cancer biology. MCF10AT1 diffusivity (D) significantly reduced with GSK219, while the rest of the conditions did not affect D. Movement directionality (v) was not affected by any treatment conditions. (G) MDA-MB-231 cell D or v did not alter with any treatment conditions, suggesting an insignificant role of TRPV4 in their cell motility. (L, M) Both ETCC-06 and ETCC-10 D increased with GSK101, but GSK219 also increased ETCC-06 diffusivity, while not altering ETCC-10 diffusivity. ETCC-06 directionality (v) increased with GSK101, but ETCC-10 directionality remained unchanged. (P) Plasma membrane association of TRPV4 (% PM TRPV4) positively scalex with cell movement D over a larger range for MCF10DCIS.comcells than for MCF10CA1a, reflecting the high cell volume plasticity observed in MCF10DCIS.comcells. This result suggests that both cell types have a pro-invasive mechanotransduction pathway. (Q) Such scaling is absent in MCF10AT1, MDA-MB-231, ETCC-06, and ETCC-10 cells. (R) The presence of such a mechanotransduction pathway is observed for MCF10CA1a and MCF10DCIS.comcells from the plot of greater than 2-fold increase in plasma membrane association of TRPV4 (x-axis; PM TRPV4_peg/PM TRPV4_ctrl) and greater than 1-fold increase in diffusivity (y-axis; Dpeg/Dctrl) by PEG-300. (S) Cell volume reduction-mediated mechanotransduction pathway can be clearly observed from plotting plasma membrane association of TRPV4 (x-axis) and relative increase in diffusivity with GSK101 versus GSK219 (y-axis), which are significantly greater than 2 for MCF10CA1a and MCF10DCIS.comcells.

MCF10CA1a cells showed plasma membrane relocation of TRPV4 in response to TRPV4 inhibition (1 nM GSK219; ND 219H), PEG 300 (74.4 mOsm/kg; ND PEG300), and OC conditions. In contrast, treatment with 0.2 pM GSK101 did not affect TRPV4 plasma membrane association (Supplementary Fig. 13 and Fig. 6A, showing IF images for TRPV4 (red) and DAPI (cyan)). This response was similar to that observed in MCF10DCIS.comcells under OC conditions. The plasma membrane associations of TRPV4 were quantified by line analysis, showing an increase under mechanical stresses (PEG and OC) and TRPV4 inhibition with GSK219, as plotted in Fig. 6B. As expected, based on the differences in plasma membrane TRPV4 association, MCF10CA1a cells exhibited increased motility (D) in response to TRPV4 inhibition (GSK219) and hyperosmotic conditions (PEG). Additionally, their movement became more directional (v) with TRPV4 activation (GSK101), similar to the observations in MCF10DCIS.comcells (Fig. 5I, 5J).

In contrast, the negative control MCF10AT1 cells did not exhibit mechanical stress and TRPV4 activity-dependent plasma membrane relocation of TRPV4, as shown by IF images (Suppl. Fig. 13, Fig. 6D) and line analysis results (Fig. 6E). In these cells, TRPV4 inhibition (GSK219) decreased diffusivity (D), while PEG 300 had no effect. Moreover, movement directionality (v) remained unchanged under all treatment conditions (Fig. 6F).

The three test group cells, MDA-MB-231 (Fig. 6G, 6H), ETCC-06 (Fig. 6J, 6K), and ETCC-10 (Fig. 6M, 6N), failed to relocate TRPV4 to the plasma membrane in response to mechanical stresses (PEG and OC), as shown by IF images and quantification. This suggests that these cells lack the pro-invasive mechanosensitive cell volume reduction capability observed in MCF10DCIS.comcells. Consistently, none of these cells showed increased motility (D) under hyperosmotic conditions or TRPV4 inhibition (Fig. 6I, 6L, 6O).

We plotted these results to show cell diffusivity (D) scaling with varying plasma membrane associations of TRPV4, where MCF10DCIS.com(green hexagon) and MCF10CA1a (red hexagon) exhibited a positive scaling (Fig. 5P). However, the changes in plasma membrane TRPV4 occurred over a narrower range (31 – 79%) for MCF10CA1a compared to MCF10DCIS.comcells (16 - 75%), reflecting the larger mechanosensitive cell volume plasticity of MCF10DCIS.comcells. In contrast, MDA-MB-231, ETCC-6, and ETCC-10 cells did not show this scaling (Fig. 5Q). Like MCF10AT1 cells, plasma membrane TRPV4 associations in these cells remained at similar percentages (Fig. 5Q).

This dichotomy in the presence of a pro-invasive mechanotransduction program is evident in the plots of diffusivity (D) increase versus plasma membrane TRPV4 relocation (Fig. 5R), where only MCF10CA1a and MCF10DCIS.comcells show a significant increase in both parameters. Furthermore, the relationship between plasma membrane-associated TRPV4 and D increase with GSK219 versus GSK101 can be used to assess the presence of a pro-invasive, mechanosensitive cell volume reduction program. Most non-mechanotransducing cells remained close to the baseline value of 1, while mechanotransducing cells showed a significant increase in D with GSK219 treatment.

Discussion

Our study explored how cell crowding, a common condition in diseases and development, impacts cell invasiveness. We employed novel assays to quantify cell invasion and single-cell motility in cells with breast tissue pathologies, including DCIS and ADH, which experience cell crowding due to abnormal proliferation in confined ducts. Our results revealed a novel mechanotransduction pathway activated by crowding in high-grade DCIS cells, establishing a direct link between crowding and invasiveness. This pathway highlights the intricate interplay between mechanical stimuli, ion channel activation, trafficking changes, cell volume reduction, stiffening, and invasiveness (Fig. 7). It distinguishes high-grade DCIS cells from normal, hyperplastic, or lower-grade DCIS cells, supporting previous findings that minimal cell volume reduction enables efficient invasion38.

Cell crowding triggers the activation of a pro-invasive mechanotransduction pathway in high-grade DCIS cells but not in less aggressive or normal cells.

This pro-invasive mechanotransduction pathway involves cell volume reduction and cortical stiffening driven by ion channel inhibition. Ion channel inhibition leads to the relocation of ion channels to the plasma membrane (PM). Under ND conditions, ion channels are mainly cytoplasmic in both high-grade DCIS and less aggressive/normal cells. However, high-grade DCIS cells have a larger cell volume under ND conditions. Cell crowding induces ion channel inhibition, leading to decreased intracellular calcium and subsequent cell volume reduction, which correlates with increased invasiveness. In contrast, less aggressive and normal cells do not exhibit this response. The inset image shows that mechanosensitive plasma membrane relocation of ion channels (orange) scales with increased invasiveness due to the activation of the cell volume reduction mechanotransduction pathway, while the absence of relocation indicates no mechanosensitive invasive increase.

The correlation between cell volume reduction and increased invasiveness highlights the specialized mechanotransduction abilities of high-grade DCIS cells. Using mass spectrometry, we identified TRPV4 as a crucial component of this pathway, providing insights into how non-invasive breast cancer cells may evolve into invasive phenotypes. By employing Fluo-4 assays and TRPV4 activators and inhibitors, we found that cell crowding triggered TRPV4 inhibition, leading to reduced intracellular calcium and subsequent cell volume reduction. Notably, TRPV4 inhibition prompted its relocation to the plasma membrane, priming it for activation under suitable signaling conditions, thereby maintaining cellular homeostasis. In high-grade DCIS cells, TRPV4 inhibition by cell crowding increased invasiveness and motility, while TRPV4 activation had the opposite effect. This phenomenon differed from other invasive breast cancer cell lines, where TRPV4 inhibition typically suppresses invasiveness, and activation enhances it. However, the modest mechanotransduction capability observed in MCF10CA1a cells suggests heterogeneity in invasive cancer progression.81–83.

Our discovery of the selective mechanotransduction pathway in high-grade DCIS cells was further validated by analyzing patient-derived breast cancer tissues. We found that TRPV4 was predominantly associated with the plasma membrane in high-grade DCIS lesions, but not in lower-grade DCIS or less aggressive pathologies. This selective relocation reinforces TRPV4’s crucial role in the pro-invasive mechanotransduction pathway unique to high-grade DCIS cells. The association of TRPV4 with the plasma membrane may serve as a valuable biomarker for identifying high-grade DCIS lesions.

Our data indicate that ion channels beyond TRPV4 may contribute to pro-invasive mechanotransduction. Notably, SCN11A and KCNN4 showed increased plasma membrane association under cell crowding conditions, suggesting their roles in modulating ion flux and cell volume and thus contributing to the mechanotransduction response. Although Piezo1 is known to respond to mechanical stimuli62, its lesser plasma membrane relocation compared to TRPV4 under cell crowding and hyperosmotic conditions suggests a lesser role in promoting invasiveness in this case. Further comparisons are needed to determine the dominant ion channel driving mechanotransduction in high-grade DCIS cells. However, it is evident that cell crowding-induced pro-invasive cell volume reduction pathway involves the coordinated action of multiple ion channels27,84.

We made the intriguing discovery that hyperosmotic conditions, which reduced’ cell volume through osmotic water outflow, mimic the effects of cell crowding by inducing similar pro-invasive cell volume reduction. Treatment with hyperosmotic agents like PEG 300 led to TRPV4 relocation to the plasma membrane and increased cell invasiveness, highlighting TRPV4’s mechanosensitive role in response to external mechanical stimuli. This suggests that the convergence of mechanical stresses, including cell crowding and hyperosmotic stress, is a crucial trigger for pro-invasive cell volume reduction pathways, leading to a unified mechanotransduction signaling cascade.

The study’s insights have far-reaching implications, unlocking new research paths in cancer and biological sciences. By elucidating mechanosensitive responses to cell crowding, we can shed light on tissue development, wound healing, and physiological processes involving cell density changes. The parallels between cell crowding and hyperosmotic conditions in driving pro-invasive behaviors warrant further investigation into the intricate mechanisms of cytoskeleton reorganization, ion channel and transporter relocation, and enhanced cell invasiveness and motility. Moreover, our findings pave the way for developing diagnostic and prognostic strategies that leverage the selective intracellular localization patterns of TRPV4 and other mechanosensitive ion channels, potentially guiding clinical decisions for patients with high-grade DCIS and other cancers where cell crowding plays a crucial role.

Materials and methods

Cell culture and treatments

MCF10A cells were purchased from ATCC, and MCF10AT1, MCF10DCIS.co, and MCF10CA1a cells were procured from the Barbara Ann Karmanos Cancer Institute after establishing material transfer. These cells were authenticated prior to purchase, and cells with a low passage number (< 15) were used for experiments. We confirmed that the cells were mycoplasma-free using PCR analysis. MCF10A and MCF10AT1 cells were maintained in DMEM/F12 supplemented with 5% horse serum, 20 ng/mL EGF, 0.5 μg/mL hydrocortisone, 10 μg/mL insulin, and 1% penicillin and streptomycin. MCF10A cells were also supplemented with 100 ng/mL cholera toxin. MCF10DCIS.comand MCF10CA1a cells were cultured in DMEM/F12 supplemented with 5% horse serum and 1% penicillin and streptomycin. cells were cultured in 5% CO2 at 37 °C. For TRPV4 activator and inhibitor, we used GSK 1016790A (GSK101; Tocris 6433) and GSK 2193874 (GSK219; Tocris 5106).

For volume measurements, parental cells were mixed with those expressing RFP at a 9:1 ratio to distinguish individual cells. Cells were treated with 2% (74.4 mOsmol/kg) or 4% PEG-300 (148.8 mOsmol/kg; v/v; Millipore Sigma 8074845000) in culture medium for 15 min, 2% PEG-300 for 48 h, or GSK1016790A (Tocris; 0.05 and 0.2 pM) or GSK2193874 (Tocris; 0.2 and 1 nM) for 48 h.

TRPV4 knockdown by shRNA

Cells in growth medium without penicillin and streptomycin were transfected with the hTRPV4 shRNA plasmid (1 and 2 µg; Santa Cruz sc-61726-SH) and Lipofectamine3000 in opti-MEM. At 6 h after transfection, the medium was replaced, and cells were cultured for 30 h before processing for western blot and immunofluorescence analyses.

Immunofluorescence

The cells were fixed with 4% paraformaldehyde (Fisher Scientific) for 20 min, followed by permeabilization with 0.1% saponin (Fisher Scientific) for 5 min at room temperature (RT). After permeabilization, the cells were washed three times in PBS and blocked in 0.1% saponin + 10% BSA in PBS at RT for 1 h. Cells were incubated with primary antibodies overnight at 4 °C in 0.1% saponin in PBS. The primary antibodies used were TRPV4 (Abcam 39260; 1:500 dilution), TfR (ThermoFisher Scientific 13-6800; 1:500 dilution), Piezo1 (Alomone APC-087; 1:500), and KCNN4 (Alomone ALM-051; 1:500). Cells were washed three times in PBS, incubated with fluorescent-tagged anti-rabbit-Alexa 550 or anti-mouse-Alexa 488 secondary antibodies (Thermo Fisher Scientific) and DAPI for 1 h at RT, and imaged by confocal microscopy. For plasma membrane staining, DiIC18(3) (’1,1’-Dioctadecyl-3,3,3’,3’-Tetramethylindocarbocyanine Perchlorate; DiIC18(3); Thermo Fisher Scientific, Cat # D282) was diluted in warm PBS to 5 uM. Media was removed from the cells, then the 5 uM DiIC18(3) in PBS was added. The dish was transferred to the incubator (37 C, 5% CO2) for 10 minutes. Following the 10 min incubation, the DiIC18(3) in PBS was removed and the cells were washed 2x with complete cell growth media. The cells were then fixed with 2% PFA in PBS

Confocal imaging and image processing

2D images were acquired using a Yokogawa spinning-disk confocal microscope (Andor Technology) installed in a Nikon Eclipse TE2000 inverted microscope using a 60x/1.49NA Plan Apo objective (Nikon). The samples were illuminated using 430, 488, 561, and 647 nm solid-state lasers (Andor Technology). Images were acquired using an iXon back-illuminated EMCCD camera (Andor Technology). For volume measurements, 3D confocal images were acquired using z-step sizes calculated based on Nyquist conditions. Images were processed using the ImageJ (NIH) or Imaris (Bitplane) software. For live cell imaging, cells were maintained in 5% CO2 at 37 °C in a stage-top incubator (Oko-lab).

Western blotting

ND cells (40–60% cell density) were lysed in SDS sample buffer, and OC cells were lysed in cytoskeleton buffer (10 mM Tris pH 7.4,100 mM NaCl, 1 mM EDTA, 1 mM EGTA, 1% Triton X-100, 10% glycerol, 0.1% SDS, 0.5% deoxycholate). SDS sample buffer (Bio-Rad) supplemented with reducing reagent was added to lysates and boiled at 100°C for 10 min. The samples were separated on 4–15% SDS-PAGE gels and transferred to nitrocellulose membranes (Bio-Rad). Membranes were blocked in TBST (TBS + 0.1% Tween-20) containing 5% non-fat milk and then incubated with primary TRPV4 antibody (ThermoFisher Scientific PA541066; 1:1000) and GAPDH (Santa Cruz sc-32233; 1:2000) overnight at 4 °C. Membranes were washed three times in TBST (5 min each) and then incubated with horseradish peroxidase (HRP)-conjugated anti-rabbit (ThermoFisher Scientific A28177) or anti-mouse (ThermoFisher Scientific A28177) secondary antibodies for 30 min at RT. Blots were visualized using West Femto maximum sensitivity chemiluminescent substrate (Thermo Fisher Scientific).

Lentiviral infection and stable cell generation

To produce lentiviruses encoding RFP, HEK-293 cells were transfected with the lentiviral vector plasmid DNA using Lenti-X Packaging Single Shots (Takara). pCSII-EF-miRFP670v1-hGem(1/110) was a gift from Vladislav Verkhusha (Addgene plasmid # 80006; http://n2t.net/addgene:80006; RRID:Addgene_80006). Supernatants containing lentiviral pseudoparticles were harvested 24 and 48 h post-transfection. harvested lentiviral particles were immediately stored at −80 °C. To establish stable cell lines, 70% confluent cells at one day post-seeding were infected with lentivirus in the presence of 8 µg/mL polybrene (Thermo Fisher Scientific). Two days after transduction, 1 µg/mL puromycin (Thermo Fisher Scientific) was added to the medium to select for stably transduced cells. The samples were visualized daily to ensure that the untransduced cells in the wells were not viable. Once the polyclonal populations had sufficiently expanded, cell stocks were prepared and harvested for protein expression assays.

Cell invasion assay

For cell invasion assays, 8-well chamber slides (Nunc Lab-Tek) were coated with 50 µg/mL poly-L-lysine (Thermo Fisher Scientific) for 30 min and washed with PBS. The slides were fixed with 0.5% glutaraldehyde (Thermo Fisher Scientific) for 20 min and washed. Gelatin was conjugated with fluorescein by mixing 200 µL fluorescein-gelatin (1 mg/mL; ThemoFisher Scientific) with 800 μL unlabeled gelatin (4 mg/mL), followed by incubation for 5 min at 60 °C. The gelatine mix was allowed to cool at RT for 5 min and then applied, and the slides were incubated for 15 min. Slides were washed with PBS and disinfected with 70% ethanol for 30 min. Following three PBS washes, the residual reactive groups were quenched with growth medium by incubating at room temperature for 30 min in the dark. Cells were seeded in a fresh growth medium and incubated undisturbed on a horizontal surface at RT for 30 min to encourage cell distribution. Chamber slides were incubated with 5% CO2 at 37 °C. To detect cell invasion, cells were fixed with 4% formaldehyde and stained with DAPI (Thermo Fisher Scientific), and imaging was performed using a x4/0.2 NA Plan Apo objective (Nikon). Cell invasion was determined by quantifying the sites of the degraded matrix, which were visible as dark areas in the bright-green fluorescent gelatin matrix. The area of gelatin digestion and the number of cells were calculated using the ImageJ software.

Nanoindenter assay for measuring cortical stiffness

Stiffness measurements were performed on live cells in 5% CO2 at 37 °C, maintained in a stage-top chamber (Oko-lab) using a nanoindenter (Optics11 Chiaro) attached to the confocal microscope. Nanoindentation was performed at the cell surface of single cells with an indentation probe with a spring constant (0.24 N/m) and tip diameter (10 μm). The Hertzian contact model was used to fit the data to extract Young’s modulus in the elastic regime.

Surface biotinylation and pull-down assay

ND cells that were 40%–70% confluent and OC cells were detached and washed twice with PBS. The cells were resuspended at a concentration of 25 × 106 cells/mL in PBS containing 2 mM sulfo-NHS-biotin (Invitrogen). The reaction mixture was then incubated at RT for 60 min. Cells were washed three times with 1 M Tris pH 8.0 to quench and remove the excess biotin reagent. Cell pellets were lysed in 1.0 mL of cold lysis buffer (50 mM HEPES pH 7.2, 150 mM NaCl, 1.0% Triton X-100, 1.0% CHAPS, 100x protease and phosphatase inhibitors). Lysates were incubated on ice for 30 min, transferred to tubes, and centrifuged (10,000 × g for 5 min) to remove insoluble material. Supernatants were collected, and protein concentrations were measured using the bicinchoninic acid assay (Pierce). We washed 50 µL resin (ThermoFisher Scientific) with immobilized streptavidin twice with binding buffer (0.1 M phosphate, 0.15 M NaCl pH 7.2) and centrifuged at 5000 x g for 1 min. Biotin-labelled cell lysate (1 mg) was added to the resin and incubated with rotation for 1 h at RT. The resin was washed by resuspending in binding buffer, centrifuging to pellet the resin, and removing the supernatant by aspiration. The wash was repeated four times. Samples were boiled in SDS-PAGE sample buffer with DTT and separated by electrophoresis.

Mass spectrometry

Peptide mixtures from each sample were analyzed by LC-MS/MS using a nano-LC system (Easy nLC1000) connected to a Q Exactive HF mass spectrometer (Thermo Fisher Scientific). The platform was configured with a nano-electrospray ion source (Easy-Spray, ThermoFisher Scientific), an Acclaim PepMap 100 C18 nanoViper trap column (3 μm particle size, 75 μm ID × 20 mm length), and an EASY-Spray C18 analytical column (2 μm particle size, 75 μm ID × 500 mm length). Peptides were eluted at a flow rate of 300 nL/min using linear gradients of 5–25% acetonitrile (in aqueous phase and 0.1% formic acid) for 40 min, followed by 45% for 10 min, and static flow at 90% for 10 min. Mass spectrometry data were collected in a data-dependent manner, switching between one full-scan MS mode (m/z:380–1400; resolution: 60 K; AGC:3e6; maximum ion time:20 ms), and 10 MS/MS scans (resolution: 15 K; AGC:1e5; maximum ion time:120 ms; nCE:27) of the top 10 target ions. Ions were sequenced once and dynamically excluded from the list for 20 s. The datasets were searched using MaxQuant at default parameters against the UniProt Human Proteome database.

Line analysis

IF images of ion channels (TRPV4, PIEZO1, or KCNN4), DiIC18(3), and DAPI were opened in ImageJ. These images were concatenated so that a line crossing a cell in all three images could be used to measure approximate percentage protein localization in the plasma membrane, cytosol, or nucleus. Background was subtracted from all images. DiIC18(3) signal was used as a guide for plasma membrane location (between 50% of the peak intensity) and DAPI signal (between 50% of the peak intensity) for nuclear locations. The cytosol location was defined as the region between the 50% peak intensity points of the DiIC18(3) and DAPI signals. The ion channel signal in each window along the line was then averaged. The percentage of protein localization in the plasma membrane, cytosol, or nucleus for a cell was calculated by dividing the average intensity of each location by the sum of intensities in all three locations.

Pipeline for analyses of patient histologic samples

Specimen collection: A retrospective clinical study was designed to collect patient tissue blocks. A pathologist chose 39 patient breast formalin-fixed paraffin-embedded (FFPE) tissue specimens of a combination of the following pathologies (benign, ADH, IDC, low-, intermediate-, or high-grade DCIS, and IDC), along with normal tissue for each patient. Specimens associated with a past cancer diagnosis and drug treatment history were excluded.

Sample preparation for histopathologic evaluation: For each FFPE block, two serial dissections were carried out. Tissue sections on unlabeled slides were stained with hematoxylin and eosin (H&E) and subjected to immunohistochemical (IHC) staining using TRPV4 antibodies (Abcam 39260; 1:100 dilution). Whole-slide images were captured using the Olympus VS200 whole-slide imaging system. To validate the binding specificity of the antibody, both negative and positive control samples were prepared.

Two sequential annotations: A breast surgical pathologist annotated the pathology stages within specified regions of interest (ROIs) on wholeslide H&E images, based on cell morphological features. Two pathologists independently evaluated the same ROIs in the corresponding IHC images. They annotated the protein distribution using a three-tier classification:

Case 1: Absence of TRPV4.

Case 2: Intracellular TRPV4 localization.

Case 3: Presence of TRPV4 in the plasma membrane, with or without intracellular TRPV4.

The independent annotations concerning protein localization by the two pathologists were then subjected to statistical tests for selectivity and specificity. Any IHC ROIs with classification disagreements between the pathologists were designated as equivocal cases.

Calcium reporter assay

This assay was completed using the Fluo-4 Direct™ Calcium Assay Kit (Thermo Fisher F10471), including the assay buffer (Thermo Fisher), 2X Fluo-4 calcium assay reagent (Thermo Fisher), and preweighed, water-soluble probenecid (Thermo Fisher P36400; 2.5 mM). The 2X Fluo-4 calcium assay reagent was first prepared by adding 10 mL of the Fluo-4 calcium assay buffer and 200 μL of 250 mM probenecid to the desiccated calcium assay reagent at room temperature (21 °C). This mixture was then vortexed and allowed to sit for 15 minutes. While waiting for the 2X Fluo-4 calcium assay reagent to finish being prepared, the media from the well of a LabTek II Chambered Coverglass with Cover #1.5 Borosilicate Sterile 8-well plate (Thermo Fisher 155409) was removed and replaced with 200 μL of complete media. Once the 2X Fluo-4 calcium assay reagent was fully prepared, 200 μL of the reagent was added to the well with 200 μL of complete media. The 8-well plate was then placed at 37 °C and 5% CO₂ for 30 minutes. Following 30 minutes at 37 °C and 5% CO₂, the 8-well plate was removed from the incubator and placed at room temperature for 30 minutes. After incubation in the reagent for 30 minutes at room temperature, the 8-well plate was imaged using the confocal microscope with a 488 nm laser for 30 minutes with 30-second intervals between images and an exposure time of 200 ms. If the sample was treated, the sample was imaged for 35 minutes with 30-second intervals, with the treatment being added after 5 minutes had elapsed. Added treatments all had a final volume of 200 μL and were composed of: 1 nM GSK 2 in 1:1 media:2X Fluo-4 calcium assay reagent, 0.2 nM GSK 2 in 1:1 media:2X Fluo-4 calcium assay reagent, 0.2 pM GSK 1 in 1:1 media:2X Fluo-4 calcium assay reagent, 1:1:1 water:media:2X Fluo-4 calcium assay reagent, and 0.3 Osm/L PEG 300 in 1:1 media:2X Fluo-4 calcium assay reagent.

Cell motility assay

Cells were first seeded in MatTek 35 mm dishes with No. 1.5 coverslip, 14 mm glass diameter, and Poly-D-Lysine coated glass bottoms (MatTek P35GC-1.5-14-C). Once the cells adhered, they were stained with 10 μg/mL WGA-488 (Thermo Fisher W11261) for 10 minutes at room temperature. Following the staining, the cells were washed with complete media twice and then incubated at 37 °C and 5% CO₂ for 1 hour. During this one-hour incubation, the cells received the following treatments: 0.2 pM GSK 101 in complete media for 1 hour, 0.05 pM GSK 101 in complete media for 1 hour, 1 nM GSK 219 in complete media for 1 hour, 0.2 nM GSK 219 in complete media for 1 hour, 74.4 mOsm/L PEG 300 in complete media for 15 minutes, and 1:1 hypoosmotic solution in complete media for 15 minutes. After the incubation, the sample was placed in an OkoLab chamber at 37 °C and 5% CO₂ for imaging. Imaging was completed using a confocal microscope with a 488 nm laser for 3 hours with 1-minute intervals.

Data availability

Data available on reasonable request from the corresponding author (I.C.).

eLife Assessment

In this important study, Bu et al investigate how cell overcrowding triggers a mechano-transduction pathway involving TRPV4 channels, focusing on high-grade ductal carcinoma in situ (DCIS) cells. The authors show that cell crowding in these malignant cells leads to a reduction in cell volume and promotes a pro-invasive phenotype through calcium homeostasis and TRPV4 channel trafficking to the plasma membrane; this phenomenon is specific to invasive cell lines like MCF10CA and DCIS and is corroborated by patient tissue samples. The work suggests the role of TRPV4 in cell motility and mechanical sensing, offering potential therapeutic insights for targeting cancer metastasis. While the study presents robust and convincing data, the absence of TRPV4 genetic ablation is a critical limitation, which would further confirm its role in these processes.

Reviewer #1 (Public review):

Summary:

In this study, Bu et al examined the dynamics of TRPV4 channel in cell overcrowding in carcinoma conditions. They investigated how cell crowding (or high cell confluence) triggers a mechano-transduction pathway involving TRPV4 channels in high-grade ductal carcinoma in situ (DCIS) cells that leads to large cell volume reduction (or cell volume plasticity) and pro-invasive phenotype.

In vitro, this pathway is highly selective for highly malignant invasive cell lines derived from a normal breast epithelial cell line (MCF10A) compared to the parent cell line, but not present in another triple-negative invasive breast epithelial cell line (MDA-MB-231). The authors convincingly showed that enhanced TRPV4 plasma membrane localization correlates with high-grade DCIS cells in patient tissue samples.

Specifically in non-invasive MCF10DCIS.com cells, they showed that overcrowding or over-confluence leads to a decrease in cell volume and intracellular calcium levels. This condition also triggers the trafficking of TRPV4 channels from intracellular stores (nucleus and potentially endosomes), to the plasma membrane (PM). When these over-confluent cells are incubated with a TRPV4 activator, there is an acute and substantial influx of calcium, attesting to the fact that there are a high number of TRPV4 channels present on the PM. Long-term incubation of these over-confluent cells with the TRPV4 activator results in the internalization of the PM-localized TRPV4 channels.

In contrast, cells plated at lower confluence primarily have TRPV4 channels localized in the nucleus and cytosol. Long-term incubation of these cells at lower confluence with a TRPV4 inhibitor leads to the relocation of TRPV4 channels to the plasma membrane from intracellular stores and a subsequent reduction in cell volume. Similarly, incubation of these cells at low confluence with PEG 3000 (a hyperosmotic agent) promotes the trafficking of TRPV4 channels from intracellular stores to the plasma membrane.

Strengths:

The study is elegantly designed and the findings are novel. Their findings on this mechano-transduction pathway involving TRPV4 channels, calcium homeostasis, cell volume plasticity, motility, and invasiveness will have a great impact in the cancer field and are potentially applicable to other fields as well. Experiments are well-planned and executed, and the data is convincing. The authors investigated TRVP4 dynamics using multiple different strategies- overcrowding, hyperosmotic stress, and pharmacological means, and showed a good correlation between different phenomena.

Weaknesses:

A major emphasis in the study is on pharmacological means to relate TRPV4 channel function to the phenotype. I believe the use of genetic means would greatly enhance the impact and provide compelling proof for the involvement of TRPV4 channels in the associated phenotype. In this regard, I wonder if siRNA-mediated knockdown of TRPV4 in over-confluent cells (or knockout) would lead to an increase in cell volume and normalize the intracellular calcium levels back to normal, thus ultimately leading to a decrease in cell invasiveness.

Reviewer #2 (Public review):

Summary:

The metastasis poses a significant challenge in cancer treatment. During the transition from non-invasive cells to invasive metastasis cells, cancer cells usually experience mechanical stress due to a crowded cellular environment. The molecular mechanisms underlying mechanical signaling during this transition remain largely elusive. In this work, the authors utilize an in vitro cell culture system and advanced imaging techniques to investigate how non-invasive and invasive cells respond to cell crowding, respectively.

Strengths:

The results clearly show that pre-malignant cells exhibit a more pronounced reduction in cell volume and are more prone to spreading compared to non-invasive cells. Furthermore, the study identifies that TRPV4, a calcium channel, relocates to the plasma membrane both in vitro and in vivo (patient samples). Activation and inhibition of the TRPV4 channel can modulate the cell volume and cell mobility. These results unveil a novel mechanism of mechanical sensing in cancer cells, potentially offering new avenues for therapeutic intervention targeting cancer metastasis by modulating TRPV4 activity. This is a very comprehensive study, and the data presented in the paper are clear and convincing. The study represents a very important advance in our understanding of the mechanical biology of cancer.

Weaknesses:

However, I do think that there are several additional experiments that could strengthen the conclusions of this work. A critical limitation is the absence of genetic ablation of the TRPV4 gene to confirm its essential role in the response to cell crowding.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

Summary:

In this study, Bu et al examined the dynamics of TRPV4 channel in cell overcrowding in carcinoma conditions. They investigated how cell crowding (or high cell confluence) triggers a mechano-transduction pathway involving TRPV4 channels in high-grade ductal carcinoma in situ (DCIS) cells that leads to large cell volume reduction (or cell volume plasticity) and proinvasive phenotype.

In vitro, this pathway is highly selective for highly malignant invasive cell lines derived from a normal breast epithelial cell line (MCF10CA) compared to the parent cell line, but not present in another triple-negative invasive breast epithelial cell line (MDA-MB-231). The authors convincingly showed that enhanced TRPV4 plasma membrane localization correlates with highgrade DCIS cells in patient tissue samples.

Specifically in invasive MCF10DCIS.com cells, they showed that overcrowding or overconfluence leads to a decrease in cell volume and intracellular calcium levels. This condition also triggers the trafficking of TRPV4 channels from intracellular stores (nucleus and potentially endosomes), to the plasma membrane (PM). When these over-confluent cells are incubated with a TRPV4 activator, there is an acute and substantial influx of calcium, attesting to the fact that there are a high number of TRPV4 channels present on the PM. Long-term incubation of these over-confluent cells with the TRPV4 activator results in the internalization of the PMlocalized TRPV4 channels.

In contrast, cells plated at lower confluence primarily have TRPV4 channels localized in the nucleus and cytosol. Long-term incubation of these cells at lower confluence with a TRPV4 inhibitor leads to the relocation of TRPV4 channels to the plasma membrane from intracellular stores and a subsequent reduction in cell volume. Similarly, incubation of these cells at low confluence with PEG 3000 (a hyperosmotic agent) promotes the trafficking of TRPV4 channels from intracellular stores to the plasma membrane.

Strengths:

The study is elegantly designed and the findings are novel. Their findings on this mechanotransduction pathway involving TRPV4 channels, calcium homeostasis, cell volume plasticity, motility, and invasiveness will have a great impact in the cancer field and are potentially applicable to other fields as well. Experiments are well-planned and executed, and the data is convincing. The authors investigated TRVP4 dynamics using multiple different strategies- overcrowding, hyperosmotic stress, and pharmacological means, and showed a good correlation between different phenomena.

Weaknesses:

A major emphasis in the study is on pharmacological means to relate TRPV4 channel function to the phenotype. I believe the use of genetic means would greatly enhance the impact and provide compelling proof for the involvement of TRPV4 channels in the associated phenotype.

In this regard, I wonder if siRNA-mediated knockdown of TRPV4 in over-confluent cells (or knockout) would lead to an increase in cell volume and normalize the intracellular calcium levels back to normal, thus ultimately leading to a decrease in cell invasiveness.

We greatly appreciate the positive feedback regarding the design of our study and the novelty of our findings. We also acknowledge the valuable suggestion to complement our pharmacological approaches with genetic manipulation of TRPV4.

In response to the comment regarding siRNA-mediated knockdown or knockout of TRPV4, we fully agree that this would further substantiate our findings. In the revised manuscript, we implemented shRNA targeting TRPV4 to investigate its functional effects on intracellular calcium level changes, cell volume plasticity, and invasiveness phenotypes, assessed through singlecell motility assays under cell crowding or hyperosmotic stress. These results have been incorporated into the revised manuscript, and detailed descriptions of these findings are included below.

Using the shRNA approach that resulted in ~50% reduction of TRPV4 expression

(Supplementary Figure 6A and 6B show TRPV4 expression levels via IF and immunoblots, respectively), we examined the effect of reduced TRPV4 on intracellular calcium levels in MCF10DCIS.com cells under normal density (ND) and stress conditions (confluent; Con and hyperosmotic; PEG) using Fluo-4 AM imaging (Fig. 4S-X). We found that shRNA TRPV4 slightly decreased calcium levels in ND cells, likely due to fewer active calcium channels at the plasma membrane resulting from lower TRPV4 expression (as shown in the summary plot in Fig. 4W). With fewer active calcium channels, cells treated with shRNA TRPV4 exhibited less reduction in intracellular calcium levels under cell crowding conditions compared to control cells. Additionally, hyperosmotic stress using PEG 300 induced smaller calcium spikes in shRNA cells compared to the significant spike observed in control cells. This reduced calcium response to Con and hyperosmotic stress in shRNA cells was reflected in the decreased cell volume reduction by PEG 300 shown in Fig. 4Y. Consequently, shRNA-mediated TRPV4 reduction impaired cell volume plasticity in MCF10DCIS.com cells and abolished the pro-invasive mechanotransduction capability involving cell volume reduction, as evidenced by no increase in cell motility (both cell diffusivity and directionality) under hyperosmotic conditions (Fig. 5H-J). These findings demonstrate the critical role of TRPV4 in conferring pro-invasive

mechanotransduction capability to MCF10DCIS.com cells through cell volume reduction.

Reviewer #2 (Public review):

Summary:

The metastasis poses a significant challenge in cancer treatment. During the transition from non-invasive cells to invasive metastasis cells, cancer cells usually experience mechanical stress due to a crowded cellular environment. The molecular mechanisms underlying mechanical signaling during this transition remain largely elusive. In this work, the authors utilize an in vitro cell culture system and advanced imaging techniques to investigate how non-invasive and invasive cells respond to cell crowding, respectively.

Strengths:

The results clearly show that pre-malignant cells exhibit a more pronounced reduction in cell volume and are more prone to spreading compared to non-invasive cells. Furthermore, the study identifies that TRPV4, a calcium channel, relocates to the plasma membrane both in vitro and in vivo (patient samples). Activation and inhibition of the TRPV4 channel can modulate the cell volume and cell mobility. These results unveil a novel mechanism of mechanical sensing in cancer cells, potentially offering new avenues for therapeutic intervention targeting cancer metastasis by modulating TRPV4 activity. This is a very comprehensive study, and the data presented in the paper are clear and convincing. The study represents a very important advance in our understanding of the mechanical biology of cancer.

Weaknesses:

However, I do think that there are several additional experiments that could strengthen the conclusions of this work. A critical limitation is the absence of genetic ablation of the TRPV4 gene to confirm its essential role in the response to cell crowding.

We are deeply grateful for the positive assessment of our study and its contribution to advancing our understanding of mechanical signaling in cancer progression. We also greatly appreciate the suggestion to incorporate genetic ablation experiments to further validate the role of TRPV4 in cell crowding responses.

As noted in our response to Reviewer #1, we employed an shRNA approach to investigate the functional effects of TRPV4 knockdown on intracellular calcium level changes, cell volume plasticity, and invasiveness phenotypes. We assessed these effects using Fluo-4 AM calcium assay, single-cell volume measurements, and single-cell motility assays under cell crowding or hyperosmotic stress. These results have been incorporated into the revised manuscript and are described in detail in our response to Reviewer #1's "weaknesses" comment.

Reducing TRPV4 expression levels by shRNA diminished mechanosensing intracellular calcium changes under cell crowding and hyperosmotic conditions using PEG 300 treatment. Furthermore, a significantly reduced cell volume plasticity was observed under hyperosmotic conditions in shRNA treated cells compared to control cells (Fig. 4S-X). This diminished mechanosensing capability abolished the pro-invasive mechanotransduction effect, as assessed by single cell motility under hyperosmotic conditions (Fig. 5H-J). These findings demonstrate the critical role of TRPV4 in conferring pro-invasive mechanotransduction capability to MCF10DCIS.com cells through cell volume reduction.

Reviewer #1 (Recommendations for the authors):

The way the results or discussion section is written. It was a little confusing for me to relate to some phenomena. For example, it is not clear how TRPV4 inhibition (due to overcrowding) leads to a decrease in intercellular calcium levels, especially when TRPV4 channels were intercellular (not on the PM) to begin with (in normal density (ND) conditions). Along the same lines, how GSK219 causes a dip in calcium levels in ND cells when TRPV4 channels are primarily intercellular (Figure 4E). If most of the TRPV4 channels that are translocated to the PM in response to cell crowding are in an inactive state, how do they confer enhanced cell volume plasticity relative to non-invasive cell lines?

Thank you very much for raising this important point. We fully agree with your concern and have significantly revised the manuscript to clarify this aspect. Specifically, we have emphasized that a modest level of TRPV4 channels are constitutively active at the plasma membrane in normal density (ND) cells. This is now discussed in detail in the context of Fig. 4:

Page 14: “Considering these factors, we hypothesized that cell crowding might inhibit calcium-permeant ion channels that are constitutively active at the plasma membrane, including TRPV4, which would then lower intracellular calcium levels and subsequently reduce cell volume via osmotic water movement.”

Page 16-17: “… However, the temporal profile of Fluo-4 intensity in Fig. 4E, which corresponds to the time points marked in Fig. 4D (t1: baseline and t2: dip), clearly shows the dip at t2, indicated by ΔCa (the vertical dashed line between the dip and baseline). This modest Fluo-4 dip at t2 represents the inhibition of activity by GSK219 on a small population of constitutively active TRPV4 channels at the plasma membrane under ND conditions.

In Con cells, 1 nM GSK219 caused a smaller dip in Fluo-4 intensity compared to the one observed in ND cells, with no subsequent changes. This is likely due to fewer constitutively active TRPV4 at the plasma membrane in Con cells than in ND cells. …These findings suggest that a substantial portion of TRPV4 channels relocated to the plasma membrane under cell crowding was inactive, and some constitutively active TRPV4 channels already present in the membrane became inactive as a result of cell crowding.”

'Internalization' might be a better word than 'uptake' in the following line in the results section

"...activating TRPV4 under cell crowding conditions triggered channel uptake, indicating that TRPV4 trafficking depended on the channel's activation status."

Thank you very much for this suggestion. As recommended, we replaced ‘uptake’ with internalization’ on page 18:

“However, in Con cells, where a large number of inactive TRPV4 channels are likely located at the plasma membrane, GSK101 treatment notably reduced plasma membrane-associated TRPV4 in a dose-dependent manner through internalization (Fig. 4O, 4Q), consistent with previous findings65. These data suggest that plasma membrane TRPV4 levels were largely

regulated by the channel activity status. Specifically, channel activation led to the internalization of TRPV4, while channel inhibition promoted the relocation of TRPV4 to the plasma membrane.”

1. Out of curiosity:

2. Is there any information on what the intercellular TRPV4 channels are doing in the cytosol and in the nucleus? Is there any role of intercellular calcium stores in the proposed pathway?

We greatly appreciate this insightful question. Although we were unable to find studies specifically exploring the roles of cytosolic TRPV4, a recent study (Reference 74) identified a role for nuclear TRPV4 in regulating calcium within the nucleus. We speculate that when TRPV4 activity is severely impaired, such as with additional TRPV4 inhibition under cell crowding conditions, some TRPV4 channels may be redirected to the nucleus. This redistribution could help maintain nuclear calcium homeostasis.

This discussion is included on page 18 of the manuscript:

“These findings suggest that further TRPV4 inhibition under crowding conditions triggers a distinct trafficking alteration. Recent studies have implicated nuclear TRPV4 in regulating nuclear Ca2+ homeostasis and Ca2+-regulated transcription74. In light of this study and our findings, TRPV4 may relocate to the nucleus as a compensatory mechanism to maintain nuclear calcium regulation. This relocation could reflect an adaptive response to preserve calcium-dependent transcriptional programs or other nuclear processes essential for cell survival under mechanical stress.”

One recommendation is to add some explanation or some minor details for the convenience of the reader. For example:

At normal or lower confluence, cells show an acute large dip in intercellular calcium when an inhibitor is applied implying that there are a few TRPV4 channels on the PM and they are constitutively active.

Thank you very much for highlighting this important point and for the helpful suggestion to improve clarity. We have significantly revised the text associated with Fig. 4 to ensure this point is clear. Specifically, we have added the following explanation on page 16:

"This modest Fluo-4 dip at t2 represents the inhibition of activity by GSK219 on a small population of constitutively active TRPV4 channels at the plasma membrane under ND conditions."

Reviewer #2 (Recommendations for the authors):

(1) Figure 1. The authors frequently change the medium to prevent acidification in overconfluent cultures. A cell viability assay should be performed to ensure that the over-confluent cells remain healthy and viable during the experiments. There are commercial kits that can be easily used to quantify the number of viable cells and the extent of cell toxicity. The number of viable cells would provide a more reliable basis for comparison between normal density and overconfluent conditions.

Thank you very much for raising this important point. We have consistently observed that cell crowding does not induce significant cell death in MCF10DCIS.com cells. To address your recommendation, we performed a viability assay using propidium iodide (PI) to selectively stain dead cells and WGA-488 to stain all live cells. Cell death was quantified under normal density (ND) conditions and at 1, 3, 5, 7, and 10 days post-confluence.

Our results indicate that cells remain similarly viable post-confluence, with minimal cell death

(~1.5%) compared to ND cells (~0.75%). These findings are summarized in Supplementary Figure 2, demonstrating that over-confluent cultures remain healthy and viable during the experiments.

(2) Figure 2. To determine whether the reduction in cell volume is reversible, over-confluent cells can be further diluted back to normal density. Additionally, the reversibility of TRPV4 channel trafficking to the plasma membrane should be assessed under these conditions in IF experiments and cell surface biotinylation.

Thank you for this suggestion. We reseeded the previously overcrowded (OC) cells at normal density and observed that their TRPV4 distribution predominantly returned to being intracellular, with only modest plasma membrane localization, as shown by line analysis (Supplementary Figure 10A-C, page 13). Furthermore, their invasiveness decreased to levels comparable to the original normal density (ND) cells (Supplementary Figure 3C and 3E, page 6). These results demonstrate the reversibility of TRPV4 trafficking changes and the increase in invasiveness under mechanical stress.

Page 6. "The enhanced invasiveness of MCF10DCIS.com cells under cell crowding was largely reversible. When OC cells were reseeded at normal density for invasion assays, their invasive cell fraction decreased to approximately 15%, slightly lower (p = 0.012) than the initial value of around 24% (Suppl. Fig. 3C, 3E)."

Page 13. “We investigated whether TRPV4 relocation to the plasma membrane induced by cell crowding is reversible, as suggested by its impact on invasiveness (Suppl. Fig. 3E). To test this, previously OC MCF10DCIS.com cells were reseeded under ND conditions. We then assessed TRPV4 localization via immunofluorescence (IF) imaging to determine if most channels returned to the cytoplasm and could be relocated to the plasma membrane under mechanical stress, such as hyperosmotic conditions. Consistent with their initial ND state, reseeded ND MCF10DCIS.com cells displayed intracellular TRPV4 distribution (Suppl. Fig. 10A). Upon exposure to hyperosmotic stress (74.4 mOsm/Kg PEG300), TRPV4 was again relocated to the plasma membrane (Suppl. Fig. 10B). These findings, quantified through line analysis (Suppl. Fig. 10C), demonstrate that the mechanosensing response of MCF10DCIS.com cells is reversible.”

(3) Figure 3B. A control using intracellular proteins such as GAPDH or Tubulin is missing. Including this control would help exclude the possibility of cell rupture or compromised cell membranes in crowded environments, which is very common in a cell crowding environment.

Thank you very much for pointing this out. The control lanes (GAPDH) were already included in the full gel results shown in Supplementary Figure 5. For the immunoprecipitation and immunoblotting of surface-biotinylated cell lysates, we did not expect to detect GAPDH; however, some GAPDH signals were still observed. As shown for MCF10DCIS.com cells, less GAPDH was detected under OC conditions, but the immunoprecipitated samples displayed significantly higher levels of TRPV4 on the cell surface compared to ND cells (Supplementary Figure 5A). For the whole cell lysates, TRPV4 protein levels were comparable across different cell lines based on the immunoblot results, with consistent GAPDH signals serving as a loading control (Supplementary Figure 5B).

(4) Figure 4. To convincingly demonstrate TRPV4 relocation to the plasma membrane, IF should be performed under non-permeable conditions (i.e., without detergents like saponin). This approach ensures that only plasma membrane proteins are accessible to antibodies, reducing intracellular background. The same approach should be applied to Piezo1 and TfR.

Thank you for this suggestion. We observed that under non-permeable conditions, primary antibodies could still access intracellular proteins. To address this issue, we employed extracellular-binding TRPV4 antibodies to selectively detect TRPV4 relocation to the plasma membrane under hyperosmotic conditions (74.4 mOsm/kg PEG 300) in live MCF10DCIS.com cells, as shown in Supplementary Figure 9. These results clearly demonstrate the plasma membrane relocation of TRPV4 under hyperosmotic conditions, distinguishing it from control conditions. Unfortunately, we were unable to identify high-affinity extracellular-binding antibodies for Piezo1 and TfR. Nevertheless, our findings strongly support the mechanosensing plasma membrane relocation of TRPV4.

Essential Weakness:

Throughout the study, only TRPV4 inhibitors and activators were used to show that TRPV4 relocation is associated with intracellular calcium concentration and cell size changes. It is crucial to use TRPV4 KO or KD cells to confirm that the observed effects are specific to TRPV4 and not due to off-target effects on other proteins. Additionally, fusing a plasma membrane targeting sequence to TRPV4 to make a constitutive plasma membrane-localized construct could demonstrate the opposite effect.

Thank you very much for this important comment. As noted in our response to Reviewer #1, we employed an shRNA approach to investigate the functional effects of TRPV4 knockdown on intracellular calcium level changes, cell volume plasticity, and invasiveness phenotypes. We assessed these effects using Fluo-4 AM calcium assay, single-cell volume measurements, and single-cell motility assays under cell crowding or hyperosmotic stress. These results have been incorporated into the revised manuscript and are described in detail in our response to Reviewer #1's "weaknesses" comment.

Reducing TRPV4 expression levels by shRNA diminished mechanosensing intracellular calcium changes under cell crowding and hyperosmotic conditions using PEG 300 treatment. Furthermore, a significantly reduced cell volume plasticity was observed under hyperosmotic conditions in shRNA treated cells compared to control cells (Fig. 4S-X). This diminished mechanosensing capability abolished the pro-invasive mechanotransduction effect, as assessed by single cell motility under hyperosmotic conditions (Fig. 5H-J). These findings demonstrate the critical role of TRPV4 in conferring pro-invasive mechanotransduction capability to MCF10DCIS.com cells through cell volume reduction.

Minor Points:

The introduction section is poorly written; many results currently included in the introduction would be more appropriately placed in the discussion section. The long redundant introduction makes the article hard to read through.

Thank you very much for pointing this out. In the revised introduction, we have significantly reduced references to the results, streamlining the section to make it more concise and focused. This adjustment ensures the introduction is clearer and avoids redundancy, improving the readability of the manuscript.

Version 2

Abstract

Abstract

Cell crowding is a common microenvironmental factor that affects various disease processes, but its impact on cell invasiveness into surrounding tissues is not well understood. This study investigates the biomechanical changes induced by cell crowding, focusing on pro-invasive cell volume reduction in ductal carcinoma in situ (DCIS) cells. DCIS is a non-invasive form of breast cancer characterized by abnormal cell growth confined within the breast duct. While DCIS can undergo invasive transition, it is unclear which DCIS cells are predisposed to this transition. We discovered that cell crowding enhanced the invasiveness of high-grade DCIS cells, which experienced significant cell volume reduction compared to hyperplasia-mimicking or normal cells. Mass spectrometry analyses revealed that cell crowding relocated ion channels, including TRPV4, a calcium-permeant ion channel, to the plasma membrane selectively in high-grade DCIS cells but not in less aggressive or normal cells. Cell crowding inhibited TRPV4 activity in high-grade DCIS cells, decreasing intracellular calcium levels and reducing cell volume. This inhibition also triggered the relocation of TRPV4 to the plasma membrane, effectively priming the inactive channel for activation and mitigating the calcium loss caused by crowding-induced inhibition. Analyses of patient-derived breast cancer tissues validated that TRPV4 is selectively associated with the plasma membrane in high-grade DCIS but not in lower-grade DCIS or less aggressive pathologies. The extent of plasma membrane TRPV4 association scaled with cell volume reduction and increased cell invasiveness and motility, suggesting its utility as an active pro-invasive mechanotransduction pathway indicator. Additionally, hyperosmotic conditions and pharmacologic TRPV4 inhibition mimicked the pro-invasive volume reduction observed under cell crowding, while TRPV4 activation reversed this effect by inducing cell volume increase. Silencing the TRPV4 gene via shRNA diminished the mechanotransduction capability of high-grade DCIS cells, as demonstrated by reduced intracellular calcium depletion, attenuated cell volume reduction, and decreased motility. In summary, this study uncovers a previously unrecognized pro-invasive mechanotransduction pathway initiated by cell crowding, which is specific to high-grade DCIS cells, revealing a potential biomarker for identifying DCIS patients at high risk of invasive transition.

Introduction

The complex interplay between cellular mechanics and invasive behaviors is crucial to understanding various physiological and pathological processes, such as wound healing1 and disease progression2,3. Numerous studies show that these processes are mediated by mechanotransduction, whereby cells translate mechanical stimuli into cellular activity2,4–6. While mechanotransduction is traditionally studied concerning fluid stress7,8, matrix stiffness4,9, and other biomechanical changes such as osmotic stress10,11, the role of cell crowding, characterized by increased cell density and spatial constraints, is relatively less explored. A recent report shows that cell crowding plays a role in facilitating wound closure and repair by enhancing cell proliferation12. As tissues develop, repair, or undergo pathological transformation, cell crowding becomes common13,14. This challenges individual cells, forcing them to perceive and respond to the mechanical constraints of a crowded environment. Our study used human breast cell line model systems to describe a novel mechanotransduction pathway triggered by cell crowding that induces invasiveness into surrounding tissues. Interestingly, this pathway exhibited unique selectivity, as it specifically associated with a type of non-invasive cancer pathology and was not present in lower-grade or less aggressive pathologies. This suggests that not all cells possess the ability to translate mechanical stimuli, such as cell crowding, into cell invasiveness.

To assess the role of cell crowding in cell invasiveness, we chose in vitro cell lines linked to different pathological states that reflect cell crowding conditions in vivo, including atypical ductal hyperplasia (ADH)15 and ductal carcinoma in situ (DCIS)16–18. ADH is an intraductal clonal epithelial cell proliferative lesion18 and represents an intermediate step between normal breast tissue and in situ carcinomas19–21. ADH is associated with high risk as it is reported that an ADH diagnosis is associated with a five-fold increased risk of breast cancer15. DCIS is non-invasive form of cancer characterized by proliferating malignant epithelial cells16–18. Unlike ADH, DCIS is considered a precursor to invasive breast cancer22. However, the mechanism of how DCIS transitions to invasive cancer is not well understood and therefore, there is currently no reliable and robust method to differentiate which DCIS cells are at high risk of becoming invasive23,24.

Both ADH and DCIS conditions can potentially expose cells to crowding in vivo. However, how ADH and DCIS cells respond to such changes in cell density remains unknown. Our study revealed that cell crowding selectively triggered a pro-invasive mechanotransduction program in a specific type of DCIS cell line associated with high-grade comedo-type pathology25,26. The mechanotransduction program induced by cell crowding involved the inhibition of ion channels, such as transient receptor potential vanilloid 4 (TRPV4)27, a calcium-permeable ion channel, as identified from our mass spectrometry assay. This inhibition prompted the relocation of TRPV4 and other ion channels to the plasma membrane, leading to decreased intracellular calcium levels, subsequent induction of reduced cell volume and cortical stiffening, thereby promoting cell invasiveness and motility.

Results

Cell Crowding Enhances Invasiveness in High-Grade DCIS Cells

Cell crowding reflects a common condition in tumor microenvironments for ADH and DCIS, arising from aberrant cell proliferation within spatially constrained intraductal spaces. We examined the influence of this prevalent environmental factor on the invasiveness of these cells in vitro. To conduct this investigation, we assembled a panel of cell lines derived from the normal breast epithelial cell line MCF10A, including its H-RAS mutation-driven derivatives associated with various pathologies28: MCF10AT1, MCF10DCIS.com, and invasive cells MCF10CA1a. MCF10AT1 resembles ADH15, MCF10DCIS.com mimics high-grade DCIS26, and MCF10CA1a represents a malignant invasive cancer that was observed to form metastatic lesions in a mouse xenograft29. The current classification of DCIS relies on histological factors such as cell growth patterns and cytonuclear features30,31. Comedo-DCIS is a histologic subtype, which is characterized by apoptotic cell death, representing a high-grade DCIS with higher invasive potential than those of lower-grades25. MCF10DCIS.com forms comedo-type DCIS lesions that can spontaneously transition to invasive cancer when xenografted26.

To compare cell invasiveness under normal cell density and cell crowding conditions in vitro, we opted for a modified 2D matrix degradation assay. This approach allowed us to quantify overall cell invasiveness by normalizing it with the total cell number, thereby accounting for differences in cell densities. We chose this method over transwell, 3D Matrigel, or spheroid assays, where quantifying cell invasiveness as a function of cell density is challenging. By modifying an existing collagen-crosslinked polyacrylamide hydrogel matrix-based invasion assay32–34, we could determine the invasive cell fraction out of the total cell population by fluorescence imaging of the invasion gel bed and cell nuclei through a low magnification (4X) imaging (Fig. 1A; for a detailed procedure, see Suppl. Fig. 1) when cells were at normal density versus under cell crowding conditions.

Cell crowding selectively increases invasiveness in high-grade DCIS cells.

(A) We used a collagen-crosslinked polyacrylamide hydrogel matrix-based invasion assay to assess the effect of cell crowding on invasiveness. Representative images show gelatin-Alexa488 conjugates, where dark areas indicate cell invasion through degradation (green; Gelatin488), and DAPI staining marks cell locations (blue; DAPI) in a two-day invasion assay of MCF10DCIS.com cells under normal density (ND; upper panel) and overconfluent (OC; lower panel) conditions. "Masked" images are thresholded to produce positive masks applied to the “DAPI” images. Individual cell locations detected in “DAPI” images are marked with purple circles in “Detected points” images. The total number of cells within the field of view is counted from these points. By overlaying the mask and DAPI images, “Masked DAPI” images are obtained, and invaded cells are detected and represented by purple circles in “Masked points” images. The invasive cell fraction is calculated by the ratio of the number of invaded cells to the total number of cells (0.24 for ND and 0.59 for OC MCF10DCIS.com cells). These data show that cell invasiveness is enhanced by cell crowding. Scale bar = 100 μm. (B) Comparison of “Gelatin488” images of MCF10A (normal breast epithelial cells), MCF10AT1 (ADH-mimicking cells), MCF10DCIS.com (high-grade DCIS mimic), and MCF10CA1a (invasive breast cancer cells) between ND and OC conditions. MCF10DCIS.com cell invasion is significantly higher under cell crowding than under ND conditions. (C) Invasive cell fractions of these cells between ND (blue circles) and OC (red circles) conditions are compared, showing that cell crowding-induced increases in invasiveness are notable only in MCF10DCIS.com cells. We used the two-tailed Mann-Whitney U test, a nonparametric and unpaired statistical method, to compare differences between groups. ****: p < 0.0001, ***: p < 0.001, **: p < 0.01, *: p < 0.05, ns: p > 0.05, throughout the manuscript.

We observed alterations in live cell morphology and invasiveness as they progressed from normal density (ND: 40-70%) to full confluence (100%) and beyond. Our goal was to determine a distinct time window, termed overconfluence (OC), where cell crowding-induced changes level off. The OC time window was achieved as cells cultured for an additional 5-10 days after reaching 100% confluence, with the growth medium replaced twice per day to prevent acidification. We confirmed that the cells remained largely viable under crowding conditions, with minimal cell death (∼1.58%) similar to that observed in ND cells (∼0.85%). (Suppl. Fig. 2A-B). The schematic of the timeline to achieve the OC conditions, and low-magnification (4X) bright-field images at different time points from the ND to OC conditions are shown in Suppl. Fig. 3A-B. The time-dependent equilibration of cell invasiveness after the cells reached confluence is demonstrated in Suppl. Fig. 3C-D using our quantifiable collagen-crosslinked polyacrylamide hydrogel matrix-based invasion assay. Cell crowding mechanosensing significantly increased the invasive cell fraction of MCF10DCIS.com cells from ∼24% (ND cells) to 59% (OC cells). The enhanced invasiveness of MCF10DCIS.com cells under cell crowding was largely reversible. When OC cells were reseeded at normal density for invasion assays, their invasive cell fraction decreased to approximately 15%, slightly lower (p = 0.012) than the initial value of around 24% (Suppl. Fig. 3C, 3E). In contrast, other cell types such as normal breast epithelial cells MCF10A and MCF10AT1, which mimic ADH conditions in vivo, did not display invasiveness in ND conditions or enhanced invasiveness in OC conditions, suggesting insensitivity to cell crowding (Fig. 1B-C).

The invasive cell fraction of ND MCF10CA1a cells was already ∼80%, and the additional increase in invasiveness due to cell crowding was not readily discernible, with a slight increase to ∼82% under crowded conditions. These data suggest a striking and selective mechanosensing effect of cell crowding on cell invasiveness in MCF10DCIS.com cells. To ensure that acidification did not affect the invasiveness of MCF10DCIS.com cells despite the frequent replacement of the cell growth medium, we incubated ND cells for two days with acidified medium from cultures of OC MCF10DCIS.com cells. We observed that medium acidity did not alter cell invasiveness (Suppl. Fig. 3F), reinforcing that the increased invasiveness under OC conditions was induced by cell crowding.

Cell crowding reduces cell volume and stiffens MCF10DCIS.com cells

As cells became crowded, we observed a reduction in cell size. Previous reports indicate an inverse relationship between cell volume and cortical stiffness35,36, leading us to hypothesize that reduced cell volume would be accompanied by increased cortical stiffness. Research on glioma cell invasion underscores the critical role of hydrodynamic cell volume changes in penetration into surroundings37,38, suggesting that significant cell volume reduction facilitates cell invasion. Additionally, increased cortical stiffness is known to help cells overcome the physical barriers of the dense extracellular matrix39,40. We thus investigated whether cell crowding selectively induces pro-invasive cell volume reduction and cell stiffening in MCF10DCIS.com cells, thereby priming them for invasion.

To measure the cell volume of individual cells, we used confocal microscopy with a 60X oil immersion objective to obtain z-stack images of live or fixed cells stably expressing red fluorescent protein (RFP)41. Using a nanoindenter attached to the confocal microscope (Suppl. Fig. 4A), we extracted Young’s modulus of live individual cells to assess changes in cortical stiffness. Cell crowding significantly reduced cell volume (Fig. 2A, B) and increased cortical stiffness (Fig. 2C) in both MCF10DCIS.com and MCF10A cells. However, the changes in volume and stiffness between ND and OC conditions were more pronounced in MCF10DCIS.com cells (Suppl. Fig. 4B, C). The cell crowding-induced alterations in cell volume and stiffness were negligible in MCF10AT1 cells (Fig. 2A-C). In MCF10CA1a cells, cell crowding led to an increase in stiffness, but a decrease in volume was not evident. This was expected because the cell volume was already too small to be accurately measured with the available resolution (Fig. 2A-C). The notable plasticity in cell volume and stiffness changes observed in MCF10DCIS.com cells in response to cell crowding potentially underscores the critical link between this plasticity and mechanosensitive increases in cell invasiveness. We assessed cell volume changes only as an effector event of cell crowding, without measuring cell stiffness, because cell volume reflects the mechanical properties of the entire cell, while Young’s modulus can vary depending on the location of indentations in the plasma membrane42,43. The observed cell volume changes by cell crowding (Fig. 2A) depended on the ND cell volume, as OC cell volumes under crowding conditions were comparable across all cells. Consequently, the ND cell volume and the volume changes from ND to OC conditions exhibited an approximately linear relationship (R² ∼ 0.97; Suppl. Fig. 4D). This finding suggests that the ND cell volume could serve as an indicator of relative cell volume plasticity.

Cell Crowding induces significant cell volume reduction and stiffening in High-Grade DCIS Cells.

(A) Cell volume (V; mean and SD) differences between ND (blue circles) and OC (red circles) conditions of MCF10A, MCF10AT1, MCF10DCIS.com, and MCF10CA1a cells are plotted. The high-grade DCIS cell mimic, MCF10DCIS.com, shows a large volume reduction due to cell crowding. (B) Representative confocal microscopy images of RFP-coexpressing cells of the four cell types in ND and OC conditions. The images include x-y (left) and x-z (right) views, with scale bar = 10 μm. The large volume reduction of MCF10DCIS.com cells is evident. (C) Plots showing changes in cortical stiffness (mean and SD) measured by Young’s modulus (Y) using a nanoindenter, displaying significant cell stiffening of MCF10DCIS.com cells due to cell crowding. (D) Hyperosmotic conditions induced by PEG 300 treatment (darker blue circles for 2% = 74.4 mOsm/Kg; navy circles for 4% = 148.8 mOsm/kg) lead to dose-dependent cell volume reduction. (E) Treatment with 2% PEG 300 (darker circles) for two days significantly increased the invasiveness (mean and SD) of MCF10DCIS.com cells, similar to the OC case (red circles). ****: p < 0.0001, ***: p < 0.001, **: p < 0.01, *: p < 0.05, ns: p > 0.05.

To validate the pro-invasive nature of cell volume reduction in MCF10DCIS.com cells, we examined the effect of hyperosmotic cell volume reduction in ND cells using polyethylene glycol (PEG) 300 for 15 minutes. The cell volume reduction effects of this treatment were previously described36,44. PEG 300 effectively reduced ND MCF10DCIS.com cell volume in a dose-dependent manner, showing a greater reduction with 148.8 mOsmol/kg PEG 300 than with 74.4 mOsmol/kg PEG 300 (Fig. 2D). To assess the impact of PEG-induced cell volume reduction on cell invasiveness, we used collagen-crosslinked polyacrylamide hydrogel matrix-based invasion assay and exposed cells to 74.4 mOsmol/kg PEG 300 for 2 days. While treatment with 148.8 mOsmol/kg PEG 300 significantly reduced cell viability, cells treated with 74.4 mOsmol/kg PEG 300 remained viable for 2 days. Monitoring changes in cell invasiveness revealed that exposure of ND MCF10DCIS.com cells to 74.4 mOsmol/kg PEG 300 increased cell invasiveness (Fig. 2E), confirming the causal relationship between cell volume reduction and increased cell invasiveness.

Cell crowding induces TRPV4 relocation to plasma membrane in MCF10DCIS.com cells

Cell volume regulation typically depends on osmotic gradients that direct the movement of water across cell membranes45. This process is facilitated by ion flux modulation, which is controlled by ion channels and ion transporters located on the plasma membrane46–49. Notably, cells capable of achieving minimal cell volumes have been shown to successfully invade neighboring tissues38. Based on this, we speculated that the high cell volume plasticity observed in MCF10DCIS.com cells underlies their ability to penetrate the surrounding tissues. To achieve such efficient cell volume changes, cell crowding-induced mechanotransduction in MCF10DCIS.com cells may involve modulation of the number of ion channels or transporters on the plasma membrane.

To compare the plasma membrane-associated proteins between ND and OC conditions, we employed mass spectrometry and profiled those proteins using the streptavidin-pulled surface-biotinylated cell lysates. Fig. 3A shows the relative protein densities on the plasma membrane under ND (blue bars) and OC (red bars) conditions in MCF10DCIS.com cells. The figure highlights the top 25 proteins that exhibit over a 5-fold increase in expression on the plasma membrane in the OC condition compared to the ND condition (triangle plots). The gene names of these 25 proteins are presented in Fig. 3A (Suppl. Table 1 provides the corresponding protein names along with the corresponding fold increases). Notably, TRPV4, a member of the transient receptor potential family of ion channels known for its mechanosensitive properties50,51, showed a remarkable 153-fold increase in plasma membrane association in response to cell crowding, as illustrated in Fig. 3A. Additionally, we observed an increase in plasma membrane association of other ion channels, such as SCN11A (the alpha subunit of the voltage-gated sodium channel Nav1.9) with ∼42-fold enrichment, and KCNN4 (the small-conductance calcium-activated potassium channel 3 SK3) showing ∼6-fold increase, as shown in Fig. 3A. To evaluate the relative plasma membrane association of proteins under OC versus ND conditions, we performed a comparable mass spectrometry analysis of cell-surface biotinylated cell lysates for MCF10A, MCF10AT1, and MCF10CA1a cells. The top 25 gene/protein names showing increased plasma membrane associations in these cells are listed in Suppl. Table 2. Unlike in MCF10DCIS.com cases, there was no cell-crowding-induced relocation of ion channels and ion transporters to the plasma membrane in MCF10A and MCF10AT1 cells. However, in MCF10CA1a cells, ion transporters such as ATP2B4 showed a ∼37-fold greater plasma membrane association under crowding conditions. This suggests that plasma membrane relocations of these ion channels and ion transporters in response to cell crowding selectively occurred in MCF10DCIS.com cells and, to an extent, in invasive ductal cancer MCF10CA1a cells.

Cell crowding induces TRPV4 relocation to plasma membrane in MCF10DCIS.com cells.

(A) Mass spectrometry data showing proteins enriched in the plasma membrane (PM) >5-fold (fold changes represented using triangle plots; OC/ND ratio on the right axis) when cells are under OC (red bars) relative to ND conditions (blue bars). Ion channels are marked with red boxes, where TRPV4 shows about a 160-fold increased association with the plasma membrane under OC conditions. (B) Proteins near and on the plasma membrane were pulled down after cell surface biotinylation with streptavidin beads and immunoblotted for TRPV4. TRPV4 is significantly associated with the plasma membrane in OC MCF10DCIS.com cells. In MCF10CA1a cells, TRPV4 appears to be associated with the plasma membrane under both ND and OC conditions, with a slight increase under OC conditions. (C) Immunoblots of whole-protein lysates demonstrate similar overall TRPV4 protein levels across MCF10A cell derivatives, regardless of cell density. This indicates that the differing plasma membrane association of TRPV4 is due to trafficking changes, not expression level changes. GAPDH is used as a loading control. (D) Representative IF images by confocal microscopy show TRPV4 (red) localization compared to the control protein transferrin receptor (TfR; green) in MCF10A, MCF10AT1, MCF10DCIS.com, and MCF10CA1a cells under ND and OC conditions. DAPI (blue) staining was used for visualizing the nuclei. As observed in the biochemical data in (B-C), cell crowding induces the relocation of TRPV4 to the plasma membrane in MCF10DCIS.com cells. TRPV4 is associated with the plasma membrane in ND MCF10CA1a cells, with a clear elevated association in OC cells. Scale bar = 10 μm. (E) Plasma membrane-associated TRPV4 (%) is quantified for the four cell lines under ND and OC conditions by line analysis, showing a significant increase in both MCF10DCIS.com cells and MCF10CA1a cells due to cell crowding. (F) IF images showing that hyperosmotic conditions induced by PEG 300 (74.4 mOsm/Kg) treatment also relocate TRPV4 (red) to the plasma membrane in MCF10DCIS.com cells. TfR localization remains intact under hyperosmotic conditions. Increased relocation is also observed in MCF10CA1a cells. Scale bar = 10 μm. (G) The increased plasma membrane association of TRPV4 due to hyperosmotic stress is quantified by line analysis. Scale bar = 10 μm. (H) Representative regions of interest (ROIs) of TRPV4-stained immunohistochemistry (IHC) images in different pathology phenotypes. High-grade DCIS and invasive ductal cancer (IDC) ROIs clearly exhibit plasma membrane association of TRPV4. Two high-grade DCIS IHC images were acquired by two different people and both show plasma membrane-associated TRPV4. Scale bar = 20 μm. (I) Statistical results from independent histological evaluations of pathologies and TRPV4 distributions of 97 ROIs from 39 patient specimens indicate a high correlation (>70%) of plasma membrane association of TRPV4 with high-grade DCIS or IDC pathologies. Y/N: Yes/no, indicating both pathologists agreed that PM ion channels were present/absent. E: Equivocal, indicating only one pathologist agreed. Significantly high proportions of high-grade DCIS (75%) and IDC (73%) ROIs exhibited plasma membrane TRPV4 association, which was not observed in lower-risk cases.

TRPV4 plays a pivotal role in facilitating the passage of calcium ions (Ca2+) and is integral in detecting various forms of mechanical stresses, including temperature fluctuations and osmotic pressure52,53. Activation of TRPV4 channels elevates intracellular calcium levels, leading to cell volume increase via osmotic water influx52,54. Nav1.9 primarily contributes to the generation and propagation of action potentials in sensory neurons and is associated with various pain disorders55–57. SK3 channels are activated by elevated intracellular calcium levels and serve diverse physiological functions58,59. However, unlike TRPV4, Nav1.9 and SK3 channels are not generally classified as mechanosensors. Consequently, we focused on TRPV4, the mechanosensor channel displaying the most significant increase in the plasma membrane association under cell crowding conditions, to further investigate its role in promoting pro-invasive cell volume reduction in MCF10DCIS.com cells.

To confirm the selective plasma membrane association of TRPV4 in OC MCF10DCIS.com cells suggested by mass spectrometry results, we conducted immunoblots on surface-biotinylated lysates from all four cell types under both conditions. Fig. 3B and Suppl. Fig. 5A showed a notable association of TRPV4 with the plasma membrane in OC MCF10DCIS.com cells, which was significantly lower in ND cells. This association was not observed in MCF10A and MCF10AT1 cells under any conditions. Interestingly, in MCF10CA1a cells, TRPV4 associated with the plasma membrane under both conditions, with a slight increase in OC conditions. This increase suggests that these cells may also possess mechanosensing abilities, enabling them to respond to cell crowding. Additionally, immunoblots from whole-cell lysates showed consistent TRPV4 expression levels across different cell types and densities (Fig. 3C, Suppl. Fig. 5B), indicating that cell crowding influences ion channel trafficking without altering overall expression. This evidence strongly suggests that the relocation of TRPV4 to the plasma membrane is a mechanosensitive response to cell crowding.

To investigate the redistribution of TRPV4 in response to cell crowding, we performed IF imaging with confocal microscopy across all four cell types under both ND and OC conditions. The binding specificity of the TRPV4 antibody was confirmed through IF imaging (Suppl. Fig. 6A) and immunoblotting (Suppl. Fig. 6B) of MCF10DCIS.com cells treated with TRPV4 shRNA. As expected from the mass spectrometry results, TRPV4 (red) showed only modest localization to the plasma membrane in ND MCF10DCIS.com cells (Fig. 3D). In contrast, it was prominently associated with the plasma membrane in OC MCF10DCIS.com cells (Fig. 3D). In MCF10CA1a cells, TRPV4 was also associated with plasma membrane under ND conditions, but cell crowding also increased the association significantly as shown in MCF10DCIS.com cells. Plasma membrane associations were verified using the plasma membrane marker DiIC18(3), which we previously used to visualize the plasma membrane60. IF images showed that TRPV4 successfully colocalized with DiIC18(3) in OC MCF10DCIS.com and OC MCF10CA1a cells (Suppl. Fig. 7A). However, the non-interacting protein transferrin receptor (TfR; green) remained similarly distributed in the OC cells compared to the ND cells, indicating that the protein trafficking changes induced by cell crowding are specific to TRPV4 (Fig. 3D).

To investigate whether other ion channels identified through mass spectrometry and a well-known mechanosensory channel PIEZO161,62, exhibit similar behavior, we conducted IF imaging on ND and OC MCF10DCIS.com cells targeting KCNN4 and PIEZO1. Unfortunately, due to the absence of specific antibodies, we were unable to examine SCN11. In ND MCF10DCIS.com cells, KCNN4 was observed to be mostly cytosolic, while PIEZO1 was mildly but slightly more associated with the plasma membrane than TRPV4 under ND conditions. However, under OC conditions, both PIEZO1 and KCNN4 also relocated to the plasma membrane, albeit to a lesser extent than TRPV4 (Suppl. Fig. 7B). The varying degrees of plasma membrane association of PIEZO1 and KCNN4 are quantified using line analysis, as summarized in Supple. Fig. 7C.

As expected, in ND MCF10A, MCF10AT1, and MCF10DCIS.com cells, TRPV4 was modestly present at the plasma membrane but was predominantly intracellular, with notable enrichment in the nuclei (Fig. 3D). However, in OC MCF10DCIS.com cells, approximately 80% of TRPV4 was located in the plasma membrane based on line analysis (Fig. 3E). In contrast, in OC MCF10A and OC MCF10AT1 cells, TRPV4 remained mostly cytosolic but migrated out of the nuclei (Fig. 3D). TRPV4 association with the plasma membrane was observed in both ND and OC MCF10CA1a cells, with an observable increase under OC conditions, further confirming the immunoprecipitation results shown in Fig. 3B. Plasma membrane-associated TRPV4 was quantified using line analysis and plotted in Fig. 3E. Refer to Suppl. Fig. 8A for plots of the relative TRPV4 associations with the plasma membrane, cytosol, and nucleus between ND and OC conditions in all four cell types. These IF imaging results underscore the mechanosensitive plasma membrane relocation of ion channels in MCF10DCIS.com, and to an extent, in MCF10CA1a cells, contrasting with the observed insensitivity of MCF10A and MCF10AT1 cells to cell crowding.

To assess the relationship between cell crowding-induced cell volume reduction and plasma membrane relocation of TRPV4, we investigated whether hyperosmotic conditions, which reduce cell volume without cell crowding, would also result in TRPV4 relocation. We subjected ND cells to the same PEG 300 condition (74.4 mOsm/kg PEG for 15 minutes) used in Fig. 2D to induce cell volume reduction. In ND MCF10DCIS.com cells, such hyperosmotic cell volume reduction prompted significant plasma membrane relocation of TRPV4, a response not observed in MCF10A and MCF10AT1 cells, but seen in MCF10CA1a cells (Fig. 3F). Quantitative line analysis confirmed these findings by assessing the relative TRPV4 association with the plasma membrane (Fig. 3G; Suppl. Fig. 8B for plots of the relative TRPV4 association with the plasma membrane, cytosol, or nucleus). We confirmed that the seeming plasma membrane-associated TRPV4 under the hyperosmotic condition was indeed localized at the plasma membrane by using an extracellular domain antibody against TRPV4 in live MCF10DCIS.com cells, both untreated and treated with 74.4 mOsm/Kg PEG 300. As expected, significant antibody binding was observed only in PEG 300-treated cells, where intracellular TRPV4 had relocated to the plasma membrane unlike in untreated live cells (Suppl. Fig. 9A-B). This finding underscores the relationship between mechanosensitive cell volume reduction and plasma membrane relocation of TRPV4, highlighting that both cell crowding and hyperosmotic stress lead to the same effect, particularly pronounced in MCF10DCIS.com cells.

We investigated whether TRPV4 relocation to the plasma membrane induced by cell crowding is reversible, as suggested by its impact on invasiveness (Suppl. Fig. 3F). To test this, previously OC MCF10DCIS.com cells were reseeded under ND conditions. We then assessed TRPV4 localization via immunofluorescence (IF) imaging to determine if most channels returned to the cytoplasm with only a modest localization at the plasma membrane, and could be relocated to the plasma membrane under mechanical stress, such as hyperosmotic conditions. Consistent with their initial ND state, reseeded ND MCF10DCIS.com cells displayed intracellular TRPV4 distribution (Suppl. Fig. 10A). Upon exposure to hyperosmotic stress (74.4 mOsm/Kg PEG300), TRPV4 was again relocated to the plasma membrane (Suppl. Fig. 10B). These findings, quantified through line analysis (Suppl. Fig. 10C), demonstrate that the mechanosensing response of MCF10DCIS.com cells is reversible.

Patient tissue analysis shows selective plasma membrane association of TRPV4 in high-grade DCIS cells

MCF10DCIS.com cells represent a basal cell model for high-grade DCIS cells driven by HRAS mutation26,63. However, the majority of patient-derived DCIS cells originate from the luminal cell population and lack the HRAS mutation63. Moreover, there are limited options for patient-derived DCIS cell line models that have not linked to well-defined pathology64,65. Thus, to seek generality of our finding about the selective association between plasma membrane relocation and high-grade DCIS pathology, we examined 39 breast cancer patient tissue blocks to assess the selectivity in vivo. We designed an early-stage retrospective study by selecting various breast tissue pathologies that range from benign to invasive cancer. Those pathologies include usual ductal hyperplasia, papilloma, columnar changes, ADH, low-to high-grade DCIS, and invasive ductal carcinoma (IDC). We also incorporated normal regions for comparison. Hematoxylin and eosin (H&E) staining was used to visualize tissue sections from each patient, excluding samples with prior cancer diagnoses or drug treatments. In total, 97 regions of interest (ROIs) from H&E-stained sections of 39 patient tissue blocks were selected. Two pathologists independently assessed TRPV4 distribution patterns at the single-cell level in corresponding ROIs in the immunohistochemistry (IHC) images (Fig. 3H). A detailed methodology is described in the Methods section and more example images are shown in Suppl. Fig. 11. High-grade DCIS cells, as depicted in Fig. 3H, clearly demonstrated plasma membrane-associated TRPV4, a feature absent in lower-grade DCIS cells (intermediate-and low-grade). This distinction sets them apart not only from cells with ADH and benign pathologies but also from lower-grade DCIS cells. As demonstrated by the IF images in Fig. 3D for MCF10CA1a (invasive cell mimic), IDC cells exhibited notable plasma membrane TRPV4, suggesting that they possess a similar pro-invasive mechanotransduction capability to high-grade DCIS cells.

We found that the sensitivity and specificity for plasma membrane TRPV4 in high-grade DCIS cells were 0.75 ± 0.19 (15/20) and 0.98 ± 0.03 (61/62) respectively (95% confidence interval) (Fig. 3I). It is important to note that the specificity calculation excluded IDC cases. Notably, even in less aggressive pathologies, a significant amount of TRPV4 was localized in the nuclei, as shown for MCF10A and MCF10AT1 cells in the IF images in Fig. 3D. This underscores our interpretation from the in vitro results in Fig. 3A-D that increased plasma membrane association of TRPV4 in high-grade DCIS cells results from changes in protein localization through trafficking alterations, rather than differences in expression levels. These initial in vivo results clearly demonstrate the selective and specific association of TRPV4 with the plasma membrane in high-grade DCIS cells. Notably, this association is absent in lower-grade DCIS, ADH, benign, and normal cells, thereby confirming our in vitro findings using MCF10A cell derivatives. These results suggest a potentially critical role for TRPV4 in the progression of high-grade DCIS.

Cell crowding inhibits ion channels and triggers their plasma membrane relocations

A calcium-permeant ion channel like TRPV4 is known to influence intracellular calcium dynamics54. While activation of TRPV4 typically elevates calcium levels and this potentially increases cell volume66–68, the impact of its inhibition is less clear, given the multifaceted nature of calcium signaling in cell volume control69. This complexity is compounded by compensatory cellular mechanisms and the involvement of other ion channels in response to altered calcium homeostasis46. Considering these factors, we hypothesized that cell crowding might inhibit calcium-permeant ion channels that are constitutively active at the plasma membrane, including TRPV4, which would then lower intracellular calcium levels and subsequently reduce cell volume via osmotic water movement. To test this, we employed the Fluo-4 calcium assay70 to compare relative intracellular calcium levels of MCF10DCIS.com cells between ND and confluent (Con) conditions through 4X confocal microscopy imaging with 488 nm excitation. For this calcium assay, we opted for confluent conditions instead of OC conditions to collect the Fluo-4 signal from confluent monolayers of MCF10DCIS cells, which demonstrates comparable cell-crowding induced TRPV4 relocation to the plasma membrane (Suppl. Fig. 12A) as observed in the OC condition. This choice was made because more than one layer of cells was occasionally observed in OC conditions, which would yield overestimation of crowding-dependent intracellular calcium levels. The intracellular Fluo-4 signal under Con conditions was significantly lower than that under ND conditions, as shown in the fluorescence images of ND versus Con cells in Fig. 4A. Time-dependent Fluo-4 images were acquired over 25 minutes. For ND cell images, the average Fluo-4 intensity values were calculated from 10-15 selected cells (highlighted in the white box in Fig. 4A). For Con cell images, intensity values were derived from all cells within the entire 50 μm by 50 μm field of view, as depicted in Fig. 4A. After background subtraction, the intensity values were plotted in Fig. 4B. The temporal Fluo-4 intensity profiles from both ND and Con cells remained largely constant over the measured 20-minute period. However, Con cells exhibited significantly lower Fluo-4 intensity compared to ND cells, indicating reduced intracellular calcium levels. Lower Fluo-4 intensity in Con cells than in ND cells was not due to limited Fluo-4 reagent in Con samples. Within cell clusters, cells experiencing crowding exhibited notably lower Fluo-4 levels compared to those at the periphery (Suppl. Fig. 12B), suggesting that cell crowding had resulted in a decrease in intracellular calcium levels, likely due to the mechanosensitive inhibition of calcium-permeable channels like TRPV4.

Cell crowding inhibits ion channels and triggers their plasma membrane relocations.

(A) To compare intracellular calcium (Ca²⁺) levels, we used a Fluo-4 AM assay, where green fluorescence intensity increases with higher intracellular Ca²⁺ levels. Calcium levels are significantly lower in confluent (Con) MCF10DCIS.com cells. (B) The temporal progression of averaged Fluo-4 intensity in ND MCF10DCIS.com cells (blue curve) in the box shown on the left image is compared with that of Con cells (red curve). Fluo-4 intensity is consistently lower in Con cells than in ND cells for approximately 25 minutes (200 ms acquisition time and 1 s time interval). (C) Fluo-4 intensity reduction due to cell crowding is significant in MCF10DCIS.com cells. (D-H) Pharmacological inhibition of TRPV4 with 1 nM GSK219 generates dips in the Fluo-4 signal. Fluo-4 images at the baseline (t1) and the dip (t2) post 1 nM GSK219 are compared in MCF10DCIS.com cells between ND (D) and Con (F) conditions. The Fluo-4 intensity time traces are compared between ND (blue; E) and Con (red; G) conditions, showing that the magnitude of the dip (marked as ΔCa) is significantly lower in Con cells, where TRPV4 activity is largely inhibited under cell crowding conditions. Notably, ΔCa increased with higher GSK219 doses (1 nM vs. 0.2 nM), but remained significantly lower in Con MCF10DCIS.com cells, with smaller changes observed under the 0.2 nM GSK219 condition. (I-M) TRPV4 activation with 0.2 pM GSK101 leads to a small spike in ND cells (I, J). However, in Con cells, the same GSK101 treatment leads to a notable spike in Fluo-4 intensity, indicating that TRPV4 inhibition and subsequent relocation to the plasma membrane by cell crowding primes the ion channels for activation. GSK101 treatment also leads to a dose-dependent increase in the spike magnitude with higher GSK101 concentrations (0.05 pM: 101L; 0.2 pM: 101H), which is strikingly high in Con MCF10DCIS.com cells. (N-Q) TRPV4 activation status-dependent intracellular localization changes. (N) IF images of TRPV4 (red) and TfR (green) in ND MCF10DCIS.com cells show that GSK101 does not increase plasma membrane association of TRPV4. However, GSK219 significantly relocates TRPV4 to the plasma membrane in a dose-dependent manner, similar to ND cells treated with 74.4 mOsm/Kg PEG 300. (O) In OC cells, while GSK219 does not significantly alter TRPV4 association with the plasma membrane, GSK101 depletes plasma membrane TRPV4 in a dose-dependent manner, suggesting that TRPV4 activation status affects its trafficking. Relative plasma membrane associations with different treatments are quantified for ND (P) and OC (Q) cells using line analysis. (R) The magnitudes of Fluo-4 spikes by GSK101 and dips by GSK219 show a linear relationship (R² ∼ 0.69), indicating a negative correlation between them. This reinforces the observation that TRPV4 inhibition increases its association with the plasma membrane, while activation shows the reverse effect. (S-V) Compared to the Fluo-4 intensity in control MCF10DCIS.com cells (S, T), shRNA showed similar Fluo-4 levels (U, V). However, hyperosmotic stress by 74.4 mOsm/Kg PEG 300 (light gray box) led to a noticeable spike only in control ND cells. Additionally, cell crowding conditions (Con) led to a decreased Fluo-4 level (at t1 baseline in the image in S and red time trace in T) and a reduction in Fluo-4 level in shRNA-treated MCF10DCIS.com cells (t1; U) compared to control cases (S, T), as shown in the image and time trace (t1; V). (W) Relative Fluo-4 time-averaged intensities are plotted for individual control ND (blue) vs. Con (red) cells, and shRNA-treated ND (semi-transparent blue) vs. Con (semi-transparent red) cells. Intracellular calcium levels in shRNA ND cells are lower than those in control ND cells, reflecting the reduced number of TRPV4 channels. The decrease in calcium levels by crowding (Con) in shRNA cells is clearly lower than in control cells, reflecting the importance of TRPV4 in mechanosensing cell volume reduction. (X) PEG 300-induced calcium spikes are significantly lower in shRNA cells (semi-transparent gray) than in control cells (gray), reinforcing TRPV4’s crucial role in MCF10DCIS.com mechanotransduction. (Y) TRPV4 silencing significantly reduced the mechanosensing cell volume reduction effect. Control ND cells underwent a 48% volume reduction in response to 74.4 mOsm/Kg PEG 300, whereas TRPV4-silenced cells reduced their volume by only 27%.

To investigate whether TRPV4 inhibition was involved in decreased intracellular calcium levels and subsequent relocation of TRPV4 to the plasma membrane, we employed TRPV4-specific pharmacological agents to alter TRPV4 activity and modulate calcium concentrations in ND MCF10DCIS.com cells without applying cell crowding conditions. To inhibit TRPV4 activity, we used TRPV4 inhibitor (GSK2193874; GSK219)71 at 0.2 and 1 nM, which we confirmed to have insignificant effects on cell viability in both ND and OC conditions for two days (Suppl. Fig. 13A). GSK219 treatments immediately caused a modest dip in the Fluo-4 signal, or intracellular calcium level, followed by a recovery, indicating cellular homeostatic activity. Fig. 4D shows Fluo-4 images taken before and at the signal dip after applying 1 nM GSK219, where the dip in the Fluo-4 signal is not visually apparent. However, the temporal profile of Fluo-4 intensity in Fig. 4E, which corresponds to the time points marked in Fig. 4D (t1: baseline and t2: dip), clearly shows the dip at t2, indicated by ΔCa (the vertical dashed line between the dip and baseline). This modest Fluo-4 dip at t2 represents the inhibition of activity by GSK219 on a small population of constitutively active TRPV4 channels at the plasma membrane under ND conditions.

In Con cells, 1 nM GSK219 caused a smaller dip in Fluo-4 intensity compared to the one observed in ND cells, with no subsequent changes. This is likely due to fewer constitutively active TRPV4 at the plasma membrane in Con cells than in ND cells. Fig. 4F shows Fluo-4 images at the baseline and the dip, and Fig. 4G presents the temporal Fluo-4 intensity profile, highlighting the two time points. Indeed, the dip in Fluo-4 intensity induced by GSK219 was smaller for Con MCF10DCIS.com cells (mean ΔCa ∼ -39 ± 9) than for ND cells (mean ΔCa ∼ -49 ± 18), as illustrated in Fig. 4H. The dose-dependent (0.2 nM and 1 nM GSK219, labeled 219L and 219H) and cell density-dependent (ND and Con) Fluo-4 signal dips are summarized by the ΔCa values plotted in Fig. 4H, showing that the most significant TRPV4 inhibition occurred in ND MCF10DCIS.com cells treated with 1 nM GSK219. These findings suggest that a substantial portion of TRPV4 channels relocated to the plasma membrane under cell crowding was inactive, and some constitutively activeTRPV4 channels already present in the membrane became inactive as a result of cell crowding.

Administering a TRPV4 activator (GSK1016790A; GSK101)72,73 immediately caused a notably greater spike in Fluo-4 signal in Con cells than in ND cells, evidencing that a large number of TRPV4 channels at the plasma membrane were inactive in Con cells. In Con cells, 0.2 pM GSK101 led to a large Fluo-4 signal spike with ΔCa ∼ +2,348 ± 597, while in ND cells, ΔCa was just ∼ +324 ± 130. This difference is depicted in the Fluo-4 images (Fig. 4I for ND and Fig. 4K for Con) and the time-dependent Fluo-4 intensity plots (Fig. 4J for ND and Fig. 4L for Con). The dose (101L and 101H representing 0.05 pM and 0.2 pM GSK101) and cell-density (ND and Con) dependent Fluo-4 signal spikes are summarized in the ΔCa values, as shown in Fig. 4M. The results show the significant GSK101 dose-responsive activation of previously inactive TRPV4 in Con MCF10DCIS.com cells, confirming that a large number of inactive TRPV4 channels at the plasma membrane were primed for activation in Con cell.

Based on the differing effects of GSK219 and GSK101 on immediate intracellular calcium responses, along with lower calcium levels in Con cells compared to ND cells, we hypothesized that cell crowding induces TRPV4 inhibition, leading to reduced intracellular calcium levels, which in turn triggers the relocation of TRPV4 to the plasma membrane as a compensatory mechanism. To test this hypothesis, we examined TRPV4 localization changes in response to GSK219 or GSK101 in a dose-dependent manner in both ND and OC MCF10DCIS.com cells using IF imaging (Fig. 4N for higher GSK101/219 concentrations; Suppl. Fig. 13B for both concentrations). Indeed, treating ND cells with the TRPV4 inhibitor GSK219 for 1 hour resulted in significant dose-dependent TRPV4 relocation to the plasma membrane (∼53 ± 7% at 0.2 nM and ∼62 ± 7% at 1 nM), compared to control ND cells (∼15 ± 6%), as shown by line analysis (Fig. 4P). These results indicate that TRPV4 inhibition induces its strategic relocation to the plasma membrane, priming it for rapid activation to counteract the effects of channel inhibition.

Confirming our observation of moderate Fluo-4 spikes induced by GSK101 in ND MCF10DCIS.com cells, which implies a small number of constitutively active TRPV4 channels (Fig. 4J), the plasma membrane-associated TRPV4 levels in ND cells remained largely unchanged following GSK101 treatment (Fig. 4N, 4P). However, in Con cells, where a large number of inactive TRPV4 channels are likely located at the plasma membrane, GSK101 treatment notably reduced plasma membrane-associated TRPV4 in a dose-dependent manner through internalization (Fig. 4O, 4Q), consistent with previous findings65. These data suggest that plasma membrane TRPV4 levels were largely regulated by the channel activity status. Specifically, channel activation led to the internalization of TRPV4, while channel inhibition promoted the relocation of TRPV4 to the plasma membrane. This relationship was approximately linear as shown in Fig. 4R, where the intracellular calcium differences compared to that of ND cells were negatively associated with plasma membrane fraction of TRPV4 (R2 ∼ 0.69).

Contrary to our expectation, GSK219 treatment in Con cells slightly reduced TRPV4 levels at the plasma membrane (Fig. 4O, 4Q). Moreover, higher doses of GSK219 increased nuclear TRPV4 levels in Con cells without affecting cytoplasmic levels (Suppl. Fig. 13B). These findings suggest that further TRPV4 inhibition under crowding conditions triggers a distinct trafficking alteration. Recent studies have implicated nuclear TRPV4 in regulating nuclear Ca2+ homeostasis and Ca2+-regulated transcription74. In light of this study and our findings, TRPV4 may relocate to the nucleus as a compensatory mechanism to maintain nuclear calcium regulation. This relocation could reflect an adaptive response to preserve calcium-dependent transcriptional programs or other nuclear processes essential for cell survival under mechanical stress.

Given the role of TRPV4 in osmoregulation53,75 and its significant relocation to the plasma membrane under hyperosmotic conditions (∼55% plasma membrane association by line analysis; Fig. 3E, 4P) in ND MCF10DCIS.com cells (∼15% plasma membrane association by line analysis), we explored the potential involvement of TRPV4 inhibition during this relocation. Using the Fluo-4 assay, we observed an initial calcium spike following 74.4 mOsm/kg PEG 300 in ND MCF10DCIS.com cells due to osmotic water outflow, which was followed by a homeostatic relaxation aimed at restoring calcium levels (Suppl. Fig. 14A). This relaxation likely involved the inhibition of ion channels like TRPV4 and other ion transporters, followed by their relocations to the plasma membrane. Indeed, we confirmed the relocation of other ion channels that responded to cell crowding, including KCNN4 and PIEZO1, to the plasma membrane following the same PEG 300 treatment (Suppl. Fig. 14B for IF images and line analysis results). The data suggest that the increased association of TRPV4 with the plasma membrane under hyperosmotic stress may reflect an adaptive inhibition response to water outflow.

To investigate the role of TRPV4 in mechanotransduction, we used shRNA to silence TRPV4 gene expression, resulting in a ∼50% reduction in TRPV4 protein levels (Suppl. Fig. 6B). We then assessed the effects of TRPV4 silencing on intracellular calcium levels in ND and Con conditions, as well as the immediate calcium response to hyperosmotic stress induced by 74.4 mOsm/Kg PEG 300. In ND cells, hyperosmotic stress triggered a significant calcium spike, as shown in the Fluo-4 image (middle; Fig. 4S) and time trace (blue; Fig. 4T). In contrast, TRPV4 silencing in ND cells led to a lower intracellular calcium level (left; Fig. 4U) and a reduced calcium spike in response to hyperosmotic stress (middle; Fig. 4U, and semi-transparent blue; Fig. 4V). Furthermore, the effect of TRPV4 silencing on intracellular calcium levels was less pronounced in Con cells (semi-transparent red; Fig. 4V) compared to the ND counterpart (semi-transparent blue; Fig. 4V). The relative intracellular calcium levels are summarized in Fig. 4W.

Notably, TRPV4 silencing significantly reduced the immediate calcium spike as summarized in Fig. 4X. This calcium spike was attributed to increased calcium concentration due to hyperosmotically reduced cell volume (Suppl. Fig. 14A). Thus, TRPV4 may play a critical role in mechanotransduction, enabling cells to sufficiently reduce their volume in response to mechanical stress (large volume plasticity). To further investigate this, we examined the effect of TRPV4 silencing on ND MCF10DCIS.com cell volume reduction in response to hyperosmotic stress. We indeed found that control ND cells underwent a 48% volume reduction in response to 74.4 mOsm/Kg PEG 300, whereas TRPV4-silenced cells only reduced their volume by 27% (Fig. 4Y). Notably, this reduced cell volume reduction in TRPV4-silenced cells is consistent with the idea that cell volume reduction-induced increased calcium concentration contributes to the calcium spike, reinforcing the concept that TRPV4 augments the ability of cells to mechanotransduce.

Cell crowding-induced inactive TRPV4 relocation to the plasma membrane indicates channel inhibition-mediated volume reduction and enhanced invasiveness and motility

To elucidate the relationships between TRPV4 inhibition, cell volume reduction, and invasiveness increase under cell crowding, we investigated the effects of TRPV4 inhibitors on these processes. We examined the effects of 2-day treatments with GSK101 and GSK219 on the cell volumes of ND and OC MCF10DCIS.com cells. This duration was chosen to correlate the treatments with changes in cell volume and invasiveness, as observed in response to cell crowding Under ND conditions, activating TRPV4 with GSK101 (0.05 or 0.2 pM for 2 days), which led to only modest calcium spikes and largely unaltered TRPV4 distribution (Fig. 4N; Suppl. Fig. 13B), did not significantly change cell volume (Fig. 5A). Conversely, TRPV4 inhibition by GSK219 treatment (1 nM for 2 days) that induced greater calcium changes (dips rather than spikes) and a significant plasma membrane relocation of TRPV4, led to a noticeable cell volume reduction (Fig. 5A). 0.2 nM GSK219 treatment had a negligible impact on ND cell volume (Fig. 5A). The GSK219 effect in ND MCF10DCIS.com cell volume was similar to that observed under hyperosmotic conditions by 74.7 mOsm/Kg PEG 300 (Fig. 5A).

Cell crowding-induced plasma membrane TRPV4 association scales with cell volume reduction and increases in invasiveness and motility

A-C. MCF10DCIS.com cell volume changes with TRPV4 inhibition and activation. (A) In ND MCF10DCIS.com cells, GSK101, which did not alter plasma membrane association of TRPV4, did not affect cell volume. Conversely, GSK219, which increased such association in a dose-dependent manner, reduced cell volume, with the effect of 1 nM GSK219 (219H) being similar to that of 74.4 mOsm/Kg PEG 300. (B) Under OC conditions, GSK101, which leads to significant Fluo-4 spikes, increased cell volume in a dose-dependent manner, while GSK219 and PEG only mildly reduced cell volume. (C) Cell volume changes in MCF10DCIS.com cells show an inverse relationship with plasma membrane association of TRPV4, reflecting the activation status of the channel (R² = 0.59). (D-F) Cell invasiveness increases with cell volume reduction and greater plasma membrane association of TRPV4. (D) Cell invasiveness significantly increases with GSK219 under ND conditions. (E) GSK101 under OC conditions shows a notable decrease in cell invasiveness in a dose-dependent manner. (F) Plasma membrane association of TRPV4 predictably reports cell invasiveness (R² ∼ 0.69), while cell invasiveness and cell volume are inversely related, reinforcing our observation that cell volume reduction promotes cell invasiveness. (H-M) To assess if cell motility also follows the trend of cell invasiveness, we performed a single-cell motility assay by tracking nuclear WGA in individual live cells every 5 s for 25 min. (H) Representative trajectories of individual cells are color-coded to reflect displacement at each time interval. Compared to untreated ND cells, 0.2 pM GSK101 treatment slowed overall cell diffusion, while 1 nM GSK219 and 74.4 mOsm/Kg PEG 300 treatments increased cell diffusion. ShRNA TRPV4 (Sh-ctrl) had no effect on cell motility under ND conditions. Furthermore, treatment with 74.4 mOsm/Kg PEG 300 failed to increase cell diffusivity (D) in shRNA-treated cells (Sh-PEG). Scale bar = 200 μm. Using single-cell analysis, we quantified cell diffusivity (D) and speed (v; movement directionality). (I) GSK101 treatment significantly reduced D, while GSK219 and PEG 300 notably increased D. ShRNA TRPV4 did not alter ND cell D. (J) GSK101 treatment increased v, while GSK219 and PEG treatments significantly increased D but those did not affect directionality. ShRNA ND cell directionality remained similar to that of control cells and was unchanged by PEG treatment. (K) Like cell invasiveness, cell motility (D) positively scales with plasma membrane association of TRPV4 (R² ∼ 0.73). (L) Cell motility (D) inversely relates to cell volume (R² ∼ 0.89). (M) Cell motility (D) and cell invasiveness show a strong linear relationship (R² ∼ 0.85), enabling the use of cell motility measurements to assess overall cell invasiveness.

Under OC conditions, GSK101 triggered significant, dose-dependent calcium spikes, leading to corresponding increases in cell volume (Fig. 5B). In contrast, GSK219 had a minimal impact on intracellular calcium levels and did not alter TRPV4 distribution at the plasma membrane, resulting in a slight reduction in cell volume at 1 nM (Fig. 5B). Similarly, PEG 300 induced a small change in cell volume, comparable to the 1 nM GSK219 treatment (Fig. 5B). As expected, we observed a significant negative linear correlation between plasma membrane-associated TRPV4 (% PM TRPV4) and cell volume (R² ∼ 0.59) (Fig. 5C). These findings (Fig. 5A-B) demonstrate that increased TRPV4 inhibition and subsequent intracellular calcium reduction in response to cell crowding are associated with decreased cell volume in MCF10DCIS.com cells.

We investigated the connections between TRPV4 activity, plasma membrane association, cell volume changes, and cell invasiveness. Using a collagen-crosslinked polyacrylamide hydrogel matrix-based invasion assay, we evaluated changes in invasiveness after 2-day treatments with GSK101 or GSK219. In ND MCF10DCIS.com cells, 0.2 pM GSK101 mildly suppressed cell invasiveness, while 0.05 pM GSK101 had no effect (Fig. 5D). In contrast, GSK219 increased the invasive cell fraction in a dose-dependent manner under ND conditions, mirroring changes in Fluo-4 signals and plasma membrane-associated TRPV4 (Fig. 5D). Under OC conditions, GSK101 increased cell volume and suppressed cell invasiveness dose-dependently (Fig. 5E). Conversely, GSK219 and PEG 300, which slightly reduced cell volume, increased cell invasiveness (Fig. 5E). Representative invasion images are shown in Suppl. Fig. 15A. These results establish a clear relationship between TRPV4 activity, plasma membrane-associated TRPV4, cell volume, and cell invasiveness. Notably, we observed a significant positive linear correlation between plasma membrane-associated TRPV4 and cell invasiveness (Fig. 5F; R² ∼ 0.69) and a negative association between cell volume and cell invasiveness (Fig. 5G; R² ∼ 0.89).

Cell invasiveness refers to the ability of cells to penetrate and move within their surrounding tissues or environments. As expected from the increased cortical stiffness of OC MCF10DCIS.com cells (Fig. 2C), we observed that OC MCF10DCIS.com cells underwent cytoskeletal rearrangements, including the increased formation of stress fibers (Suppl. Fig. 15B), which likely enhanced cell motility75. To assess how TRPV4 inhibition and activation, which reduced and increased cell volume, respectively, affected the motility of ND MCF10DCIS.com cells and the role of these motility changes on cell invasiveness, we conducted single-cell tracking experiments similar to single molecule tracking methods60,76–80. We utilized wheat germ agglutinin (WGA) as a nuclear marker, which stained the cell nucleus within 1 hour of incubation with cells. We then analyzed the diffusivity (D) and directionality (v) of single cell movements. We tracked overall single-cell trajectories for 3 hr with 1 min intervals, marking cell displacement over each interval in different colors, ranging from navy for the slowest to red for the fastest (Fig. 5H). Consistent with the trend in cell invasiveness, TRPV4 activation with 1 pM GSK101 significantly reduced single-cell movements, while inhibition with 0.2 nM GSK219 and 74.4 mOsm/kg PEG 300 notably increased these movements. Using the single-cell motility assay, we assessed TRPV4’s role in increasing cell invasiveness through a mechanotransduction mechanism involving cell-volume reduction. Reducing TRPV4 expression (∼50%) via shRNA diminished the cell volume reduction effect of mechanosensing. We predicted that shRNA would also attenuate the increased cell motility effect of mechanosensing in MCF10DCIS.com cells. Indeed, while shRNA (sh-ctrl; Fig. 5H, 5I) did not alter MCF10DCIS.com cell motility, it disabled mechanotransduction capability of the cells as there was no increase in D and v induced by hyperosmotic stress using 74.4 mOsm/Kg PEG 300 (sh-PEG; Fig. 5H, 5I). This result reinforces the critical role of TRPV4 in increasing cell invasiveness through mechanosensing activity inhibition-induced cell volume reduction.

Interestingly, while GSK219 and hyperosmotic conditions promoted increased cell movement (D; Fig. 5I), they also reduced cell directionality (v; Fig. 5I). This reduction in directionality suggests that while the cells become more mobile, their movements are less targeted, potentially indicating a more random or exploratory behavior under mechanical stress. Remarkably, cell movement increased in proportion to the amount of TRPV4 at the plasma membrane (Fig. 5K; R² ∼ 0.73) and decreased with increased cell volume (Fig. 5L; R² ∼ 0.89). Cell movement and invasiveness showed a strong linear correlation (Fig. 5M; R² ∼ 0.85), suggesting that these behaviors are governed by the same underlying mechanisms. Thus, our motility assay can inform cell invasiveness at the single-cell level.

Mechanosensitive TRPV4 relocation to plasma membrane indicates enhanced invasiveness

We investigated whether the mechanosensitive relocation of TRPV4 to the plasma membrane can reliably predict a cell’s ability to undergo pro-invasive mechanosensitive cell volume reduction. To achieve this, we used changes in cell motility as markers for overall shifts in cell invasiveness, given the strong correlation observed between cell motility and invasiveness (Fig. 5M). Our hypothesis is that cells with such mechanotransduction capability, like MCF10DCIS.com cells, should show increased plasma membrane association of TRPV4 when inhibited with GSK219 or under hyperosmotic conditions, evidenced by increased cell motility (D) while demonstrating the opposite effect with GSK101 treatment.

For comparison, we employed three uncharacterized cell types: MDA-MB-231, a triple-negative invasive breast cancer cell line, and two recently developed patient-derived DCIS cell lines, ETCC-06 and ETCC-1064,65, which have not yet been histologically classified by grade. These cells were compared with positive control groups, including MCF10DCIS.com and MCF10CA1a, which were observed to possess pro-invasive mechanotransduction capability (Fig. 5D, 6A-6C). MCF10AT1 served as a negative control, as it lacks this capability (Fig. 6D-6F).

Pro-invasive cell volume reduction mechanotransduction pathway is indicated by TRPV4 inhibition and hyperosmotic stress-driven increases in TRPV4 association with the plasma membrane and cell motility, as well as TRPV4 activation-induced decreases in cell motility.

(A) MCF10CA1a cells respond to 15 minutes of PEG 300 (74.4 mOsm/kg) and OC conditions by relocating TRPV4 to the plasma membrane, increasing the channel’s association with the membrane. IF images show a predominantly intracellular TRPV4 (red) distribution in the ND control MCF10CA1a cells, whereas plasma membrane association of TRPV4 occurs in response to PEG 300 and OC. DAPI (blue) staining indicates the nucleus. Scale bars throughout Fig. 6 = 10 μm. (B) Line analysis results of plasma membrane-associated TRPV4 (%) show significant increases in plasma membrane TRPV4 (PM TRPV4) with 1 hour of GSK219 (1 nM), 15 minutes of 74.4 mOsm/kg PEG 300, and OC conditions. (C) Cell movement diffusivity (D) increases with GSK219 and PEG treatments, while movement directionality (v) increases with GSK101 (0.2 pM) but shows no other effects. (D-O) MCF10AT1 (D, E), MDA-MB-231 (G, H), ETCC-06 (J, K), and ETCC-10 (M, N) do not relocate TRPV4 to the plasma membrane in response to TRPV4 inhibition by GSK219, hyperosmotic conditions by PEG-300, or OC conditions, as evidenced by IF images (TRPV4: red; DAPI: blue) and line analysis results for plasma membrane associations of TRPV4. (F) None of these cells’ motility responded to PEG 300. However, their responses to TRPV4 activation (GSK101) and inhibition (GSK219) varied, suggesting different roles of TRPV4 in their cancer biology. MCF10AT1 diffusivity (D) significantly reduced with GSK219, while the rest of the conditions did not affect D. Movement directionality (v) was not affected by any treatment conditions. (G) MDA-MB-231 cell D or v did not alter with any treatment conditions, suggesting an insignificant role of TRPV4 in their cell motility. (L, M) Both ETCC-06 and ETCC-10 D increased with GSK101, but GSK219 also increased ETCC-06 diffusivity, while not altering ETCC-10 diffusivity. ETCC-06 directionality (v) increased with GSK101, but ETCC-10 directionality remained unchanged. (P) Plasma membrane association of TRPV4 (% PM TRPV4) positively scalex with cell movement D over a larger range for MCF10DCIS.com cells than for MCF10CA1a, reflecting the high cell volume plasticity observed in MCF10DCIS.com cells. This result suggests that both cell types have a pro-invasive mechanotransduction pathway. (Q) Such scaling is absent in MCF10AT1, MDA-MB-231, ETCC-06, and ETCC-10 cells. (R) The presence of such a mechanotransduction pathway is observed for MCF10CA1a and MCF10DCIS.com cells from the plot of greater than 2-fold increase in plasma membrane association of TRPV4 (x-axis; PM TRPV4_peg/PM TRPV4_ctrl) and greater than 1-fold increase in diffusivity (y-axis; Dpeg/Dctrl) by PEG-300. (S) Cell volume reduction-mediated mechanotransduction pathway can be clearly observed from plotting plasma membrane association of TRPV4 (x-axis) and relative increase in diffusivity with GSK101 versus GSK219 (y-axis), which are significantly greater than 2 for MCF10CA1a and MCF10DCIS.com cells.

MCF10CA1a cells showed plasma membrane relocation of TRPV4 in response to TRPV4 inhibition (1 nM GSK219; ND 219H), PEG 300 (74.4 mOsm/kg; ND PEG300), and OC conditions. In contrast, treatment with 0.2 pM GSK101 did not affect TRPV4 plasma membrane association (Supplementary Fig. 16 and Fig. 6A, showing IF images for TRPV4 (red) and DAPI (cyan)). This response was similar to that observed in MCF10DCIS.com cells under OC conditions. The plasma membrane associations of TRPV4 were quantified by line analysis, showing an increase under mechanical stresses (PEG and OC) and TRPV4 inhibition with GSK219, as plotted in Fig. 6B. As expected, based on the differences in plasma membrane TRPV4 association, MCF10CA1a cells exhibited increased motility (D) in response to TRPV4 inhibition (GSK219) and hyperosmotic conditions (PEG). Additionally, their movement became more directional (v) with TRPV4 activation (GSK101), similar to the observations in MCF10DCIS.com cells (Fig. 5I, 5J).

In contrast, the negative control MCF10AT1 cells did not exhibit TRPV4 inhibition-dependent plasma membrane relocation of TRPV4, as shown by IF images (Suppl. Fig. 14, Fig. 6D) and line analysis results (Fig. 6E). In these cells, TRPV4 inhibition (GSK219) decreased diffusivity (D), while PEG 300 had no effect. Moreover, movement directionality (v) remained unchanged under all treatment conditions (Fig. 6F).

The three test group cells, MDA-MB-231 (Fig. 6G, 6H), ETCC-06 (Fig. 6J, 6K), and ETCC-10 (Fig. 6M, 6N), failed to relocate TRPV4 to the plasma membrane in response to mechanical stresses (PEG and OC), as shown by IF images and quantification. This suggests that these cells lack the pro-invasive mechanosensitive cell volume reduction capability observed in MCF10DCIS.com cells. Consistently, none of these cells showed increased motility (D) under hyperosmotic conditions or TRPV4 inhibition (Fig. 6I, 6L, 6O).

We plotted these results to show cell diffusivity (D) scaling with varying plasma membrane associations of TRPV4, where MCF10DCIS.com (green hexagon) and MCF10CA1a (red hexagon) exhibited a positive scaling (Fig. 6P). However, the changes in plasma membrane TRPV4 occurred over a narrower range (31 – 79%) for MCF10CA1a compared to MCF10DCIS.com cells (16 -75%), reflecting the larger mechanosensitive cell volume plasticity of MCF10DCIS.com cells. In contrast, MDA-MB-231, ETCC-6, and ETCC-10 cells did not show this scaling (Fig. 6Q). Like MCF10AT1 cells, plasma membrane TRPV4 associations in these cells remained at similar percentages (Fig. 6Q).

This dichotomy in the presence of a pro-invasive mechanotransduction program is evident in the plots of diffusivity (D) increase versus plasma membrane TRPV4 relocation (Fig. 6R), where only MCF10CA1a and MCF10DCIS.com cells show a significant increase in both parameters. Furthermore, the relationship between plasma membrane-associated TRPV4 and D increase with GSK219 versus GSK101 can be used to assess the presence of a pro-invasive, mechanosensitive cell volume reduction program. Most non-mechanotransducing cells remained close to the baseline value of 1, while mechanotransducing cells showed a significant increase in D with GSK219 treatment.

Discussion

Our study explored how cell crowding, a common condition in diseases and development, impacts cell invasiveness. We employed novel assays to quantify cell invasion and single-cell motility in cells with breast tissue pathologies, including DCIS and ADH, which experience cell crowding due to abnormal proliferation in confined ducts. Our results revealed a novel mechanotransduction pathway activated by crowding in high-grade DCIS cells, establishing a direct link between crowding and invasiveness. This pathway highlights the intricate interplay between mechanical stimuli, ion channel activation, trafficking changes, cell volume reduction, stiffening, and invasiveness (Fig. 7). It distinguishes high-grade DCIS cells from normal, hyperplastic, or lower-grade DCIS cells, supporting previous findings that minimal cell volume reduction enables efficient invasion38.

Cell crowding triggers the activation of a pro-invasive mechanotransduction pathway in high-grade DCIS cells but not in less aggressive or normal cells.

This pro-invasive mechanotransduction pathway involves cell volume reduction and cortical stiffening driven by ion channel inhibition. Ion channel inhibition leads to the relocation of ion channels to the plasma membrane (PM). Under ND conditions, ion channels are mainly cytoplasmic in both high-grade DCIS and less aggressive/normal cells. However, high-grade DCIS cells have a larger cell volume under ND conditions. Cell crowding induces ion channel inhibition, leading to decreased intracellular calcium and subsequent cell volume reduction, which correlates with increased invasiveness. In contrast, less aggressive and normal cells do not exhibit this response. The inset image shows that mechanosensitive plasma membrane relocation of ion channels (orange) scales with increased invasiveness due to the activation of the cell volume reduction mechanotransduction pathway, while the absence of relocation indicates no mechanosensitive invasive increase.

The correlation between cell volume reduction and increased invasiveness highlights the specialized mechanotransduction abilities of high-grade DCIS cells. Using mass spectrometry, we identified TRPV4 as a crucial component of this pathway, providing insights into how non-invasive breast cancer cells may evolve into invasive phenotypes. By employing Fluo-4 assays and TRPV4 activators and inhibitors, we found that cell crowding triggered TRPV4 inhibition, leading to reduced intracellular calcium and subsequent cell volume reduction in MCF10DCIS.com cells. Notably, TRPV4 inhibition prompted its relocation to the plasma membrane, priming it for activation under suitable signaling conditions, thereby maintaining cellular homeostasis. TRPV4 inhibition by cell crowding increased invasiveness and motility, while TRPV4 activation had the opposite effect. Reducing TRPV4 protein expression by 50% via shRNA diminished the calcium decrease and volume reduction under mechanical stress, such as cell crowding and hyperosmotic conditions. Consequently, these cells did not exhibit increased motility under hyperosmotic conditions, demonstrating TRPV4’s critical role in pro-invasive cell mechanotransduction via cell volume reduction. Invasive MCF10CA1a cells also demonstrated the modest mechanotransduction capability. In contrast, other invasive breast cancer cell lines, such as MDA-MB-231 cells, exhibited opposing responses to TRPV4 inhibition and activation, highlighting the heterogeneity of TRPV4 roles in invasive cancer progression81–83.

Our discovery of the selective mechanotransduction pathway in high-grade DCIS cells was further validated by analyzing patient-derived breast cancer tissues. We found that TRPV4 was predominantly associated with the plasma membrane in high-grade DCIS lesions, but not in lower-grade DCIS or less aggressive pathologies. This selective relocation reinforces TRPV4’s crucial role in the pro-invasive mechanotransduction pathway unique to high-grade DCIS cells. The association of TRPV4 with the plasma membrane may serve as a valuable biomarker for identifying high-grade DCIS lesions.

Our data indicate that ion channels beyond TRPV4 may contribute to pro-invasive mechanotransduction. Notably, SCN11A and KCNN4 showed increased plasma membrane association under cell crowding conditions, suggesting their roles in modulating ion flux and cell volume and thus contributing to the mechanotransduction response. Although Piezo1 is known to respond to mechanical stimuli62, its lesser plasma membrane relocation compared to TRPV4 under cell crowding and hyperosmotic conditions suggests a lesser role in promoting invasiveness in this case. Further comparisons are needed to determine the dominant ion channel driving mechanotransduction in high-grade DCIS cells. However, it is evident that cell crowding-induced pro-invasive cell volume reduction pathway involves the coordinated action of multiple ion channels27,84.

We made the intriguing discovery that hyperosmotic conditions, which reduced’ cell volume through osmotic water outflow, mimic the effects of cell crowding by inducing similar pro-invasive cell volume reduction. Treatment with hyperosmotic agents like PEG 300 led to TRPV4 relocation to the plasma membrane and increased cell invasiveness, highlighting TRPV4’s mechanosensing role in response to external mechanical stimuli. This suggests that the convergence of mechanical stresses, including cell crowding and hyperosmotic stress, is a crucial trigger for pro-invasive cell volume reduction pathways, leading to a unified mechanotransduction signaling cascade.

The study’s insights have far-reaching implications, unlocking new research paths in cancer and biological sciences. By elucidating mechanosensitive responses to cell crowding, we can shed light on tissue development, wound healing, and physiological processes involving cell density changes. The parallels between cell crowding and hyperosmotic conditions in driving pro-invasive behaviors warrant further investigation into the intricate mechanisms of cytoskeleton reorganization, ion channel and transporter relocation, and enhanced cell invasiveness and motility. Importantly, our findings elucidate the mechanotransduction capability of high-grade DCIS cells that differentiate them from less aggressive cells. This understanding paves the way for developing diagnostic and prognostic strategies that leverage the selective intracellular localization patterns of TRPV4 and other mechanosensitive ion channels, ultimately guiding clinical decisions for patients with high-risk DCIS and other cancers.

Materials and methods

Cell culture and treatments

MCF10A cells were purchased from ATCC, and MCF10AT1, MCF10DCIS.com, and MCF10CA1a cells were procured from the Barbara Ann Karmanos Cancer Institute after establishing material transfer. These cells were authenticated prior to purchase, and cells with a low passage number (< 15) were used for experiments. We confirmed that the cells were mycoplasma-free using PCR analysis. MCF10A and MCF10AT1 cells were maintained in DMEM/F12 supplemented with 5% horse serum, 20 ng/mL EGF, 0.5 μg/mL hydrocortisone, 10 μg/mL insulin, and 1% penicillin and streptomycin. MCF10A cells were also supplemented with 100 ng/mL cholera toxin. MCF10DCIS.com and MCF10CA1a cells were cultured in DMEM/F12 supplemented with 5% horse serum and 1% penicillin and streptomycin. cells were cultured in 5% CO2 at 37 °C. For TRPV4 activator and inhibitor, we used GSK 1016790A (GSK101; Tocris 64T33) and GSK 2193874 (GSK219; Tocris 5106).

For volume measurements, parental cells were mixed with those expressing RFP at a 9:1 ratio to distinguish individual cells. Cells were treated with 2% (74.4 mOsmol/kg) or 4% PEG-300 (148.8 mOsmol/kg; v/v; Millipore Sigma 8074845000) in culture medium for 15 min, 2% PEG-300 for 48 h, or GSK1016790A (Tocris; 0.05 and 0.2 pM) or GSK2193874 (Tocris; 0.2 and 1 nM) for 48 h.

TRPV4 knockdown by shRNA

Cells in growth medium without penicillin and streptomycin were transfected with the hTRPV4 shRNA plasmid (1 and 2 µg; Santa Cruz sc-61726-SH) and Lipofectamine3000 in opti-MEM. At 6 h after transfection, the medium was replaced, and cells were cultured for 30 h before processing for western blot and immunofluorescence analyses.

Immunofluorescence

The cells were fixed with 4% paraformaldehyde (Fisher Scientific) for 20 min, followed by permeabilization with 0.1% saponin (Fisher Scientific) for 5 min at room temperature (RT). After permeabilization, the cells were washed three times in PBS and blocked in 0.1% saponin + 10% BSA in PBS at RT for 1 h. Cells were incubated with primary antibodies overnight at 4 °C in 0.1% saponin in PBS. The primary antibodies used were TRPV4 (Abcam 39260; 1:500 dilution), TfR (ThermoFisher Scientific 13-6800; 1:500 dilution), Piezo1 (Alomone APC-087; 1:500), and KCNN4 (Alomone ALM-051; 1:500). Cells were washed three times in PBS, incubated with fluorescent-tagged anti-rabbit-Alexa 550 or anti-mouse-Alexa 488 secondary antibodies (Thermo Fisher Scientific) and DAPI for 1 h at RT, and imaged by confocal microscopy. For plasma membrane staining, DiIC18(3) (’1,1’-Dioctadecyl-3,3,3’,3’-Tetramethylindocarbocyanine Perchlorate; DiIC18(3); Thermo Fisher Scientific, Cat # D282) was diluted in warm PBS to 5 uM.

Media was removed from the cells, then the 5 uM DiIC18(3) in PBS was added. The dish was transferred to the incubator (37 C, 5% CO2) for 10 minutes. Following the 10 min incubation, the DiIC18(3) in PBS was removed and the cells were washed 2x with complete cell growth media. The cells were then fixed with 2% PFA in PBS

Confocal imaging and image processing

2D images were acquired using a Yokogawa spinning-disk confocal microscope (Andor Technology) installed in a Nikon Eclipse TE2000 inverted microscope using a 60x/1.49NA Plan Apo objective (Nikon). The samples were illuminated using 430, 488, 561, and 647 nm solid-state lasers (Andor Technology). Images were acquired using an iXon back-illuminated EMCCD camera (Andor Technology). For volume measurements, 3D confocal images were acquired using z-step sizes calculated based on Nyquist conditions. Images were processed using the ImageJ (NIH) or Imaris (Bitplane) software. For live cell imaging, cells were maintained in 5% CO2 at 37 °C in a stage-top incubator (Oko-lab).

Western blotting

ND cells (40–60% cell density) were lysed in SDS sample buffer, and OC cells were lysed in cytoskeleton buffer (10 mM Tris pH 7.4,100 mM NaCl, 1 mM EDTA, 1 mM EGTA, 1% Triton X-100, 10% glycerol, 0.1% SDS, 0.5% deoxycholate). SDS sample buffer (Bio-Rad) supplemented with reducing reagent was added to lysates and boiled at 100°C for 10 min. The samples were separated on 4–15% SDS-PAGE gels and transferred to nitrocellulose membranes (Bio-Rad). Membranes were blocked in TBST (TBS + 0.1% Tween-20) containing 5% non-fat milk and then incubated with primary TRPV4 antibody (ThermoFisher Scientific PA541066; 1:1000) and GAPDH (Santa Cruz sc-32233; 1:2000) overnight at 4 °C. Membranes were washed three times in TBST (5 min each) and then incubated with horseradish peroxidase (HRP)-conjugated anti-rabbit (ThermoFisher Scientific A28177) or anti-mouse (ThermoFisher Scientific A28177) secondary antibodies for 30 min at RT. Blots were visualized using West Femto maximum sensitivity chemiluminescent substrate (Thermo Fisher Scientific).

Lentiviral infection and stable cell generation

To produce lentiviruses encoding RFP, HEK-293 cells were transfected with the lentiviral vector plasmid DNA using Lenti-X Packaging Single Shots (Takara). pCSII-EF-miRFP670v1-hGem(1/110) was a gift from Vladislav Verkhusha (Addgene plasmid # 80006; http://n2t.net/addgene:80006; RRID:Addgene_80006). Supernatants containing lentiviral pseudoparticles were harvested 24 and 48 h post-transfection. harvested lentiviral particles were immediately stored at −80 °C. To establish stable cell lines, 70% confluent cells at one day post-seeding were infected with lentivirus in the presence of 8 µg/mL polybrene (Thermo Fisher Scientific). Two days after transduction, 1 µg/mL puromycin (Thermo Fisher Scientific) was added to the medium to select for stably transduced cells. The samples were visualized daily to ensure that the untransduced cells in the wells were not viable. Once the polyclonal populations had sufficiently expanded, cell stocks were prepared and harvested for protein expression assays.

Cell invasion assay

For cell invasion assays, 8-well chamber slides (Nunc Lab-Tek) were coated with 50 µg/mL poly-L-lysine (Thermo Fisher Scientific) for 30 min and washed with PBS. The slides were fixed with 0.5% glutaraldehyde (Thermo Fisher Scientific) for 20 min and washed. Gelatin was conjugated with fluorescein by mixing 200 µL fluorescein-gelatin (1 mg/mL; ThemoFisher Scientific) with 800 μL unlabeled gelatin (4 mg/mL), followed by incubation for 5 min at 60 °C. The gelatine mix was allowed to cool at RT for 5 min and then applied, and the slides were incubated for 15 min. Slides were washed with PBS and disinfected with 70% ethanol for 30 min. Following three PBS washes, the residual reactive groups were quenched with growth medium by incubating at room temperature for 30 min in the dark. Cells were seeded in a fresh growth medium and incubated undisturbed on a horizontal surface at RT for 30 min to encourage cell distribution. Chamber slides were incubated with 5% CO2 at 37 °C. To detect cell invasion, cells were fixed with 4% formaldehyde and stained with DAPI (Thermo Fisher Scientific), and imaging was performed using a x4/0.2 NA Plan Apo objective (Nikon). Cell invasion was determined by quantifying the sites of the degraded matrix, which were visible as dark areas in the bright-green fluorescent gelatin matrix. The area of gelatin digestion and the number of cells were calculated using the ImageJ software.

Nanoindenter assay for measuring cortical stiffness

Stiffness measurements were performed on live cells in 5% CO2 at 37 °C, maintained in a stage-top chamber (Oko-lab) using a nanoindenter (Optics11 Chiaro) attached to the confocal microscope. Nanoindentation was performed at the cell surface of single cells with an indentation probe with a spring constant (0.24 N/m) and tip diameter (10 μm). The Hertzian contact model was used to fit the data to extract Young’s modulus in the elastic regime.

Surface biotinylation and pull-down assay

ND cells that were 40%–70% confluent and OC cells were detached and washed twice with PBS. The cells were resuspended at a concentration of 25 × 106 cells/mL in PBS containing 2 mM sulfo-NHS-biotin (Invitrogen). The reaction mixture was then incubated at RT for 60 min. Cells were washed three times with 1 M Tris pH 8.0 to quench and remove the excess biotin reagent. Cell pellets were lysed in 1.0 mL of cold lysis buffer (50 mM HEPES pH 7.2, 150 mM NaCl, 1.0% Triton X-100, 1.0% CHAPS, 100x protease and phosphatase inhibitors). Lysates were incubated on ice for 30 min, transferred to tubes, and centrifuged (10,000 × g for 5 min) to remove insoluble material. Supernatants were collected, and protein concentrations were measured using the bicinchoninic acid assay (Pierce). We washed 50 µL resin (ThermoFisher Scientific) with immobilized streptavidin twice with binding buffer (0.1 M phosphate, 0.15 M NaCl pH 7.2) and centrifuged at 5000 x g for 1 min. Biotin-labelled cell lysate (1 mg) was added to the resin and incubated with rotation for 1 h at RT. The resin was washed by resuspending in binding buffer, centrifuging to pellet the resin, and removing the supernatant by aspiration. The wash was repeated four times. Samples were boiled in SDS-PAGE sample buffer with DTT and separated by electrophoresis.

Mass spectrometry

Peptide mixtures from each sample were analyzed by LC-MS/MS using a nano-LC system (Easy nLC1000) connected to a Q Exactive HF mass spectrometer (Thermo Fisher Scientific). The platform was configured with a nano-electrospray ion source (Easy-Spray, ThermoFisher Scientific), an Acclaim PepMap 100 C18 nanoViper trap column (3 μm particle size, 75 μm ID × 20 mm length), and an EASY-Spray C18 analytical column (2 μm particle size, 75 μm ID × 500 mm length). Peptides were eluted at a flow rate of 300 nL/min using linear gradients of 5–25% acetonitrile (in aqueous phase and 0.1% formic acid) for 40 min, followed by 45% for 10 min, and static flow at 90% for 10 min. Mass spectrometry data were collected in a data-dependent manner, switching between one full-scan MS mode (m/z:380–1400; resolution: 60 K; AGC:3e6; maximum ion time:20 ms), and 10 MS/MS scans (resolution: 15 K; AGC:1e5; maximum ion time:120 ms; nCE:27) of the top 10 target ions. Ions were sequenced once and dynamically excluded from the list for 20 s. The datasets were searched using MaxQuant at default parameters against the UniProt Human Proteome database.

Line analysis

IF images of ion channels (TRPV4, PIEZO1, or KCNN4), DiIC18(3), and DAPI were opened in ImageJ. These images were concatenated so that a line crossing a cell in all three images could be used to measure approximate percentage protein localization in the plasma membrane, cytosol, or nucleus. Background was subtracted from all images. DiIC18(3) signal was used as a guide for plasma membrane location (between 50% of the peak intensity) and DAPI signal (between 50% of the peak intensity) for nuclear locations. The cytosol location was defined as the region between the 50% peak intensity points of the DiIC18(3) and DAPI signals. The ion channel signal in each window along the line was then averaged. The percentage of protein localization in the plasma membrane, cytosol, or nucleus for a cell was calculated by dividing the average intensity of each location by the sum of intensities in all three locations.

Live cell detection of plasma membrane associated TRPV4

Cells were seeded in a LabTek II 8-well Chambered Coverglass dish (Thermo Fisher 155409) at a density of 10,000 cells per well and allowed to adhere and grow for 2 days. Afterward, the cells were incubated on ice for 1 hour with 1:300 Anti-TRPV4 (extracellular) rabbit primary antibody (Alomone ACC-124) in either complete media or 74 mOsm/Kg PEG 300 in complete media. Following the incubation, the cells were washed twice with the respective media (complete media or 74 mOsm/Kg PEG 300 in complete media) while remaining on ice. Subsequently, the cells were stained for 1 hour at 22 °C with 1:5,000 Donkey anti-Rabbit IgG H+L Highly Cross-Adsorbed Secondary Antibody, Alexa Fluor™ 647 (Thermo Fisher A31573) in the same respective media. After the staining period, the cells were imaged.

Pipeline for analyses of patient histologic samples

Specimen collection: A retrospective clinical study was designed to collect patient tissue blocks. A pathologist chose 39 patient breast formalin-fixed paraffin-embedded (FFPE) tissue specimens of a combination of the following pathologies (benign, ADH, IDC, low-, intermediate-, or high-grade DCIS, and IDC), along with normal tissue for each patient. Specimens associated with a past cancer diagnosis and drug treatment history were excluded.

Sample preparation for histopathologic evaluation: For each FFPE block, two serial dissections were carried out. Tissue sections on unlabeled slides were stained with hematoxylin and eosin (H&E) and subjected to immunohistochemical (IHC) staining using TRPV4 antibodies (Abcam 39260; 1:100 dilution). Whole-slide images were captured using the Olympus VS200 whole-slide imaging system. To validate the binding specificity of the antibody, both negative and positive control samples were prepared.

Two sequential annotations: A breast surgical pathologist annotated the pathology stages within specified regions of interest (ROIs) on wholeslide H&E images, based on cell morphological features. Two pathologists independently evaluated the same ROIs in the corresponding IHC images. They annotated the protein distribution using a three-tier classification:

Case 1: Absence of TRPV4.

Case 2: Intracellular TRPV4 localization.

Case 3: Presence of TRPV4 in the plasma membrane, with or without intracellular TRPV4.

The independent annotations concerning protein localization by the two pathologists were then subjected to statistical tests for selectivity and specificity. Any IHC ROIs with classification disagreements between the pathologists were designated as equivocal cases.

Calcium reporter assay

This assay was completed using the Fluo-4 Direct™ Calcium Assay Kit (Thermo Fisher F10471), including the assay buffer (Thermo Fisher), 2X Fluo-4 calcium assay reagent (Thermo Fisher), and preweighed, water-soluble probenecid (Thermo Fisher P36400; 2.5 mM). The 2X Fluo-4 calcium assay reagent was first prepared by adding 10 mL of the Fluo-4 calcium assay buffer and 200 μL of 250 mM probenecid to the desiccated calcium assay reagent at room temperature (21 °C). This mixture was then vortexed and allowed to sit for 15 minutes. While waiting for the 2X Fluo-4 calcium assay reagent to finish being prepared, the media from the well of a LabTek II Chambered Coverglass with Cover #1.5 Borosilicate Sterile 8-well plate (Thermo Fisher 155409) was removed and replaced with 200 μL of complete media. Once the 2X Fluo-4 calcium assay reagent was fully prepared, 200 μL of the reagent was added to the well with 200 μL of complete media. The 8-well plate was then placed at 37 °C and 5% CO₂ for 30 minutes. Following 30 minutes at 37 °C and 5% CO₂, the 8-well plate was removed from the incubator and placed at room temperature for 30 minutes. After incubation in the reagent for 30 minutes at room temperature, the 8-well plate was imaged using the confocal microscope with a 488 nm laser for 30 minutes with 30-second intervals between images and an exposure time of 200 ms. If the sample was treated, the sample was imaged for 35 minutes with 30-second intervals, with the treatment being added after 5 minutes had elapsed. Added treatments all had a final volume of 200 μL and were composed of: 1 nM GSK 2 in 1:1 media:2X Fluo-4 calcium assay reagent, 0.2 nM GSK 2 in 1:1 media:2X Fluo-4 calcium assay reagent, 0.2 pM GSK 1 in 1:1 media:2X Fluo-4 calcium assay reagent, 1:1:1 water:media:2X Fluo-4 calcium assay reagent, and 0.3 Osm/L PEG 300 in 1:1 media:2X Fluo-4 calcium assay reagent.

Cell motility assay

Cells were first seeded in MatTek 35 mm dishes with No. 1.5 coverslip, 14 mm glass diameter, and Poly-D-Lysine coated glass bottoms (MatTek P35GC-1.5-14-C). Once the cells adhered, they were stained with 10 μg/mL WGA-488 (Thermo Fisher W11261) for 10 minutes at room temperature. Following the staining, the cells were washed with complete media twice and then incubated at 37 °C and 5% CO₂ for 1 hour. During this one-hour incubation, the cells received the following treatments: 0.2 pM GSK 101 in complete media for 1 hour, 0.05 pM GSK 101 in complete media for 1 hour, 1 nM GSK 219 in complete media for 1 hour, 0.2 nM GSK 219 in complete media for 1 hour, 74.4 mOsm/Kg PEG 300 in complete media for 15 minutes, and 1:1 hypoosmotic solution in complete media for 15 minutes. After the incubation, the sample was placed in an OkoLab chamber at 37 °C and 5% CO₂ for imaging. Imaging was completed using a confocal microscope with a 488 nm laser for 3 hours with 1-minute intervals.

Cell viability assay

MCF10DCIS.com cells were plated in a LabTek II 8-well Chambered Coverglass dish (Thermo Fisher 155409) at a seeding density of 10,000 cells per well. Once the cells in the well reached the desired number of days post-confluence, they were stained using Propidium Iodide Ready Flow Reagent (Thermo Fisher R37169). Specifically, 1 drop of propidium iodide (PI) was added to 500 µL of cell growth medium per well and incubated for 15 minutes at room temperature (22°C). After staining, the cells were washed twice with complete growth medium. Next, the cells were stained with 1 µg/mL WGA-488 (Thermo Fisher w11261) in complete growth medium for 15 minutes at 37°C in a 5% CO₂ atmosphere. Following the incubation with WGA-488, the cells were washed three times with complete growth medium and then incubated for 2 hours at 37°C in 5% CO₂ to allow WGA migration into the nuclei. Finally, the cells were imaged using confocal microscopy at 4X magnification, with laser excitation at 488 nm for WGA-488 and 560 nm for propidium iodide.

eLife Assessment

This fundamental study demonstrates the role of TRPV4 channels in mechano-transduction during carcinoma progression, showing how high cell confluence in DCIS cells activates pathways regulating calcium homeostasis, cell volume reduction, and invasiveness. The authors addressed all prior concerns, with the shRNA knockdown providing compelling evidence for TRPV4's role in metastasis.

Reviewer #1 (Public review):

Summary:

In this study, Bu et al examined the dynamics of TRPV4 channel in cell overcrowding in carcinoma conditions. They investigated how cell crowding (or high cell confluence) triggers a mechano-transduction pathway involving TRPV4 channels in high-grade ductal carcinoma in situ (DCIS) cells that leads to large cell volume reduction (or cell volume plasticity) and pro-invasive phenotype.

In vitro, this pathway is highly selective for highly malignant invasive cell lines derived from a normal breast epithelial cell line (MCF10CA) compared to the parent cell line, but not present in another triple-negative invasive breast epithelial cell line (MDA-MB-231). The authors convincingly showed that enhanced TRPV4 plasmamembrane localization correlates with high-grade DCIS cells in patient tissue samples. Specifically in invasive MCF10DCIS.com cells they showed that overcrowding or over-confluence leads to a decrease in cell volume and intracellular calcium levels. This condition also triggers the trafficking of TRPV4 channels from intracellular stores (nucleus and potentially endosomes), to the plasma membrane (PM). When these over-confluent cells are incubated with a TRPV4 activator, there is an acute and substantial influx of calcium, attesting the fact that there are high number of TRPV4 channels present on the PM. Long-term incubation of these over-confluent cells with the TRPV4 activator results in the internalization of the PM-localized TRPV4 channels.

In contrast, cells plated at lower confluence primarily have TRPV4 channels localized in the nucleus and cytosol. Long-term incubation of these cells at lower confluence with a TRPV4 inhibitor leads to the relocation of TRPV4 channels to the plasma membrane from intracellular stores and a subsequent reduction in cell volume. Similarly, incubation of these cells at low confluence with PEG 3000 (a hyperosmotic agent) promotes the trafficking of TRPV4 channels from intracellular stores to the plasma membrane.

Strengths:

The study is elegantly designed and the findings are novel. Their findings on this mechano-transduction pathway involving TRPV4 channels, calcium homeostasis, cell volume plasticity, motility and invasiveness will have a great impact in the cancer field and potentially applicable to other fields as well. Experiments are well-planned and executed, and the data is convincing. Authors investigated TRVP4 dynamics using multiple different strategies- overcrowding, hyperosmotic stress, pharmacological and genetic means, and showed a good correlation between different phenomena.

All of my previous concerns have been addressed. The quality of the manuscript has improved significantly.

Reviewer #2 (Public review):

Summary:

The metastasis poses a significant challenge in cancer treatment. During the transition from non-invasive cells to invasive metastasis cells, cancer cells usually experience mechanical stress due to a crowded cellular environment. The molecular mechanisms underlying mechanical signaling during this transition remain largely elusive. In this work, the authors utilize an in vitro cell culture system and advanced imaging techniques to investigate how non-invasive and invasive cells respond to cell crowding, respectively.

The results clearly show that pre-malignant cells exhibit a more pronounced reduction in cell volume and are more prone to spreading compared to non-invasive cells. Furthermore, the study identifies that TRPV4, a calcium channel, relocates to the plasma membrane both in vitro and in vivo (patient's samples). Activation and inhibition of TRPV4 channel can modulate the cell volume and cell mobility. These results unveil a novel mechanism of mechanical sensing in cancer cells, potentially offering new avenues for therapeutic intervention targeting cancer metastasis by modulating TRPV4 activity. This is a very comprehensive study, and the data presented in the paper are clear and convincing. The study represents a very important advance in our understanding of the mechanical biology of cancer.

Author response:

The following is the authors’ response to the previous reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

In this study, Bu et al examined the dynamics of TRPV4 channel in cell overcrowding in carcinoma conditions. They investigated how cell crowding (or high cell confluence) triggers a mechano-transduction pathway involving TRPV4 channels in high-grade ductal carcinoma in situ (DCIS) cells that leads to large cell volume reduction (or cell volume plasticity) and proinvasive phenotype.

In vitro, this pathway is highly selective for highly malignant invasive cell lines derived from a normal breast epithelial cell line (MCF10CA) compared to the parent cell line, but not present in another triple-negative invasive breast epithelial cell line (MDA-MB-231). The authors convincingly showed that enhanced TRPV4 plasmamembrane localization correlates with highgrade DCIS cells in patient tissue samples. Specifically in invasive MCF10DCIS.com cells they showed that overcrowding or over-confluence leads to a decrease in cell volume and intracellular calcium levels. This condition also triggers the trafficking of TRPV4 channels from intracellular stores (nucleus and potentially endosomes), to the plasma membrane (PM). When these over-confluent cells are incubated with a TRPV4 activator, there is an acute and substantial influx of calcium, attesting the fact that there are high number of TRPV4 channels present on the PM. Long-term incubation of these over-confluent cells with the TRPV4 activator results in the internalization of the PM-localized TRPV4 channels.

In contrast, cells plated at lower confluence primarily have TRPV4 channels localized in the nucleus and cytosol. Long-term incubation of these cells at lower confluence with a TRPV4 inhibitor leads to the relocation of TRPV4 channels to the plasma membrane from intracellular stores and a subsequent reduction in cell volume. Similarly, incubation of these cells at low confluence with PEG 3000 (a hyperosmotic agent) promotes the trafficking of TRPV4 channels from intracellular stores to the plasma membrane.

Strengths:

The study is elegantly designed and the findings are novel. Their findings on this mechanotransduction pathway involving TRPV4 channels, calcium homeostasis, cell volume plasticity, motility and invasiveness will have a great impact in the cancer field and potentially applicable to other fields as well. Experiments are well-planned and executed, and the data is convincing. Authors investigated TRVP4 dynamics using multiple different strategies- overcrowding, hyperosmotic stress, pharmacological and genetic means, and showed a good correlation between different phenomena.

All of my previous concerns have been addressed. The quality of the manuscript has improved significantly.

We are deeply grateful to the reviewer for their thoughtful assessment and invaluable suggestions, including crucial additional experiments and more effective presentation and description of our findings, which have greatly enhanced the quality of our manuscript.

Reviewer #2 (Public review):

Summary:

The metastasis poses a significant challenge in cancer treatment. During the transition from non-invasive cells to invasive metastasis cells, cancer cells usually experience mechanical stress due to a crowded cellular environment. The molecular mechanisms underlying mechanical signaling during this transition remain largely elusive. In this work, the authors utilize an in vitro cell culture system and advanced imaging techniques to investigate how non-invasive and invasive cells respond to cell crowding, respectively.

The results clearly show that pre-malignant cells exhibit a more pronounced reduction in cell volume and are more prone to spreading compared to non-invasive cells. Furthermore, the study identifies that TRPV4, a calcium channel, relocates to the plasma membrane both in vitro and in vivo (patient's samples). Activation and inhibition of TRPV4 channel can modulate the cell volume and cell mobility. These results unveil a novel mechanism of mechanical sensing in cancer cells, potentially offering new avenues for therapeutic intervention targeting cancer metastasis by modulating TRPV4 activity. This is a very comprehensive study, and the data presented in the paper are clear and convincing. The study represents a very important advance in our understanding of the mechanical biology of cancer.

We sincerely appreciate the reviewer’s insightful evaluation and invaluable recommendations for key additional experiments, which have significantly strengthened our manuscript.

Version 3

Abstract

Abstract

Cell crowding is a common microenvironmental factor influencing various disease processes, but its role in promoting cell invasiveness remains unclear. This study investigates the biomechanical changes induced by cell crowding, focusing on pro-invasive cell volume reduction in ductal carcinoma in situ (DCIS). Crowding specifically enhanced invasiveness in high-grade DCIS cells through significant volume reduction compared to hyperplasia-mimicking or normal cells. Mass spectrometry revealed that crowding selectively relocated ion channels, including TRPV4, to the plasma membrane in high-grade DCIS cells. TRPV4 inhibition triggered by crowding decreased intracellular calcium levels, reduced cell volume, and increased invasion and motility. During this process, TRPV4 membrane relocation primed the channel for later activation, compensating for calcium loss. Analyses of patient-derived breast cancer tissues confirmed that plasma membrane- associated TRPV4 is specific to high-grade DCIS and indicates the presence of a pro-invasive cell volume reduction mechanotransduction pathway. Hyperosmotic conditions and pharmacologic TRPV4 inhibition mimicked crowding-induced effects, while TRPV4 activation reversed them. Silencing TRPV4 diminished mechanotransduction in high-grade DCIS cells, reducing calcium depletion, volume reduction, and motility. This study uncovers a novel pro-invasive mechanotransduction pathway driven by cell crowding and identifies TRPV4 as a potential biomarker for predicting invasion risk in DCIS patients.

Introduction

The complex interplay between cellular mechanics and invasive behaviors is crucial to understanding various physiological and pathological processes, such as wound healing1 and disease progression2,3. Numerous studies show that these processes are mediated by mechanotransduction, whereby cells translate mechanical stimuli into cellular activity2,4–6. While mechanotransduction is traditionally studied concerning fluid stress7,8, matrix stiffness4,9, and other biomechanical changes such as osmotic stress10,11, the role of cell crowding, characterized by increased cell density and spatial constraints, is relatively less explored. A recent report shows that cell crowding plays a role in facilitating wound closure and repair by enhancing cell proliferation12. As tissues develop, repair, or undergo pathological transformation, cell crowding becomes common13,14. This challenges individual cells, forcing them to perceive and respond to the mechanical constraints of a crowded environment. Our study used human breast cell line model systems to describe a novel mechanotransduction pathway triggered by cell crowding that induces invasiveness into surrounding tissues. Interestingly, this pathway exhibited unique selectivity, as it specifically associated with a type of non-invasive cancer pathology and was not present in lower-grade or less aggressive pathologies. This suggests that not all cells possess the ability to translate mechanical stimuli, such as cell crowding, into cell invasiveness.

To assess the role of cell crowding in cell invasiveness, we chose in vitro cell lines linked to different pathological states that reflect cell crowding conditions in vivo, including atypical ductal hyperplasia (ADH)15 and ductal carcinoma in situ (DCIS)16–18. ADH is an intraductal clonal epithelial cell proliferative lesion18 and represents an intermediate step between normal breast tissue and in situ carcinomas19–21. ADH is associated with high risk as it is reported that an ADH diagnosis is associated with a five-fold increased risk of breast cancer15. DCIS is non-invasive form of cancer characterized by proliferating malignant epithelial cells16–18. Unlike ADH, DCIS is considered a precursor to invasive breast cancer22. However, the mechanism of how DCIS transitions to invasive cancer is not well understood and therefore, there is currently no reliable and robust method to differentiate which DCIS cells are at high risk of becoming invasive23,24.

Both ADH and DCIS conditions can potentially expose cells to crowding in vivo. However, how ADH and DCIS cells respond to such changes in cell density remains unknown. Our study revealed that cell crowding selectively triggered a pro-invasive mechanotransduction program in a specific type of DCIS cell line associated with high-grade pathology25,26. The mechanotransduction program induced by cell crowding involved the inhibition of ion channels, such as transient receptor potential vanilloid 4 (TRPV4)27, a calcium-permeable ion channel, as identified from our mass spectrometry assay. This inhibition prompted the relocation of TRPV4 and other ion channels to the plasma membrane, leading to decreased intracellular calcium levels, subsequent induction of reduced cell volume and cortical stiffening, thereby promoting cell invasiveness and motility.

Results

Cell crowding enhances invasiveness in high-grade DCIS cells

Cell crowding reflects a common condition in tumor microenvironments for ADH and DCIS, arising from aberrant cell proliferation within spatially constrained intraductal spaces. We examined the influence of this prevalent environmental factor on the invasiveness of these cells in vitro. To conduct this investigation, we assembled a panel of cell lines derived from the normal breast epithelial cell line MCF10A, including its H-RAS mutation-driven derivatives associated with various pathologies28: MCF10AT1, MCF10DCIS.com, and invasive cells MCF10CA1a. MCF10AT1 resembles ADH15, MCF10DCIS.com mimics high-grade DCIS26, and MCF10CA1a represents a malignant invasive cancer that was observed to form metastatic lesions in a mouse xenograft29. The current classification of DCIS relies on histological factors such as cell growth patterns and cytonuclear features30,31. Comedo-DCIS is a histologic subtype, which is characterized by apoptotic cell death, representing a high-grade DCIS with higher invasive potential than those of lower-grades25. MCF10DCIS.com forms comedo-type DCIS lesions that can spontaneously transition to invasive cancer when xenografted26.

To compare cell invasiveness under normal cell density and cell crowding conditions in vitro, we opted for a modified 2D matrix degradation assay. This approach allowed us to quantify overall cell invasiveness by normalizing it with the total cell number, thereby accounting for differences in cell densities. We chose this method over transwell, 3D Matrigel, or spheroid assays, where quantifying cell invasiveness as a function of cell density is challenging. By modifying an existing collagen-crosslinked polyacrylamide hydrogel matrix-based invasion assay32–34, we could determine the invasive cell fraction out of the total cell population by fluorescence imaging of the invasion gel bed and cell nuclei through low magnification (4X) imaging (Fig. 1A; for a detailed procedure, see Suppl. Fig. 1) when cells were at normal density versus under cell crowding conditions.

Cell crowding selectively increases invasiveness in high-grade DCIS cells.

(A) We used a collagen-crosslinked polyacrylamide hydrogel matrix-based invasion assay to assess the effect of cell crowding on cell invasiveness. Representative images show gelatin-Alexa488 conjugates, where dark areas in the green (Gelatin488) background indicate cell invasion through degradation, and DAPI staining marks cell locations (blue; DAPI) in a two-day invasion assay of MCF10DCIS.com cells under normal density (ND; upper panel) and overconfluent (OC; lower panel) conditions. “Masked” images are thresholded to produce positive masks applied to the “DAPI” images. Individual cell locations detected in “DAPI” images are marked with purple circles in “Detected points” images. The total number of cells within the field of view is counted from these points. By overlaying the mask and DAPI images, “Masked DAPI” images are obtained, and invaded cells are detected and represented by purple circles in “Masked points” images. The invasive cell fraction is calculated by the ratio of the number of invaded cells to the total number of cells (0.24 for ND and 0.59 for OC MCF10DCIS.com cells). These data show that cell invasiveness is enhanced by cell crowding. Scale bar = 100 μm. (B) Comparison of “Gelatin488” images of MCF10A (normal breast epithelial cells), MCF10AT1 (ADH-mimicking cells), MCF10DCIS.com (high-grade DCIS mimic), and MCF10CA1a (invasive breast cancer cells) between ND and OC conditions. MCF10DCIS.com cell invasion is significantly higher under cell crowding than under ND conditions. (C) Invasive cell fractions of these cells between ND (blue circles) and OC (red circles) conditions are compared, showing that cell crowding-induced increases in invasiveness are notable only in MCF10DCIS.com cells. The number of cell invasion analyses was as follows: MCF10A (ND: 10; OC: 7), MCF10AT1 (ND: 6; OC: 6), MCF10DCIS.com (ND: 9; OC: 9), and MCF10CA1a (ND: 6; OC: 7). We used the two-tailed Mann-Whitney U test, a nonparametric and unpaired statistical method, to compare differences between groups. ****: p < 0.0001, ***: p < 0.001, **: p < 0.01, *: p < 0.05, ns: p > 0.05, throughout the manuscript.

We observed alterations in live cell morphology and invasiveness as they progressed from normal density (ND: 40-70%) to full confluence (100%) and beyond. Our goal was to determine a distinct time window, termed overconfluence (OC), where cell crowding-induced changes level off. The OC time window was achieved in cells cultured for an additional 5-10 days after reaching 100% confluence, with the growth medium replaced twice per day to prevent acidification. We confirmed that the cells remained largely viable under crowding conditions, with minimal cell death (∼1.58%) similar to that observed in ND cells (∼0.85%) (Suppl. Fig. 2A-B). The schematic of the timeline to achieve the OC conditions, and low-magnification (4X) bright-field images at different time points from the ND to OC conditions are shown in Suppl. Fig. 3A-B. The time-dependent equilibration of cell invasiveness after the cells reached confluence is demonstrated in Suppl. Fig. 3C-D using our quantifiable collagen-crosslinked polyacrylamide hydrogel matrix-based invasion assay. Cell crowding mechanosensing significantly increased the invasive cell fraction of MCF10DCIS.com cells from ∼24% (ND cells) to 59% (OC cells). The enhanced invasiveness of MCF10DCIS.com cells under cell crowding was largely reversible. When OC cells were reseeded at normal density for invasion assays, their invasive cell fraction decreased to approximately 15%, slightly lower (p = 0.012) than the initial value of around 24% (Suppl. Fig. 3C, 3E). In contrast, normal breast epithelial cells (MCF10A) and the ADH-mimicking MCF10AT1 cells did not display invasiveness under ND conditions or enhanced invasiveness under OC conditions, indicating insensitivity to cell crowding (Fig. 1B-C).

The invasive cell fraction of ND MCF10CA1a cells was already ∼80%, and the additional increase in invasiveness due to cell crowding was not readily discernible, with a slight increase to ∼82% under crowded conditions (Fig. 1B-C). These data suggest a striking and selective mechanosensing effect of cell crowding on cell invasiveness in MCF10DCIS.com cells. To ensure that acidification did not affect the invasiveness of MCF10DCIS.com cells despite the frequent replacement of the cell growth medium, we incubated ND cells for two days with acidified medium from cultures of OC MCF10DCIS.com cells. We observed that medium acidity did not alter cell invasiveness (Suppl. Fig. 3F), reinforcing that the increased invasiveness under OC conditions was induced by cell crowding.

Cell crowding reduces cell volume and stiffens MCF10DCIS.com cells

As cells became crowded, we observed a reduction in cell size. Previous reports indicate an inverse relationship between cell volume and cortical stiffness35,36, leading us to hypothesize that reduced cell volume would be accompanied by increased cortical stiffness. Research on glioma cell invasion underscores the critical role of hydrodynamic cell volume changes in penetration into surroundings37,38, suggesting that significant cell volume reduction facilitates cell invasion. Additionally, increased cortical stiffness is known to help cells overcome the physical barriers of the dense extracellular matrix39,40. We thus investigated whether cell crowding selectively induces pro-invasive cell volume reduction and cell stiffening in MCF10DCIS.com cells, thereby priming them for invasion.

To measure the cell volume of individual cells, we used confocal microscopy with a 60X oil immersion objective to obtain z-stack images of live or fixed cells stably expressing red fluorescent protein (RFP)41. Using a nanoindenter attached to the confocal microscope (Suppl. Fig. 4A), we extracted Young’s modulus of live individual cells to assess changes in cortical stiffness. Cell crowding significantly reduced cell volume (Fig. 2A, B) and increased cortical stiffness (Fig. 2C) in both MCF10DCIS.com and MCF10A cells. However, the changes in volume and stiffness from ND to OC conditions were more pronounced in MCF10DCIS.com cells (Suppl. Fig. 4B, C). The cell crowding-induced alterations in cell volume and stiffness were negligible in MCF10AT1 cells (Fig. 2A-C). In MCF10CA1a cells, cell crowding led to an increase in stiffness, but a decrease in volume was not evident. This was expected because the cell volume was already too small to be accurately measured with the available resolution (Fig. 2A-C). The notable plasticity in cell volume and stiffness changes observed in MCF10DCIS.com cells in response to cell crowding potentially underscores the critical link between this plasticity and mechanosensitive increases in cell invasiveness. We assessed cell volume changes only as an effector event of cell crowding, without measuring cell stiffness, because cell volume reflects the mechanical properties of the entire cell, while Young’s modulus can vary depending on the location of indentations in the plasma membrane42,43. The observed cell volume changes by cell crowding (Fig. 2A) depended on the ND cell volume, as OC cell volumes under crowding conditions were comparable across all cells. Consequently, the ND cell volume and the volume changes from ND to OC conditions exhibited an approximately linear relationship (R² ∼ 0.97; Suppl. Fig. 4D). This finding suggests that the ND cell volume could serve as an indicator of relative cell volume plasticity.

Cell crowding induces significant cell volume reduction and stiffening in high-grade DCIS cells.

(A) Cell volume (V; mean and SD) differences between ND (blue circles) and OC (red circles) conditions of MCF10A, MCF10AT1, MCF10DCIS.com, and MCF10CA1a cells are plotted. The high-grade DCIS cell mimic, MCF10DCIS.com, shows a large volume reduction due to cell crowding. The number of single cell volume analyses (technical replicates merged from three independent experimental repeats) was as follows: MCF10A (ND: 44; OC: 14), MCF10AT1 (ND: 16; OC: 16), MCF10DCIS.com (ND: 38; OC: 24), and MCF10CA1a (ND: 29; OC: 31). (B) Representative confocal microscopy images of RFP-coexpressing cells of the four cell types in ND and OC conditions. The images include x-y (left) and x-z (right) views, with scale bar = 10 μm. The large volume reduction of MCF10DCIS.com cells is evident. (C) Plots showing changes in cortical stiffness (mean and SD) measured by Young’s modulus (Y) using a nanoindenter, displaying significant cell stiffening of MCF10DCIS.com cells due to cell crowding. The number of single cell stiffness measurements (technical replicates merged from two independent experimental repeats) was as follows: MCF10A (ND: 21; OC: 19), MCF10AT1 (ND: 19; OC: 14), MCF10DCIS.com (ND: 19; OC: 21), and MCF10CA1a (ND: 11; OC: 10). (D) Hyperosmotic conditions induced by PEG 300 treatment (light blue and darker blue circles for untreated and 2% PEG 300 = 74.4 mOsm/Kg, respectively; navy circles for 4% PEG 300 = 148.8 mOsm/kg) lead to dose-dependent cell volume reduction. The number of single cell volume analyses (technical replicates merged from three independent experimental repeats) was as follows: MCF10DCIS.com (ND control: 62; ND + 2% PEG: 23; ND + 4% PEG: 10). (E) Treatment with 2% PEG 300 (darker blue circles) for two days significantly increased the invasiveness (mean and SD) of MCF10DCIS.com cells, similar to the OC case (red circles). The number of cell invasion analyses (technical replicates merged from two independent experimental repeats) was as follows: MCF10DCIS.com (ND control: 13; ND + 4% PEG: 12; OC: 10). ****: p < 0.0001, ***: p < 0.001, **: p < 0.01, *: p < 0.05, ns: p > 0.05.

To validate the pro-invasive nature of cell volume reduction in MCF10DCIS.com cells, we examined the effect of hyperosmotic cell volume reduction in ND cells using polyethylene glycol (PEG) 300 for 15 minutes. The cell volume reduction effects of this treatment were previously described36,44. PEG 300 effectively reduced ND MCF10DCIS.com cell volume in a dose-dependent manner, showing a greater reduction with 148.8 mOsmol/kg PEG 300 than with 74.4 mOsmol/kg PEG 300 (Fig. 2D). To assess the impact of PEG-induced cell volume reduction on cell invasiveness, we used a collagen-crosslinked polyacrylamide hydrogel matrix-based invasion assay and exposed cells to 74.4 mOsmol/kg PEG 300 for 2 days. While treatment with 148.8 mOsmol/kg PEG 300 significantly reduced cell viability, cells treated with 74.4 mOsmol/kg PEG 300 remained viable for 2 days. Monitoring changes in cell invasiveness revealed that exposure of ND MCF10DCIS.com cells to 74.4 mOsmol/kg PEG 300 increased cell invasiveness (Fig. 2E), confirming the causal relationship between cell volume reduction and increased cell invasiveness.

Cell crowding induces TRPV4 relocation to plasma membrane in MCF10DCIS.com cells

Cell volume regulation typically depends on osmotic gradients that direct the movement of water across cell membranes45. This process is facilitated by ion flux modulation, which is controlled by ion channels and ion transporters located on the plasma membrane46–49. Notably, cells capable of achieving minimal cell volumes have been shown to successfully invade neighboring tissues38. Based on this, we speculated that the high cell volume plasticity observed in MCF10DCIS.com cells underlies their ability to penetrate the surrounding tissues. To achieve such efficient cell volume changes, cell crowding-induced mechanotransduction in MCF10DCIS.com cells may involve modulation of the number of ion channels or transporters on the plasma membrane.

To compare the plasma membrane-associated proteins between ND and OC conditions, we employed mass spectrometry and profiled those proteins using streptavidin-pulled surface-biotinylated cell lysates. Fig. 3A shows the relative protein densities on the plasma membrane under ND (blue bars) and OC (red bars) conditions in MCF10DCIS.com cells. The figure highlights the top 25 proteins that exhibit over a 5-fold increase in expression on the plasma membrane in the OC condition compared to the ND condition (triangle plots). The gene names of these 25 proteins are presented in Fig. 3A (Suppl. Table 1 provides the corresponding protein names along with the corresponding fold increases). Notably, TRPV4, a member of the transient receptor potential family of ion channels known for its mechanosensitive properties50,51, showed a remarkable 153-fold increase in plasma membrane association in response to cell crowding (Fig. 3A). Additionally, we observed an increase in plasma membrane association of other ion channels, such as SCN11A (the alpha subunit of the voltage-gated sodium channel Nav1.9) with ∼42-fold enrichment, and KCNN4 (the small-conductance calcium-activated potassium channel 3 SK3) showing ∼6-fold increase (Fig. 3A). To evaluate the relative plasma membrane association of proteins under OC versus ND conditions, we performed a comparable mass spectrometry analysis of cell-surface biotinylated cell lysates for MCF10A, MCF10AT1, and MCF10CA1a cells. The top 25 gene/protein names showing increased plasma membrane associations in these cells are listed in Suppl. Table 2. Unlike in MCF10DCIS.com cases, there was no cell-crowdinginduced relocation of ion channels and ion transporters to the plasma membrane in MCF10A and MCF10AT1 cells. However, in MCF10CA1a cells, ion transporters such as ATP2B4 showed a ∼37-fold greater plasma membrane association under crowding conditions (Suppl. Table 2). This suggests that plasma membrane relocations of these ion channels and ion transporters in response to cell crowding selectively occurred in MCF10DCIS.com cells and, to an extent, in invasive ductal cancer MCF10CA1a cells (Suppl. Table 2).

Cell crowding induces TRPV4 relocation to plasma membrane in MCF10DCIS.com cells.

(A) Mass spectrometry data showing proteins enriched in the plasma membrane (PM) >5-fold (fold changes represented using triangle plots; OC/ND ratio on the right axis) when cells are under OC (red bars) relative to ND conditions (blue bars). Ion channels are marked with red boxes, where TRPV4 shows about a 160-fold increased association with the plasma membrane under OC conditions. (B) Proteins near and on the plasma membrane were pulled down after cell surface biotinylation with streptavidin beads and immunoblotted for TRPV4. TRPV4 is significantly associated with the plasma membrane in OC MCF10DCIS.com cells. In MCF10CA1a cells, TRPV4 appears to be associated with the plasma membrane under both ND and OC conditions, with a slight increase under OC conditions. (C) Immunoblots of whole-protein lysates demonstrate similar overall TRPV4 protein levels across MCF10A cell derivatives, regardless of cell density. This indicates that the differing plasma membrane association of TRPV4 is due to trafficking changes, not expression level changes. GAPDH is used as a loading control. (D) Representative IF images by confocal microscopy show TRPV4 (red) localization compared to the control protein transferrin receptor (TfR; green) in MCF10A, MCF10AT1, MCF10DCIS.com, and MCF10CA1a cells under ND and OC conditions. DAPI (blue) staining was used for visualizing the nuclei. As observed in the biochemical data in (B-C), cell crowding induces the relocation of TRPV4 to the plasma membrane in MCF10DCIS.com cells. TRPV4 is associated with the plasma membrane in ND MCF10CA1a cells, with a clear elevated association in OC cells. Scale bar = 10 μm. (E) Plasma membrane-associated TRPV4 (%) is quantified for the four cell lines under ND and OC conditions by line analysis, showing a significant increase in both MCF10DCIS.com cells and MCF10CA1a cells due to cell crowding. The number of cells used for line analyses (technical replicates merged from three independent experimental repeats) was as follows: MCF10A (ND: 6; OC: 12), MCF10AT1 (ND: 6; OC: 11), MCF10DCIS.com (ND: 12; OC: 8), and MCF10CA1a (ND: 10; OC: 10). (F) IF images showing that hyperosmotic conditions induced by PEG 300 (74.4 mOsm/Kg) treatment also relocate TRPV4 (red) to the plasma membrane in MCF10DCIS.com cells. TfR localization remains consistent under hyperosmotic conditions. Increased relocation is also observed in MCF10CA1a cells. Scale bar = 10 μm. (G) The increased plasma membrane association of TRPV4 due to hyperosmotic stress is quantified by line analysis. The number of cells used for line analyses (technical replicates merged from two independent experimental repeats) was as follows: MCF10A (ND control: 6; ND + 4% PEG: 15), MCF10AT1 (ND control: 6; ND + 4% PEG: 9), MCF10DCIS.com (ND control: 12; ND + 4% PEG:8), and MCF10CA1a (ND control: 10; ND + 4% PEG: 21). Scale bar = 10 μm. (H) Representative regions of interest (ROIs) of TRPV4-stained immunohistochemistry (IHC) images in different pathology phenotypes. High-grade DCIS and invasive ductal cancer (IDC) ROIs clearly exhibit plasma membrane association of TRPV4. Two high-grade DCIS IHC images were acquired by two different people and both show plasma membrane-associated TRPV4. Scale bar = 20 μm. (I) Statistical results from independent histological evaluations of pathologies and TRPV4 distributions of 97 ROIs from 39 patient specimens indicate a high correlation (>70%) of plasma membrane association of TRPV4 with high-grade DCIS or IDC pathologies. Y/N: Yes/no, indicating both pathologists agreed that PM ion channels were present/absent. E: Equivocal, indicating the pathologists disagreed. Significantly high proportions of high-grade DCIS (75%) and IDC (73%) ROIs exhibited plasma membrane TRPV4 association, which was not observed in lower-risk cases. ****: p < 0.0001, ***: p < 0.001, **: p < 0.01, *: p < 0.05, ns: p > 0.05.

TRPV4 plays a pivotal role in facilitating the passage of calcium ions (Ca2+) and is integral in detecting various forms of mechanical stresses, including temperature fluctuations and osmotic pressure52,53. Activation of TRPV4 channels elevates intracellular calcium levels, leading to cell volume increase via osmotic water influx52,54. Nav1.9 primarily contributes to the generation and propagation of action potentials in sensory neurons and is associated with various pain disorders55–57. SK3 channels are activated by elevated intracellular calcium levels and serve diverse physiological functions58,59. However, unlike TRPV4, Nav1.9 and SK3 channels are not generally classified as mechanosensors. Consequently, we focused on TRPV4, the mechanosensor channel displaying the most significant increase in the plasma membrane association under cell crowding conditions, to further investigate its role in promoting pro-invasive cell volume reduction in MCF10DCIS.com cells.

To confirm the selective plasma membrane association of TRPV4 in OC-treated MCF10DCIS.com cells suggested by the mass spectrometry results, we performed immunoprecipitation using avidin beads on surface-biotinylated lysates from all four cell types under both control and OC conditions, followed by immunoblotting. The immunoblots showed a notable association of TRPV4 with the plasma membrane in OC MCF10DCIS.com cells, which was significantly lower in ND cells (Fig. 3B and Suppl. Fig. 5A). This association was not observed in MCF10A and MCF10AT1 cells under any conditions (Fig. 3B). Interestingly, in MCF10CA1a cells, TRPV4 associated with the plasma membrane under both conditions, with a slight increase in OC conditions (Fig. 3B). This increase suggests that these cells may also possess mechanosensing abilities, enabling them to respond to cell crowding. Additionally, immunoblots from whole-cell lysates showed consistent TRPV4 expression levels across different cell types and densities (Fig. 3C, Suppl. Fig. 5B), indicating that cell crowding influences ion channel trafficking without altering overall expression. This evidence strongly suggests that the relocation of TRPV4 to the plasma membrane is a mechanosensitive response to cell crowding.

To investigate the redistribution of TRPV4 in response to cell crowding, we performed IF imaging with confocal microscopy across all four cell types under both ND and OC conditions. The binding specificity of the TRPV4 antibody was validated in MCF10DCIS.com cells using TRPV4-specific shRNA, as demonstrated by IF imaging (Suppl. Fig. 6A) and immunoblotting (Suppl. Fig. 6B). As expected from the mass spectrometry results, TRPV4 (red) showed only modest localization to the plasma membrane in ND MCF10DCIS.com cells (Fig. 3D). In contrast, it was prominently associated with the plasma membrane in OC MCF10DCIS.com cells (Fig. 3D). In MCF10CA1a cells, TRPV4 was also associated with plasma membrane under ND conditions, but cell crowding also increased the association significantly as shown in MCF10DCIS.com cells (Fig. 3D). Plasma membrane associations were verified using the plasma membrane marker DiIC18(3), which we previously used to visualize the plasma membrane60. IF images showed that TRPV4 successfully colocalized with DiIC18(3) in OC MCF10DCIS.com and OC MCF10CA1a cells (Suppl. Fig. 7A). However, the non-interacting protein transferrin receptor (TfR; green) remained similarly distributed in the OC cells compared to the ND cells, indicating that the protein trafficking changes induced by cell crowding are specific to TRPV4 (Fig. 3D).

To investigate whether other ion channels identified through mass spectrometry and the wellknown mechanosensory channel PIEZO161,62, exhibit similar behavior, we conducted IF imaging on ND and OC MCF10DCIS.com cells targeting KCNN4 and PIEZO1. Unfortunately, due to the absence of specific antibodies, we were unable to examine SCN11. In ND MCF10DCIS.com cells, KCNN4 was observed to be mostly cytosolic, while PIEZO1 was mildly but slightly more associated with the plasma membrane than TRPV4 under ND conditions. However, under OC conditions, both PIEZO1 and KCNN4 also relocated to the plasma membrane, albeit to a lesser extent than TRPV4 (Suppl. Fig. 7B). The varying degrees of plasma membrane association of PIEZO1 and KCNN4 are quantified using line analysis, as summarized in Supple. Fig. 7C.

As expected, in MCF10A, MCF10AT1, and MCF10DCIS.com cells under ND conditions, TRPV4 was modestly present at the plasma membrane but predominantly localized intracellularly, with notable enrichment in the nuclei (Fig. 3D). Under OC conditions, TRPV4 remained largely cytosolic in MCF10A and MCF10AT1 cells but migrated out of the nuclei (Fig. 3D). In contrast, in OC MCF10DCIS.com cells, approximately 80% of TRPV4 relocated to the plasma membrane, as shown by line analysis (Fig. 3E). In MCF10CA1a cells, TRPV4 association with the plasma membrane was evident under both ND and OC conditions, with a further increase under OC, aligning with the immunoprecipitation results (Fig. 3B). Plasma membrane-associated TRPV4 was quantified using line analysis and plotted in Fig. 3E. Refer to Suppl. Fig. 8A for plots of the relative TRPV4 associations with the plasma membrane, cytosol, and nucleus between ND and OC conditions in all four cell types. These IF imaging results underscore the mechanosensitive plasma membrane relocation of ion channels in MCF10DCIS.com, and to an extent, in MCF10CA1a cells, contrasting with the observed insensitivity of MCF10A and MCF10AT1 cells to cell crowding.

To assess the relationship between cell crowding-induced cell volume reduction and plasma membrane relocation of TRPV4, we investigated whether hyperosmotic conditions, which reduce cell volume without cell crowding, would also result in TRPV4 relocation. We subjected ND cells to the same PEG 300 condition (74.4 mOsm/kg for 15 minutes) used in Fig. 2D to induce cell volume reduction. In ND MCF10DCIS.com cells, such hyperosmotic cell volume reduction prompted significant plasma membrane relocation of TRPV4, a response not observed in MCF10A and MCF10AT1 cells, but observed in MCF10CA1a cells (Fig. 3F). Quantitative line analysis confirmed these findings by assessing the relative TRPV4 association with the plasma membrane (Fig. 3G; Suppl. Fig. 8B for plots of the relative TRPV4 association with the plasma membrane, cytosol, or nucleus). We confirmed that the apparent plasma membrane-associated TRPV4 under the hyperosmotic condition was indeed localized at the plasma membrane by using an extracellular domain antibody against TRPV4 in live MCF10DCIS.com cells, both untreated and treated with 74.4 mOsm/Kg PEG 300. As expected, significant antibody binding was observed only in PEG 300-treated cells, where intracellular TRPV4 had relocated to the plasma membrane unlike in untreated live cells (Suppl. Fig. 9A-B). This finding underscores the relationship between mechanosensitive cell volume reduction and plasma membrane relocation of TRPV4, highlighting that both cell crowding and hyperosmotic stress lead to the same effect, which is particularly pronounced in MCF10DCIS.com cells.

We investigated whether TRPV4 relocation to the plasma membrane induced by cell crowding is reversible, as suggested by its impact on invasiveness (Suppl. Fig. 3F). To test this, previously OC MCF10DCIS.com cells were reseeded under ND conditions. We then assessed TRPV4 localization via IF imaging to determine if most channels returned to the cytoplasm with only a modest localization at the plasma membrane, and could be relocated to the plasma membrane under mechanical stress, such as hyperosmotic conditions, again. Consistent with their initial ND state, reseeded ND MCF10DCIS.com cells displayed intracellular TRPV4 distribution (Suppl. Fig. 10A). Upon exposure to hyperosmotic stress (74.4 mOsm/Kg PEG 300), TRPV4 was again relocated to the plasma membrane (Suppl. Fig. 10B). These findings, quantified through line analysis (Suppl. Fig. 10C), demonstrate that the mechanosensing response of MCF10DCIS.com cells is reversible.

Patient tissue analysis shows selective plasma membrane association of TRPV4 in high-grade DCIS cells

MCF10DCIS.com cells represent a basal cell model for high-grade DCIS cells driven by HRAS mutation26,63. However, the majority of patient-derived DCIS cells originate from the luminal cell population and lack the HRAS mutation63. Moreover, there are limited options for patient-derived DCIS cell line models that have not linked to well-defined pathology64,65. Thus, to seek generality of our finding regarding the selective association between plasma membrane relocation and high-grade DCIS pathology, we examined 39 breast cancer patient tissue blocks to assess the selectivity in vivo. We designed an early-stage retrospective study by selecting various breast tissue pathologies that range from benign to invasive cancer. Those pathologies include usual ductal hyperplasia, papilloma, columnar changes, ADH, low- to high-grade DCIS, and invasive ductal carcinoma (IDC). We also incorporated normal regions for comparison. Hematoxylin and eosin (H&E) staining was used to visualize tissue sections from each patient, excluding samples with prior cancer diagnoses or drug treatments. In total, 97 regions of interest (ROIs) from H&E-stained sections of 39 patient tissue blocks were selected. Two pathologists independently assessed TRPV4 distribution patterns at the single-cell level in corresponding ROIs in the immunohistochemistry (IHC) images (Fig. 3H). A detailed methodology is described in the Methods section and more example images are shown in Suppl. Fig. 11. High-grade DCIS cells, as depicted in Fig. 3H, clearly demonstrated plasma membrane-associated TRPV4 (Fig. 3H), a feature absent in lower-grade DCIS cells (intermediate- and low-grade). This distinction sets them apart not only from cells with ADH and benign pathologies but also from lower-grade DCIS cells. As demonstrated by the IF images in Fig. 3D for MCF10CA1a (invasive cell mimic), IDC cells exhibited notable plasma membrane TRPV4 (Fig. 3H), suggesting that they possess a similar pro-invasive mechanotransduction capability to high-grade DCIS cells.

We found that the sensitivity and specificity for plasma membrane TRPV4 in high-grade DCIS cells were 0.75 ± 0.19 (15/20) and 0.98 ± 0.03 (61/62) respectively (95% confidence interval) (Fig. 3I). It is important to note that the specificity calculation excluded IDC cases. Notably, even in less aggressive pathologies, a significant amount of TRPV4 was localized in the nuclei, as shown for MCF10A and MCF10AT1 cells in the IF images in Fig. 3D. This underscores our interpretation from the in vitro results in Fig. 3A-D that increased plasma membrane association of TRPV4 in high-grade DCIS cells results from changes in protein localization through trafficking alterations, rather than differences in expression levels. These initial in vivo results clearly demonstrate the selective and specific association of TRPV4 with the plasma membrane in high-grade DCIS cells. Notably, this association is absent in lower-grade DCIS, ADH, benign, and normal cells, thereby confirming our in vitro findings using MCF10A cell derivatives. These results suggest a potentially critical role for TRPV4 in the progression of high-grade DCIS.

Cell crowding inhibits ion channels and triggers their plasma membrane relocations

A calcium-permeant ion channel like TRPV4 is known to influence intracellular calcium dynamics54. While activation of TRPV4 typically elevates calcium levels and this potentially increases cell volume66–68, the impact of its inhibition is less clear, given the multifaceted nature of calcium signaling in cell volume control69. This complexity is compounded by compensatory cellular mechanisms and the involvement of other ion channels in response to altered calcium homeostasis46. Considering these factors, we hypothesized that cell crowding might inhibit calcium-permeant ion channels that are constitutively active at the plasma membrane, including TRPV4, which would then lower intracellular calcium levels and subsequently reduce cell volume via osmotic water movement. To test this, we employed the Fluo-4 calcium assay70 to compare relative intracellular calcium levels of MCF10DCIS.com cells between ND and confluent (Con) conditions through 4X confocal microscopy imaging with 488 nm excitation. For this calcium assay, we opted for confluent conditions instead of OC conditions to collect the Fluo-4 signal from confluent monolayers of MCF10DCIS cells, which demonstrates comparable cell-crowding induced TRPV4 relocation to the plasma membrane (Suppl. Fig. 12A) as observed in the OC condition. This choice was made because more than one layer of cells was occasionally observed in OC conditions, which would yield overestimation of crowding-dependent intracellular calcium levels. The intracellular Fluo-4 signal under Con conditions was significantly lower than that under ND conditions, as shown in the fluorescence images of ND versus Con cells in Fig. 4A. Time-dependent Fluo-4 images were acquired over 25 minutes. For ND cell images, the average Fluo-4 intensity values were calculated from 10-15 selected cells (highlighted in the white box in Fig. 4A). For Con cell images, intensity values were derived from all cells within the entire 50 μm by 50 μm field of view, as depicted in Fig. 4A. After background subtraction, the intensity values were plotted in Fig. 4B. The temporal Fluo-4 intensity profiles from both ND and Con cells remained largely constant over the measured 20-minute period. However, Con cells exhibited significantly lower Fluo-4 intensity compared to ND cells, indicating reduced intracellular calcium levels (Fig. 4C). Lower Fluo-4 intensity in Con cells than in ND cells was not due to limited Fluo-4 reagent in Con samples. Within cell clusters, cells experiencing crowding exhibited notably lower Fluo-4 levels compared to those at the periphery (Suppl. Fig. 12B), suggesting that cell crowding had resulted in a decrease in intracellular calcium levels, likely due to the mechanosensitive inhibition of calcium-permeable channels like TRPV4.

Cell crowding inhibits ion channels and triggers their plasma membrane relocations.

(A) To compare intracellular calcium (Ca²⁺) levels, we used a Fluo-4 AM assay, where green fluorescence intensity increases with higher intracellular Ca²⁺ levels. Calcium levels are significantly lower in confluent (Con) MCF10DCIS.com cells. (B) The temporal progression of averaged Fluo-4 intensity in ND MCF10DCIS.com cells (blue curve) in the box shown on the left image is compared with that of Con cells (red curve). Fluo-4 intensity is consistently lower in Con cells than in ND cells for approximately 25 minutes (200 ms acquisition time and 30 s time interval). (C) Fluo-4 intensity reduction due to cell crowding is significant in MCF10DCIS.com cells. 10 images were used for calculating average Fluo-4 intensities for both ND and Con cells. (D-H) Pharmacological inhibition of TRPV4 with 1 nM GSK219 generates dips in the Fluo-4 signal. Fluo-4 images at the baseline (t1) and the dip (t2) post 1 nM GSK219 are compared in MCF10DCIS.com cells between ND (D) and Con (F) conditions. The Fluo-4 intensity time traces are compared between ND (blue; E) and Con (red; G) conditions, showing that the magnitude of the dip (marked as ΔCa) is significantly lower in Con cells, where TRPV4 activity is largely inhibited under cell crowding conditions. Notably, ΔCa increased with higher GSK219 doses (1 nM vs. 0.2 nM), but remained significantly lower in Con MCF10DCIS.com cells, with smaller changes observed under the 0.2 nM GSK219 condition. The number of ΔCa measurements (H) (technical replicates merged from two independent experimental repeats) was follows: MCF10DCIS.com (ND + GSK219 Low: 9; ND + GSK219 High: 12; Con + GSK219 High: 10; Con + GSK219 Low: 14). (I-M) TRPV4 activation with 0.2 pM GSK101 leads to a small spike in ND cells (I, J). However, in Con cells, the same GSK101 treatment leads to a notably larger spike in Fluo-4 intensity, indicating that TRPV4 inhibition and subsequent relocation to the plasma membrane by cell crowding primes the ion channels for activation. GSK101 treatment also leads to a dose-dependent increase in the spike magnitude with a higher GSK101 concentration being strikingly high in Con MCF10DCIS.com cells (0.05 pM: 101L; 0.2 pM: 101H). The number of ΔCa measurements (M) (technical replicates merged from two independent experimental repeats) was follows: MCF10DCIS.com (ND + GSK101 Low: 7; ND + GSK101 High: 9; Con + GSK101 Low: 9; Con + GSK101 High: 9). (N-Q) TRPV4 activation status-dependent intracellular localization changes. (N) IF images of TRPV4 (red) and TfR (green) in ND MCF10DCIS.com cells show that GSK101 does not increase plasma membrane association of TRPV4. However, GSK219 significantly relocates TRPV4 to the plasma membrane in a dose-dependent manner (Supple. Fig. 13B for all dose cases), similar to ND cells treated with 74.4 mOsm/Kg PEG 300. (O) In OC cells, while GSK219 does not significantly alter TRPV4 association with the plasma membrane, GSK101 depletes plasma membrane TRPV4 in a dose-dependent manner (Supple. Fig. 13B for all dose cases), suggesting that TRPV4 activation status affects its trafficking. Relative plasma membrane associations with different treatments are quantified for ND (P) and OC (Q) cells using line analysis. The number of line analyses (P, Q) (technical replicates merged from two independent experimental repeats) was follows: ND and OC MCF10DCIS.com (control: 12 and 8; GSK219 Low: 6 and 7; GSK219 High: 6 and 13; GSK101 Low: 7 and 8; GSK101 High: 7 and 10; 2% PEG: 12 and 12). (R) The magnitudes of Fluo-4 spikes by GSK101 and dips by GSK219 show a linear relationship (R² ∼ 0.69), indicating a negative correlation between them. This reinforces the observation that TRPV4 inhibition increases its association with the plasma membrane, while activation shows the reverse effect. (S-V) Compared to the Fluo-4 intensity in control MCF10DCIS.com cells (S, T), shRNA showed similar baseline Fluo-4 levels (U, V). However, hyperosmotic stress by 74.4 mOsm/Kg PEG 300 (light gray box) led to a noticeable spike only in control ND cells. Additionally, cell crowding conditions (Con) led to a decreased Fluo-4 level (at t1 baseline in the image in S and red time trace in T); but a reduced in Fluo-4 level diference in shRNA-treated MCF10DCIS.com cells (t1; U) compared to control cases (S, T), as shown in the image and time trace (t1; V). (W) Relative Fluo-4 time-averaged intensities are plotted for individual control ND (blue) vs. Con (red) cells, and shRNA-treated ND (semi-transparent blue) vs. Con (semi-transparent red) cells. Intracellular calcium levels in shRNA ND cells are lower than those in control ND cells, reflecting the reduced number of TRPV4 channels. The decrease in calcium levels by crowding (Con) in shRNA cells is clearly lower than in control cells, reflecting the importance of TRPV4 in mechanosensing cell volume reduction. The number of Fluo-4 average measurements (technical replicates merged from two independent experimental repeats) was follows: MCF10DCIS.com control and TRPV4 shRNA groups (ND: 19 and 18; Con: 11 and 17). (X) PEG 300-induced calcium spikes are significantly lower in shRNA cells (semi-transparent gray) than in control cells (gray), reinforcing TRPV4’s crucial role in MCF10DCIS.com mechanotransduction. The number of ΔCa measurements under 2% PEG 300 condition was follows: ND MCF10DCIS.com (ND: 19; TRPV4shRNA: 16). (Y) TRPV4 silencing significantly reduced the mechanosensing cell volume reduction effect. Control ND cells underwent a 48% volume reduction in response to 74.4 mOsm/Kg PEG 300, whereas TRPV4-silenced cells reduced their volume by only 27%. The number of single cell volume measurements (technical replicates merged from two independent experimental repeats) was follows: ND and TRPV4shRNA treated MCF10DCIS.com (Control: 11 and 13; 2% PEG 8 and 11). ****: p < 0.0001, ***: p < 0.001, **: p < 0.01, *: p < 0.05, ns: p > 0.05.

To investigate whether TRPV4 inhibition was involved in decreased intracellular calcium levels and subsequent relocation of TRPV4 to the plasma membrane, we employed TRPV4-specific pharmacological agents to alter TRPV4 activity and modulate calcium concentrations in ND MCF10DCIS.com cells without applying cell crowding conditions. To inhibit TRPV4 activity, we used a TRPV4 inhibitor (GSK2193874; GSK219)71 at 0.2 and 1 nM, which we confirmed to have insignificant effects on cell viability in both ND and OC conditions for two days (Suppl. Fig. 13A). GSK219 treatments immediately caused a modest dip in the Fluo-4 signal -- or intracellular calcium level -- followed by a recovery, indicating cellular homeostatic activity. Fig. 4D shows Fluo-4 images taken before and at the signal dip after applying 1 nM GSK219, where the dip in the Fluo-4 signal is not visually apparent. However, the temporal profile of Fluo-4 intensity in Fig. 4E, which corresponds to the time points marked in Fig. 4D (t1: baseline and t2: dip), clearly shows the dip at t2, indicated by ΔCa (the vertical dashed line between the dip and the baseline). This modest Fluo-4 dip at t2 represents the inhibition of activity by GSK219 on a small population of constitutively active TRPV4 channels at the plasma membrane under ND conditions.

In Con cells, 1 nM GSK219 caused a smaller dip in Fluo-4 intensity compared to the one observed in ND cells, with no subsequent changes. This is likely due to fewer constitutively active TRPV4 at the plasma membrane in Con cells than in ND cells. Fig. 4F shows Fluo-4 images at the baseline and the dip, and Fig. 4G presents the temporal Fluo-4 intensity profile, highlighting the two time points. Indeed, the dip in Fluo-4 intensity induced by GSK219 was smaller for Con MCF10DCIS.com cells (mean ΔCa ∼ −39 ± 9) than for ND cells (mean ΔCa ∼ −49 ± 18), as illustrated in Fig. 4H. The dose-dependent (0.2 nM and 1 nM GSK219, labeled 219L and 219H) and cell density-dependent (ND and Con) Fluo-4 signal dips are summarized by the ΔCa values plotted in Fig. 4H, showing that the most significant TRPV4 inhibition occurred in ND MCF10DCIS.com cells treated with 1 nM GSK219 (219H). These findings suggest that a portion of TRPV4 channels that are constitutively active at the plasma membrane became inactive in response to cell crowding and that a significant fraction of TRPV4 channels relocated to the plasma membrane under cell crowding remained inactive.

Administering a TRPV4 activator (GSK1016790A; GSK101)72,73 immediately caused a notably greater spike in Fluo-4 signal in Con cells than in ND cells, evidencing that a large number of TRPV4 channels at the plasma membrane were inactive in Con cells. In Con cells, 0.2 pM GSK101 led to a large Fluo-4 signal spike with ΔCa ∼ +2,348 ± 597, while in ND cells, ΔCa was just ∼ +324 ± 130. This difference is depicted in the Fluo-4 images (Fig. 4I for ND and Fig. 4K for Con) and the time-dependent Fluo-4 intensity plots (Fig. 4J for ND and Fig. 4L for Con). The dose (101L and 101H representing 0.05 pM and 0.2 pM GSK101) and cell-density (ND and Con) dependent Fluo-4 signal spikes are summarized in the plotted ΔCa values (Fig. 4M). The results show the significant GSK101 dose-responsive activation of previously inactive TRPV4 in Con MCF10DCIS.com cells, suggesting that many inactive TRPV4 channels present at the plasma membrane were primed for activation.

Based on the differing effects of GSK219 and GSK101 on immediate intracellular calcium responses, along with lower baseline calcium levels in Con cells compared to ND cells, we hypothesized that cell crowding induces TRPV4 inhibition, leading to reduced intracellular calcium levels, which in turn triggers the relocation of TRPV4 to the plasma membrane as a compensatory mechanism. To test this hypothesis, we examined TRPV4 localization changes in response to GSK219 or GSK101 in a dose-dependent manner in both ND and OC MCF10DCIS.com cells using IF imaging (Fig. 4N for higher GSK101/219 concentrations; Suppl. Fig. 13B for both concentrations).

Indeed, treating ND cells with the TRPV4 inhibitor GSK219 for 1 hour resulted in significant dose- dependent TRPV4 relocation to the plasma membrane (∼53 ± 7% at 0.2 nM and ∼62 ± 7% at 1 nM), compared to control ND cells (∼15 ± 6%), as shown by line analysis (Fig. 4P). These results indicate that TRPV4 inhibition induces its strategic relocation to the plasma membrane, priming the channels for rapid activation to counteract the effects of mechanically-induced inhibition.

In contrast, in Con cells, where a greater number of inactive TRPV4 channels are likely present at the plasma membrane, GSK101 treatment led to a dose-dependent reduction in plasma membrane-associated TRPV4, likely due to internalization of the activated channels (Fig. 4O, 4Q). This finding is consistent with previous findings65. These data suggest that plasma membrane TRPV4 levels were largely regulated by the channel’s activity status. Specifically, channel activation led to the internalization of TRPV4, while channel inhibition promoted the relocation of TRPV4 to the plasma membrane. This relationship was approximately linear, as shown in Fig. 4R, with intracellular calcium changes inversely proportional to the plasma membrane fraction of TRPV4 (R² ≈ 0.69).

Contrary to our expectation, GSK219 treatment in Con cells slightly reduced TRPV4 levels at the plasma membrane (Fig. 4O, 4Q). Moreover, higher doses of GSK219 increased nuclear TRPV4 levels in Con cells without affecting cytoplasmic levels (Suppl. Fig. 13B). These findings suggest that further TRPV4 inhibition under crowding conditions triggers a distinct trafficking alteration. Recent studies have implicated nuclear TRPV4 in regulating nuclear Ca2+ homeostasis and Ca2+- regulated transcription74, and, in light of this study and our findings, TRPV4 may relocate to the nucleus as a compensatory mechanism to maintain nuclear calcium regulation. This relocation could reflect an adaptive response to preserve calcium-dependent transcriptional programs or other nuclear processes essential for cell survival under mechanical stress.

Given the role of TRPV4 in osmoregulation53,75, we investigated its significant relocation to the plasma membrane under hyperosmotic conditions. In ND MCF10DCIS.com cells, TRPV4 plasma membrane association increased from a baseline of ∼15% to ∼55% following 74.4 mOsm/kg PEG 300 treatment, as shown by line analysis (Fig. 3E, 4P). To explore whether TRPV4 inhibition contributed to this relocation, we conducted Fluo-4 assays under the same treatment conditions. The results revealed that osmotic water outflow induced an initial spike in intracellular calcium concentration, followed by a homeostatic relaxation aimed at restoring calcium levels (Suppl. Fig. 14A). This relaxation likely involved the inhibition of ion channels, including TRPV4, and their subsequent relocation to the plasma membrane.

Consistent with this hypothesis, we confirmed the plasma membrane relocation of other ion channels, namely KCNN4 and PIEZO1, following the same PEG 300 treatment (Suppl. Fig. 14B, showing IF images and line analysis results). These findings suggest that the increased TRPV4 association with the plasma membrane under hyperosmotic stress reflects an adaptive response to water outflow, driven by the inhibition and redistribution of ion channels to regulate osmotic balance.

To investigate the role of TRPV4 in mechanotransduction, we used shRNA to silence TRPV4 gene expression, resulting in a ∼50% reduction in TRPV4 protein levels (Suppl. Fig. 6B). We then assessed the effects of TRPV4 silencing on intracellular calcium levels in ND and Con conditions, as well as the immediate calcium response to hyperosmotic stress induced by 74.4 mOsm/Kg PEG 300. In ND cells, hyperosmotic stress triggered a significant calcium spike, as shown in the Fluo-4 image (middle; Fig. 4S) and time trace (blue; Fig. 4T). In contrast, TRPV4 silencing in ND cells led to a lower intracellular calcium level (left; Fig. 4U) and a reduced calcium spike in response to hyperosmotic stress (middle; Fig. 4U, and semi-transparent blue; Fig. 4V). Furthermore, the effect of TRPV4 silencing on intracellular calcium levels was less pronounced in Con cells (semi-transparent red; Fig. 4V) compared to the ND counterpart (semi-transparent blue; Fig. 4V). The relative intracellular calcium levels are summarized in Fig. 4W.

Notably, TRPV4 silencing significantly reduced the immediate calcium spike as summarized in Fig. 4X. This calcium spike was attributed to increased calcium concentration due to hyperosmotically reduced cell volume (Suppl. Fig. 14A). Thus, TRPV4 may play a critical role in mechanotransduction, enabling cells to sufficiently reduce their volume in response to mechanical stress (large volume plasticity). To further investigate this, we examined the effect of TRPV4 silencing on ND MCF10DCIS.com cell volume reduction in response to hyperosmotic stress. We indeed found that control ND cells underwent a 48% volume reduction in response to 74.4 mOsm/Kg PEG 300, whereas TRPV4-silenced cells only reduced their volume by 27% (Fig. 4Y). Notably, the reduced cell volume change observed in TRPV4-silenced cells under hyperosmotic conditions supports the idea that the observed calcium spikes are driven by cell volume reduction. This finding reinforces the concept that TRPV4 enhances cellular mechanosensitivity by facilitating volume changes and maintaining calcium homeostasis in response to mechanical stress.

TRPV4 inhibition drives cell volume reduction and increased invasiveness under cell crowding

To elucidate the relationship between TRPV4 inhibition, cell volume reduction, and increased invasiveness under cell crowding, we investigated the effects of TRPV4 activity status on these processes. Specifically, we evaluated how 2-day treatments with GSK101 and GSK219 impacted the cell volumes of ND and OC MCF10DCIS.com cells. This treatment duration was selected to align with the timeframe in which cell volume reduction and invasiveness changes are observed in response to cell crowding.

Under ND conditions, activating TRPV4 with GSK101 (0.05 or 0.2 pM for 2 days), which led to only modest calcium spikes and largely unaltered TRPV4 distribution (Fig. 4I, 4N; Suppl. Fig. 13B), did not significantly impact cell volume (Fig. 5A). Conversely, TRPV4 inhibition by GSK219 treatment (1 nM for 2 days) that induced greater calcium dips and a significant plasma membrane relocation of TRPV4 (Fig. 4D, 4N: Suppl. Fig 13B), led to a noticeable cell volume reduction (Fig. 5A). 0.2 nM GSK219 treatment had a negligible impact on ND cell volume (Fig. 5A). The GSK219 effect in ND MCF10DCIS.com cell volume was similar to that observed under hyperosmotic conditions by 74.7 mOsm/Kg PEG 300 (Fig. 5A).

Cell crowding-induced plasma membrane TRPV4 association scales with cell volume reduction and increases in invasiveness and motility

(A-C). MCF10DCIS.com cell volume changes with TRPV4 inhibition and activation. (A) In ND MCF10DCIS.com cells, TRPV4 agonist GSK101, which did not alter plasma membrane association of TRPV4, did not affect cell volume. Conversely, TRPV4 inhibitor GSK219, which increased plasma membrane association in a dose-dependent manner, reduced cell volume, with the effect of 1 nM GSK219 (219H) being similar to that of 74.4 mOsm/Kg (2%) PEG 300. (B) Under OC conditions, GSK101, which led to significant Fluo-4 spikes, increased cell volume in a dose-dependent manner, while GSK219 and PEG only mildly reduced cell volume. (C) Cell volume changes in MCF10DCIS.com cells show an inverse relationship (R² = 0.59) with plasma membrane association of TRPV4, reflecting the activation status of the channel. The number of single cell volume measurements (technical replicates merged from three independent experimental repeats): ND (A) and OC (B) MCF10DCIS.com cells (Control: 33 and 43; GSK101 0.05 pM: 19 and 15; GSK101 0.2 pM: 9 and 22; GSK219 0.1 nM: 10 and 36; GSK101 1 nM: 15 and 9; 2% PEG 300: 23 and 8). (D-F) Cell invasiveness increases with greater cell volume reduction and plasma membrane association of TRPV4. (D) Cell invasiveness significantly increased with higher GSK219 concentrations under ND conditions. (E) GSK101 under OC conditions caused a notable decrease in cell invasiveness in a dose-dependent manner. (F-G) Plasma membrane association of TRPV4 predictably reports cell invasiveness (R² ∼ 0.69; F), while cell invasiveness and cell volume are inversely related (R² ∼ 0.69; G), reinforcing our observation that cell volume reduction promotes cell invasiveness. The number of invasive cell fraction measurements (technical replicates merged from two independent experimental repeats) : ND (D) and OC (E) MCF10DCIS.com cells (Control: 6 and 4; GSK101 0.05 pM: 4 and 4; GSK101 0.2 pM: 4 and 4; GSK219 0.1 nM: 4 and 4; GSK101 1 nM: 4 and 4; 2% PEG 300: 4 and 7). (H-M) To assess if cell motility also follows the trend of cell invasiveness, we performed a single-cell motility assay by tracking nuclear WGA in individual live cells every 60 s for 25 min. (H) Representative trajectories of individual cells were color-coded to reflect displacement at each time interval. Compared to untreated ND cells, 0.2 pM GSK101 treatment slowed overall cell diffusion, while 1 nM GSK219 and 74.4 mOsm/Kg PEG 300 treatments increased cell diffusion. ShRNA TRPV4 (Sh-ctrl) increased cell motility under ND conditions. However, with TRPV4 depletion, treatment with 74.4 mOsm/Kg PEG 300 failed to increase cell diffusivity (D) in shRNA-treated cells (Sh-PEG), unlike in the untreatec cells. Scale bar = 200 μm. Using single-cell analysis, we quantified cell diffusivity (D) and speed (v; movement directionality). (I) GSK101 treatment significantly reduced D, while GSK219 and PEG 300 notably increased it. shRNA TRPV4 also increased ND cell D, but PEG treatment did not change D in the shRNA-treated cells. (J) GSK101, GSK219, PEG 300, and shRNA treatments increased v, with GSK101 causing the most significant increase. The directionality of shRNA-treated ND cells was unaffected by PEG treatment. The number of single-cell motility measurements of MCF10DCIS.com cells (technical replicates merged from two independent experimental repeats): Control: 81; GSK101: 100; GSK219: 100, PEG: 100; shRNA: 102; shRNA + PEG: 104. (K) Like cell invasiveness, cell motility (D) positively scales with plasma membrane association of TRPV4 (R² ∼ 0.73). (L) Cell motility (D) inversely relates to cell volume (R² ∼ 0.89). (M) Cell motility (D) and cell invasiveness show a strong linear relationship (R² ∼ 0.85), enabling the use of cell motility measurements to assess overall cell invasiveness. ****: p < 0.0001, ***: p < 0.001, **: p < 0.01, *: p < 0.05, ns: p > 0.05.

Under Con conditions, GSK101 triggered significant dose-dependent calcium spikes, leading to corresponding increases in cell volume (Fig. 5B). In contrast, GSK219 under Con conditions had minimal effects on intracellular calcium levels and did not alter TRPV4 distribution at the plasma membrane, resulting in a slight cell volume reduction at 1 nM (Fig. 5B). Similarly, PEG 300 caused a modest decrease in cell volume, comparable to the effect of 1 nM GSK219 (Fig. 5B). As expected, plasma membrane-associated TRPV4 (% PM TRPV4) showed a significant inverse linear correlation with cell volume (R² ≈ 0.59) (Fig. 5C). These findings (Fig. 5A-B) indicate that increased TRPV4 inhibition and the subsequent reduction in intracellular calcium levels in response to cell crowding are associated with decreased cell volume in MCF10DCIS.com cells.

We investigated the connections between TRPV4 activity, plasma membrane association, cell volume changes, and cell invasiveness. Using a collagen-crosslinked polyacrylamide hydrogel matrix-based invasion assay, we evaluated changes in invasiveness after 2-day treatments with GSK101 or GSK219. In ND MCF10DCIS.com cells, 0.2 pM GSK101 mildly suppressed cell invasiveness, while 0.05 pM GSK101 caused no effect (Fig. 5D). In contrast, GSK219 increased the invasive cell fraction in a dose-dependent manner under ND conditions, mirroring changes in Fluo-4 signals and plasma membrane-associated TRPV4 (Fig. 5D). Under OC conditions, GSK101 increased cell volume and suppressed cell invasiveness dose-dependently (Fig. 5E). Conversely, GSK219 and PEG 300, which slightly reduced cell volume, increased cell invasiveness (Fig. 5E). Representative invasion images are shown in Suppl. Fig. 15A. These results establish a clear relationship between TRPV4 activity, plasma membrane-associated TRPV4, cell volume, and cell invasiveness. Notably, we observed a significant positive linear correlation between plasma membrane-associated TRPV4 and cell invasiveness (Fig. 5F; R² ∼ 0.69) and an inverse relationship between cell volume and cell invasiveness (Fig. 5G; R² ∼ 0.89).

Next, we investigated cell-crowding induced cell motility changes in MCF10DCIS.com cells, since invasiveness encompasses both the ability to penetrate and migrate through surrounding tissues. As expected from the increased cortical stiffness of OC MCF10DCIS.com cells (Fig. 2C), we observed that OC MCF10DCIS.com cells underwent cytoskeletal rearrangements, including the increased formation of stress fibers (Suppl. Fig. 15B), which likely enhanced cell motility75. To assess how TRPV4 inhibition and activation, which reduced and increased cell volume, respectively, affected the motility of ND MCF10DCIS.com cells and the role of these motility changes on cell invasiveness, we conducted single-cell tracking experiments similar to single molecule tracking methods60,76–80. We utilized wheat germ agglutinin (WGA) as a nuclear marker, which stained the cell nucleus within 1 hour of incubation with cells. We then analyzed the diffusivity (D) and directionality (v) of single cell movements. We tracked overall single-cell trajectories for 3 hr with 1 min intervals, marking cell displacement over each interval in different colors, ranging from navy for the slowest to red for the fastest (Fig. 5H). Consistent with the trend in cell invasiveness, TRPV4 activation with 1 pM GSK101 significantly reduced single-cell movements, while inhibition with 0.2 nM GSK219 and 74.4 mOsm/kg PEG 300 notably increased these movements.

Using the single-cell motility assay, we investigated how TRPV4 enhances cell invasiveness through a mechanotransduction mechanism involving cell-volume reduction. Reducing TRPV4 expression (∼50%) via shRNA diminished the cell volume reduction effect of mechanosensing (Fig. 4Y). We predicted that shRNA would also attenuate the increased cell motility effect of mechanosensing in MCF10DCIS.com cells. Additionally, since reduced TRPV4 in ND cells lowered intracellular calcium level as shown in Fig. 4W, shRNA effect would follow the effect of GSK219. Indeed, shRNA (Sh-ctrl; Fig. 5H, 5I) slightly increased both D and v of MCF10DCIS.com cell movement, mimicking the effects of GSK219. However, the shRNA disabled the mechanotransduction capability of the cells given that there was no increase in D and v induced by hyperosmotic stress using 74.4 mOsm/Kg PEG 300 (Sh-PEG; Fig. 5H, 5I, 5I). This result reinforces the critical role of TRPV4 in increasing cell invasiveness through mechanosensing activity inhibition-induced cell volume reduction.

Interestingly, while GSK219 and hyperosmotic conditions promoted increased cell movement (D; Fig. 5I), they also reduced cell directionality (v; Fig. 5J). This reduction in directionality suggests that while the cells become more mobile, their movements are less targeted, potentially indicating a more random or exploratory behavior under mechanical stress. Remarkably, cell movement increased in proportion to the amount of TRPV4 at the plasma membrane (Fig. 5K; R² ∼ 0.73) and decreased with greater cell volume (Fig. 5L; R² ∼ 0.89). Cell movement and invasiveness showed a strong linear correlation (Fig. 5M; R² ∼ 0.85), suggesting that these behaviors are governed by the same underlying mechanisms. Thus, our motility assay can inform cell invasiveness at the single-cell level.

Mechanosensitive TRPV4 relocation to plasma membrane indicates enhanced invasiveness

We investigated whether the mechanosensitive relocation of TRPV4 to the plasma membrane can reliably predict a cell’s ability to undergo pro-invasive mechanosensitive cell volume reduction. To achieve this, we used changes in cell motility as markers for overall shifts in cell invasiveness, given the strong correlation observed between cell motility and invasiveness (Fig. 5M). Our hypothesis is that cells with such mechanotransduction capability, like MCF10DCIS.com cells, should show increased plasma membrane association of TRPV4 when inhibited with GSK219 or under hyperosmotic conditions, evidenced by increased cell motility (D) while demonstrating the opposite effect with GSK101 treatment.

For comparison, we employed three uncharacterized cell types: MDA-MB-231, a triple-negative invasive breast cancer cell line, and two recently developed patient-derived DCIS cell lines, ETCC-006 and ETCC-01064,65, which have not yet been histologically classified by grade. These cells were compared with positive control groups, including MCF10DCIS.com and MCF10CA1a, which were observed to possess pro-invasive mechanotransduction capability (Fig. 5D, 6A-6C). MCF10AT1 served as a negative control, as it lacks this capability (Fig. 6D-6F).

Pro-invasive cell volume mechanotransduction pathway indicators.

(A) In MCF10CA1a cells, plasma membrane relocation of TRPV4 was induced by a 15-minute treatment with 2% PEG 300 (74.4 mOsm/kg) or by OC conditions. IF images show a largely intracellular distribution of TRPV4 (red) in ND control cells, whereas cells treated with PEG 300 or subjected to OC exhibit a significant increase in plasma membrane-associated TRPV4. Nuclei are stained with DAPI (blue). Scale bars in all panels of Fig. 6 represent 10 μm. (B) Line analysis quantifying plasma membrane-associated TRPV4 (%) reveals significant increases following a 1-hour treatment with GSK219 (1 nM), a 15-minute exposure to 74.4 mOsm/kg PEG 300, or under OC conditions in MCF10CA1a cells. In contrast, no significant increase in plasma membrane TRPV4 is observed with GSK101 treatment. (C) In MCF10CA1a cells, cell movement diffusivity (D) increased following GSK219 or PEG 300 treatments, whereas it decreased with GSK101. Conversely, movement directionality (v) increased significantly with GSK101 (0.2 pM) but remained unchanged with GSK219 or PEG treatments. (D-O) No plasma membrane relocation of TRPV4 was observed in response to inhibition by GSK219, hyperosmotic stress induced by PEG 300, or cell crowding (OC) in MCF10AT1 (D, E, F), MDA-MB-231 (G, H, I), ETCC-006 (J, K, L), and ETCC-010 (M, N, O) cells. Similarly, GSK219 or PEG 300 did not increase single-cell motility in these cells. This is demonstrated by IF images (TRPV4: red; DAPI: blue) (D, G, J, M), line analysis results for plasma membrane-associated TRPV4 (E, H, K, N), and single-cell motility analyses for diffusivity (D) and directionality (v) (F, I, L, O). Notably, none of these cell lines showed motility changes in response to PEG 300 treatment. However, responses to TRPV4 activation (GSK101) and inhibition (GSK219) varied across cell types, suggesting distinct roles of TRPV4 in their cancer biology. In MCF10AT1 cells (F), GSK219 significantly reduced diffusivity (D), while no other treatment affected D or v. In MDA-MB-231 cells (I), neither D nor v was altered by any treatment, indicating that TRPV4 has an insignificant role in their motility. Both ETCC-006 and ETCC-010 cells exhibited increased diffusivity with GSK101; however, GSK219 also increased diffusivity in ETCC-006 cells (L), while having no effect on ETCC-010 cells (O). Directionality (v) increased with GSK101 in ETCC-006 cells (L), whereas ETCC-010 cells showed no change in v across all conditions (O). The number of line analyses for plasma membrane-associated TRPV4 under ND control, ND + 0.2 pM GSK101, ND + 1 nM GSK219, ND + 2% PEG 300, and OC conditions (technical replicates merged from three independent experimental repeats) were: MCF10CA1a (B): 11, 6, 8, 12, 10; MCF10AT1 (E): 13, 9, 8, 9, 19; MDA-MB-231 (H): 13, 9, 8, 9, 19; ETCC-006 (K): 12, 10, 10, 5, 10; and ETCC-010 (N): 5, 5, 7, 5, 6. The number of single-cell motility analyses under ND control, ND + 0.2 pM GSK101, ND + 1 nM GSK219, and ND + 2% PEG 300 conditions were: MCF10CA1a (C): 100, 100, 100, 100; MCF10AT1 (F): 130, 161, 582, 183; MDA-MB-231 (I): 57, 6, 21, 442; ETCC-006 (L): 65, 66, 24, 100; and ETCC-010 (O): 317, 1136, 43, 71. (P) Plasma membrane association of TRPV4 (% PM TRPV4) scaled positively with cell diffusivity (D) over a broader range in MCF10DCIS.com cells compared to MCF10CA1a cells, consistent with the higher cell volume plasticity observed in MCF10DCIS.com cells. This finding suggests that both cell types engage a pro-invasive mechanotransduction pathway. (Q) In contrast, this scaling relationship is absent in MCF10AT1, MDA-MB-231, ETCC-006, and ETCC-010 cells, indicating a lack of the mechanotransduction response. (R) The presence of this pathway in MCF10CA1a and MCF10DCIS.com cells is further supported by the observed >2-fold increase in TRPV4 plasma membrane association (x-axis; PM TRPV4_peg/PM TRPV4_ctrl) and >1-fold increase in diffusivity (y-axis; Dpeg/Dctrl) following PEG-300 treatment. (S) The cell volume reduction-driven mechanotransduction pathway is further demonstrated by plotting PEG-300-induced changes in TRPV4 plasma membrane association (x-axis; PM TRPV4_peg/PM TRPV4_ctrl) against the diffusivity ratio with GSK219 versus GSK101 (y-axis), where both cell types show a significantly greater than 2-fold increase, highlighting the activation of this pathway in MCF10CA1a and MCF10DCIS.com cells.****: p < 0.0001, ***: p < 0.001, **: p < 0.01, *: p < 0.05, ns: p > 0.05.

MCF10CA1a cells showed plasma membrane relocation of TRPV4 in response to TRPV4 inhibition (1 nM GSK219; ND 219H), PEG 300 (74.4 mOsm/kg; ND PEG300), and OC conditions. In contrast, treatment with 0.2 pM GSK101 did not affect TRPV4 plasma membrane association (Supplementary Fig. 16 and Fig. 6A, showing IF images for TRPV4 (red) and DAPI (cyan)). This response was similar to that observed in MCF10DCIS.com cells under OC conditions. The plasma membrane associations of TRPV4 were quantified by line analysis, showing an increase under mechanical stresses (PEG and OC) and TRPV4 inhibition with GSK219 (Fig. 6B). As expected, based on the differences in plasma membrane TRPV4 association, MCF10CA1a cells exhibited increased motility (D) in response to TRPV4 inhibition (GSK219) and hyperosmotic conditions (PEG) (Fig. 6C). Additionally, their movement became more directional (v) with TRPV4 activation (GSK101) (Fig. 6C), similar to the observations in MCF10DCIS.com cells (Fig. 5I, 5J).

In contrast, the negative control MCF10AT1 cells did not exhibit TRPV4 inhibition-dependent plasma membrane relocation of TRPV4, as shown by IF images (Suppl. Fig. 16, Fig. 6D) and line analysis results (Fig. 6E). In these cells, TRPV4 inhibition (GSK219) decreased diffusivity (D), while PEG 300 had no effect. Moreover, movement directionality (v) remained unchanged under all treatment conditions (Fig. 6F).

The three test group cells, MDA-MB-231 (Fig. 6G, 6H), ETCC-006 (Fig. 6J, 6K), and ETCC-010 (Fig. 6M, 6N), failed to relocate TRPV4 to the plasma membrane in response to mechanical stresses (PEG and OC) or TRPV4 inhibition (GSK219) (Suppl. Fig. 16), as shown by IF images and quantification. This suggests that these cells lack the pro-invasive mechanosensitive cell volume reduction capability observed in MCF10DCIS.com cells. Consistently, none of these cells showed increased motility (D) or directionality (v) under hyperosmotic conditions or TRPV4 inhibition (Fig. 6I, 6L, 6O).

We plotted these results to show cell diffusivity (D) scaling with varying plasma membrane associations of TRPV4, where MCF10DCIS.com (green hexagon) and MCF10CA1a (red hexagon) exhibited a positive scaling (Fig. 6P). However, the changes in plasma membrane TRPV4 occurred over a narrower range (31 – 79%) for MCF10CA1a compared to MCF10DCIS.com cells (16 - 75%), reflecting the larger mechanosensitive cell volume plasticity of MCF10DCIS.com cells. In contrast, MDA-MB-231, ETCC-006, and ETCC-010 cells did not show this scaling (Fig. 6Q). Like MCF10AT1 cells, plasma membrane TRPV4 associations in these cells remained at similar percentages (Fig. 6Q).

This dichotomy in the presence of a pro-invasive mechanotransduction program is evident in the plots of diffusivity (D) increase versus plasma membrane TRPV4 relocation (Fig. 6R, 6S), where only MCF10CA1a and MCF10DCIS.com cells show a significant increase in both parameters. Furthermore, the relationship between plasma membrane-associated TRPV4 and D increase with GSK219 versus GSK101 can be used to assess the presence of a pro-invasive, mechanosensitive cell volume reduction program. Most non-mechanotransducing cells remained close to the baseline value of 1, while mechanotransducing cells showed a significant increase in D with GSK219 treatment (Fig. 6S).

Discussion

Our study explored how cell crowding, a common condition in diseases and development, impacts cell invasiveness. We employed novel assays to quantify cell invasion and single-cell motility in cells with breast tissue pathologies, including DCIS and ADH, which experience cell crowding due to abnormal proliferation in confined ducts. Our results revealed a mechanotransduction pathway activated by crowding selectively in high-grade DCIS cells, establishing a direct link between crowding and invasiveness. This pathway includes ion channel inhibition, intracellular calcium reduction, compensatory channel relocation to the plasma membrane for priming them for later activation, cell volume reduction and cell stiffening, and increased invasiveness and motility (Fig. 7). Notably, the compensatory relocation of ion channels to the plasma membrane serves as a reliable indicator of an active mechanotransduction pathway. This mechanotransduction capability that enables pro-invasive cell volume reduction distinguishes high-grade DCIS cells from normal, hyperplastic, or lower-grade DCIS cells, supporting previous findings that the ability of cells to reach minimal cell volume promotes efficient invasion38.

Cell crowding activates a pro-invasive mechanotransduction pathway in high-grade DCIS cells but not in less aggressive or normal cells.

This cell-crowding-induced pro-invasive pathway involves a cascade of events, including ion channel inhibition, intracellular calcium reduction, cell volume reduction and cortical stiffening, and increased cell invasiveness and motility. In high-grade DCIS cells, calcium-permeable ion channels such as TRPV4 relocate to the plasma membrane upon inhibition, compensating for reduced intracellular calcium levels by priming the channels for later activation under mechanical stress. The pro-invasive mechanotransduction pathway is selectively triggered by cell crowding or hyperosmotic stress in high-grade DCIS cells, which exhibit significant TRPV4 plasma membrane relocation, pronounced cell volume reduction, and increased motility and invasiveness. In contrast, less aggressive or normal cells remain significantly less or non-responsive to these stimuli. Notably, MCF10DCIS.com cells exhibit greater cell volume reduction compared to other cells, likely due to their larger baseline cell volume at normal density, demonstrating their high cell volume plasticity that correlates with crowding-induced invasiveness. The extent of TRPV4 plasma membrane relocation, cell volume reduction, and increased invasiveness and motility scales with each other, where the increased TRPV4 association with the plasma membrane can robustly serve as a marker of pro-invasive mechanotransduction activation. This mechanotransduction capability sets high-grade DCIS cells apart from less aggressive cells, providing a critical criterion that may help identify high-risk cells with invasive potential. This mechanotransduction capability was also validated in patient specimens, suggesting its relevance in clinical settings. Future investigations will include utilizing TRPV4 localization patterns as a diagnostic tool to assist in pathological grading and as a prognostic marker to identify high-risk DCIS cells likely to activate this pathway under mechanical stress.

Using mass spectrometry, we identified TRPV4 as a crucial component of this pathway, providing insights into how non-invasive breast cancer cells may evolve into invasive phenotypes under mechanical stress. Through Fluo-4 assays and TRPV4 activators and inhibitors, we found that cell crowding triggered TRPV4 inhibition, leading to reduced intracellular calcium and subsequent cell volume reduction in MCF10DCIS.com cells. Notably, TRPV4 inhibition prompted its relocation to the plasma membrane, priming it for later activation under suitable signaling conditions to maintain cellular homeostasis. TRPV4 inhibition by cell crowding increased invasiveness and motility, while TRPV4 activation had the opposite effect. Reducing TRPV4 protein expression by 50% via shRNA diminished both calcium decrease and volume reduction under mechanical stress, such as cell crowding and hyperosmotic conditions. These TRPV4-knockdown cells did not exhibit increased motility under hyperosmotic conditions, demonstrating TRPV4’s critical role in mechanotransduction-induced invasiveness through lowering cell volume plasticity. Invasive MCF10CA1a cells showed similar but modest mechanotransduction capability. In contrast, other invasive breast cancer cell lines, such as MDA-MB-231 cells, exhibited opposing responses to TRPV4 inhibition and activation, highlighting the heterogeneity of TRPV4 roles in invasive cancer progression81–83.

Our discovery of a selective mechanotransduction pathway in high-grade DCIS cells was further validated through the analysis of patient-derived breast cancer tissues. Plasma membrane association of TRPV4 was predominantly observed in high-grade DCIS and IDC lesions but was absent in lower-grade DCIS and less aggressive pathologies. This selective relocation highlights TRPV4’s critical role in the pro-invasive mechanotransduction pathway unique to high-grade DCIS cells and certain IDC subsets. These findings suggest that plasma membrane-associated TRPV4 may serve as a valuable biomarker for identifying high-grade DCIS lesions.

Our data indicate that ion channels beyond TRPV4 may contribute to pro-invasive mechanotransduction. Notably, SCN11A and KCNN4 showed increased plasma membrane association under cell crowding conditions, suggesting their roles in modulating ion flux and cell volume and thus contributing to the mechanotransduction response. Although PIEZO1 is known to respond to mechanical stimuli62, its lesser plasma membrane relocation compared to TRPV4 under cell crowding and hyperosmotic conditions suggests a lesser role in promoting invasiveness in this case. Further comparisons are needed to determine the dominant ion channel driving mechanotransduction in high-grade DCIS cells. However, it is evident that cell crowding-induced pro-invasive cell volume reduction pathway involves the coordinated action of multiple ion channels to significantly reduce intracellular calcium levels27,84.

We made the novel discovery that hyperosmotic conditions, which reduce cell volume through osmotic water efflux, mimic the effects of cell crowding by inducing similar pro-invasive cell volume reduction. Treatment with hyperosmotic agents such as PEG 300 caused TRPV4 to relocate to the plasma membrane and increased cell invasiveness, underscoring TRPV4’s role as a key mechanosensor responding to external mechanical stresses in this pathway. These findings suggest that the convergence of different mechanical stresses, including cell crowding and hyperosmotic stress, serves as a critical trigger for pro-invasive cell volume reduction pathways, activating a unified mechanotransduction signaling cascade.

The study’s insights have far-reaching implications, unlocking new research paths in cancer and biological sciences. By elucidating mechanosensitive responses to cell crowding, we can shed light on tissue development, wound healing, and physiological processes involving cell density changes. The parallels between cell crowding and hyperosmotic conditions in driving pro-invasive behaviors warrant further investigation into the intricate mechanisms of cytoskeleton reorganization, ion channel and transporter relocation, and enhanced cell invasiveness and motility. Importantly, our findings illuminate the mechanotransduction capability of high-grade DCIS cells that differentiate them from less aggressive cells. This understanding paves the way for developing diagnostic and prognostic strategies that leverage the selective intracellular localization patterns of TRPV4 and other mechanosensitive ion channels, ultimately guiding clinical decisions for patients with high-risk DCIS and other cancers.

Materials and methods

Cell lines and treatments

MCF10A cells were purchased from ATCC, and MCF10AT1, MCF10DCIS.com, and MCF10CA1a cells were obtained from the Barbara Ann Karmanos Cancer Institute through a material transfer agreement. ETCC-006 and ETCC-010 cells were purchased from the Leibniz Institute DSMZ, and MDA-MB-231 cells were from ATCC. All cell lines were authenticated prior to purchase, used at a low passage number (<15), and confirmed to be mycoplasma-free by PCR analysis. MCF10A and MCF10AT1 cells were maintained in DMEM/F12 medium supplemented with 5% horse serum, 20 ng/mL EGF, 0.5 μg/mL hydrocortisone, 10 μg/mL insulin, and 1% penicillin and streptomycin. Additionally, MCF10A cells were supplemented with 100 ng/mL cholera toxin. MCF10DCIS.com and MCF10CA1a cells were cultured in DMEM/F12 with 5% horse serum and 1% penicillin and streptomycin. ETCC-006 and ETCC-010 cells were maintained in RPMI 1640 supplemented with 10% fetal bovine serum (FBS) and 1% penicillin and streptomycin. MDA-MB-231 cells were cultured in DMEM/F12 with 10% FBS and 1% penicillin and streptomycin. All cells were cultured at 37°C in 5% CO₂. For TRPV4 modulation, we used the TRPV4 activator GSK1016790A (GSK101; Tocris 64T33) and the TRPV4 inhibitor GSK2193874 (GSK219; Tocris 5106).

For volume measurements, parental cells were mixed with those expressing RFP at a 9:1 ratio to distinguish individual cells. Cells were treated with 2% (74.4 mOsmol/kg) or 4% PEG 300 (148.8 mOsmol/kg; v/v; Millipore Sigma 8074845000) in culture medium for 15 min, 2% PEG-00 for 48 h, GSK1016790A (Tocris; 0.05 and 0.2 pM), or GSK2193874 (Tocris; 0.2 and 1 nM) for 48 h.

TRPV4 knockdown by shRNA

Cells in growth medium without penicillin and streptomycin were transfected with the hTRPV4 shRNA plasmid (1 and 2 µg; Santa Cruz sc-61726-SH) and Lipofectamine3000 in opti-MEM. At 6 h after transfection, the medium was replaced, and cells were cultured for 30 h before processing for western blot and immunofluorescence analyses.

Immunofluorescence

The cells were fixed with 4% paraformaldehyde (Fisher Scientific) for 20 min, followed by permeabilization with 0.1% saponin (Fisher Scientific) for 5 min at room temperature (RT). After permeabilization, the cells were washed three times in PBS and blocked in 0.1% saponin + 10% BSA in PBS at RT for 1 h. Cells were incubated with primary antibodies overnight at 4 °C in 0.1% saponin in PBS. The primary antibodies used were TRPV4 (Abcam 39260; different lot numbers were tested and selected for optimal performance; 1:500 dilution), TfR (ThermoFisher Scientific 13-6800; 1:500 dilution), PIEZO1 (Alomone APC-087; 1:500), and KCNN4 (Alomone ALM-051; 1:500). Cells were washed three times in PBS, incubated with fluorescent-tagged anti-rabbit-Alexa 555 or anti-mouse-Alexa 488 secondary antibodies (Thermo Fisher Scientific) and DAPI for 1 h at RT, imaged by confocal microscopy. For plasma membrane staining, DiIC18(3) (1,1’-Dioctadecyl-3,3,3’,3’-Tetramethylindocarbocyanine Perchlorate; DiIC18(3); Thermo Fisher Scientific, Cat # D282) was diluted in warm PBS to 5 μM. Media was removed from the cells, then the 5 μM DiIC18(3) in PBS was added. The dish was transferred to the incubator (37 C, 5% CO2) for 10 min. Following the 10 min incubation, the DiIC18(3) in PBS was removed and the cells were washed 2x with complete cell growth media. The cells were then fixed with 2% PFA in PBS.

Confocal imaging and image processing

2D images were acquired using a Yokogawa spinning-disk confocal microscope (Andor Technology) installed in a Nikon Eclipse TE2000 inverted microscope using a 60x/1.49NA Plan Apo objective (Nikon). The samples were illuminated using 430, 488, 561, and 647 nm solid-state lasers (Andor Technology). Images were acquired using an iXon back-illuminated EMCCD camera (Andor Technology). For volume measurements, 3D confocal images were acquired using z-step sizes calculated based on Nyquist conditions. Images were processed using ImageJ (NIH) or Imaris (Bitplane) software. For live cell imaging, cells were maintained in 5% CO2 at 37 °C in a stage-top incubator (Oko-lab).

Western blotting

ND cells (40–60% cell density) were lysed in SDS sample buffer, and OC cells were lysed in cytoskeleton buffer (10 mM Tris pH 7.4,100 mM NaCl, 1 mM EDTA, 1 mM EGTA, 1% Triton X-100, 10% glycerol, 0.1% SDS, 0.5% deoxycholate). SDS sample buffer (Bio-Rad) supplemented with reducing reagent was added to lysates and boiled at 100°C for 10 min. The samples were separated on 4–15% SDS-PAGE gels and transferred to nitrocellulose membranes (Bio-Rad). Membranes were blocked in TBST (TBS + 0.1% Tween-20) containing 5% non-fat milk or 10% BSA and then incubated with primary TRPV4 antibody (ThermoFisher Scientific PA541066; 1:1000) and GAPDH (Santa Cruz sc-32233; 1:2000) overnight at 4 °C. Membranes were washed three times in TBST (5 min each) and then incubated with horseradish peroxidase (HRP)-conjugated anti-rabbit (ThermoFisher Scientific A28177) or anti-mouse (ThermoFisher Scientific A28177) secondary antibodies for 30 min at RT. Blots were visualized using West Femto maximum sensitivity chemiluminescent substrate (Thermo Fisher Scientific).

Lentiviral infection and stable cell generation

To produce lentiviruses encoding RFP, HEK-293 cells were transfected with the lentiviral vector plasmid DNA using Lenti-X Packaging Single Shots (Takara). pCSII-EF-miRFP670v1-hGem(1/110) was a gift from Vladislav Verkhusha (Addgene plasmid # 80006; http://n2t.net/addgene:80006; RRID:Addgene_80006). Supernatants containing lentiviral pseudoparticles were harvested 24 and 48 h post-transfection. Harvested lentiviral particles were immediately stored at −80 °C. To establish stable cell lines, 70% confluent cells at one day post-seeding were infected with lentivirus in the presence of 8 µg/mL polybrene (Sigma-Aldrich). Two days after transduction, 1 µg/mL puromycin (Thermo Fisher Scientific) was added to the medium to select for stably transduced cells. The samples were visualized daily to ensure that the untransduced cells in the wells were not viable. Once the polyclonal populations had sufficiently expanded, cell stocks were prepared and harvested for protein expression assays.

Cell invasion assay

For cell invasion assays, 8-well chamber slides (Nunc Lab-Tek) were coated with 50 µg/mL poly-L-lysine (Fisher Scientific) for 30 min and washed with PBS. The slides were fixed with 0.5% glutaraldehyde (VWR) for 20 min and washed. Gelatin was conjugated with fluorescein by mixing 200 µL fluorescein-gelatin (1 mg/mL; ThemoFisher Scientific) with 800 μL unlabeled gelatin (4 mg/mL) (Sigma-Aldrich), followed by incubation for 5 min at 60 °C. The gelatine mix was allowed to cool at RT for 5 min and then applied, and the slides were incubated for 15 min. Slides were washed with PBS and disinfected with 70% ethanol for 30 min. Following three PBS washes, the residual reactive groups were quenched with growth medium by incubating at room temperature for 30 min in the dark. Cells were seeded in a fresh growth medium and incubated undisturbed on a horizontal surface at RT for 30 min to encourage cell distribution. Chamber slides were incubated with 5% CO2 at 37 °C. To detect cell invasion, cells were fixed with 4% formaldehyde and stained with DAPI (Thermo Fisher Scientific), and imaging was performed using a 4x/0.2 NA Plan Apo objective (Nikon). Cell invasion was determined by quantifying the sites of the degraded matrix, which were visible as dark areas in the bright-green fluorescent gelatin matrix. The area of gelatin digestion and the number of cells were calculated using the ImageJ software.

Nanoindenter assay for measuring cortical stiffness

Stiffness measurements were performed on live cells in 5% CO2 at 37 °C, maintained in a stage-top environmental chamber (Oko-lab) using a nanoindenter (Optics11 Chiaro) attached to the confocal microscope. Nanoindentation was performed at the cell surface of single cells with an indentation probe with a spring constant (0.24 N/m) and tip diameter (10 μm). The Hertzian contact model was used to fit the data to extract Young’s modulus in the elastic regime.

Surface biotinylation and pull-down assay

ND cells that were 40%–70% confluent and OC cells were detached and washed twice with PBS. The cells were resuspended at a concentration of 25 × 106 cells/mL in PBS containing 2 mM sulfo-NHS-biotin (Invitrogen). The reaction mixture was then incubated at RT for 60 min. Cells were washed three times with 1 M Tris pH 8.0 to quench and remove the excess biotin reagent. Cell pellets were lysed in 1.0 mL of cold lysis buffer (50 mM HEPES pH 7.2, 150 mM NaCl, 1.0% Triton X-100, 1.0% CHAPS, 100x protease and phosphatase inhibitors). Lysates were incubated on ice for 30 min, transferred to tubes, and centrifuged (10,000 × g for 5 min) to remove insoluble material. Supernatants were collected, and protein concentrations were measured using the bicinchoninic acid assay (Pierce). We washed 50 µL resin (ThermoFisher Scientific) with immobilized streptavidin twice with binding buffer (0.1 M phosphate, 0.15 M NaCl pH 7.2) and centrifuged at 5000 x g for 1 min. Biotin-labelled cell lysate (1 mg) was added to the resin and incubated with rotation for 1 h at RT. The resin was washed by resuspending in binding buffer, centrifuging to pellet the resin, and removing the supernatant by aspiration. The wash was repeated four times. Samples were boiled in SDS-PAGE sample buffer with DTT and separated by electrophoresis.

Mass spectrometry

Peptide mixtures from each sample were analyzed by LC-MS/MS using a nano-LC system (Easy nLC1000) connected to a Q Exactive HF mass spectrometer (Thermo Fisher Scientific). The platform was configured with a nano-electrospray ion source (Easy-Spray, ThermoFisher Scientific), an Acclaim PepMap 100 C18 nanoViper trap column (3 μm particle size, 75 μm ID × 20 mm length), and an EASY-Spray C18 analytical column (2 μm particle size, 75 μm ID × 500 mm length). Peptides were eluted at a flow rate of 300 nL/min using linear gradients of 5–25% acetonitrile (in aqueous phase and 0.1% formic acid) for 40 min, followed by 45% for 10 min, and static flow at 90% for 10 min. Mass spectrometry data were collected in a data-dependent manner, switching between one full-scan MS mode (m/z:380–1400; resolution: 60 K; AGC:3e6; maximum ion time: 20 ms), and 10 MS/MS scans (resolution: 15 K; AGC:1e5; maximum ion time: 120 ms; nCE:27) of the top 10 target ions. Ions were sequenced once and dynamically excluded from the list for 20 s. The datasets were searched using MaxQuant at default parameters against the UniProt Human Proteome database.

Line analysis

IF images of ion channels (TRPV4, PIEZO1, or KCNN4), TfR, DiIC18(3), and DAPI were opened in ImageJ. These images were concatenated so that a line crossing a cell in all three images could be used to measure approximate percentage protein localization in the plasma membrane, cytosol, or nucleus. Background was subtracted from all images. DiIC18(3) signal was used as a guide for plasma membrane location (between 50% of the peak intensity) and DAPI signal (between 50% of the peak intensity) for nuclear locations. The cytosol location was defined as the region between the 50% peak intensity points of the DiIC18(3) and DAPI signals. The ion channel signal in each window along the line was then averaged. The percentage of protein localization in the plasma membrane, cytosol, or nucleus for a cell was calculated by dividing the average intensity of each location by the sum of intensities in all three locations.

Live cell detection of plasma membrane associated TRPV4

Cells were seeded in a LabTek II 8-well Chambered Coverglass dish (Thermo Fisher 155409) at a density of 10,000 cells per well and allowed to adhere and grow for 2 days. Afterward, the cells were incubated on ice for 1 h with 1:300 Anti-TRPV4 (extracellular) rabbit primary antibody (Alomone ACC-124) in either complete media or 74 mOsm/Kg PEG 300 in complete media. Following the incubation, the cells were washed twice with the respective media (complete media or 74 mOsm/Kg PEG 300 in complete media) while remaining on ice. Subsequently, the cells were stained for 1 h at 22 °C with 1:5,000 Donkey anti-Rabbit IgG H+L Highly Cross-Adsorbed Secondary Antibody, Alexa Fluor™ 647 (Thermo Fisher A31573) in the same respective media. After the staining period, the cells were imaged.

Pipeline for analyses of patient histologic samples

Specimen collection: A retrospective clinical study was conducted to collect patient tissue blocks under IRB approval from George Washington University (IRB# NCR203065), with patient consent waived. Patient personal information was not disclosed to the researchers, ensuring that sample collection and analysis remained blinded. A pathologist chose 39 patient breast formalin-fixed paraffin-embedded (FFPE) tissue specimens, representing a combination of the following pathologies: benign, ADH, IDC, low-, intermediate-, or high-grade DCIS, and IDC, along with normal tissue from each patient. Specimens associated with a past cancer diagnosis or drug treatment history were excluded.

Sample preparation for histopathologic evaluation: For each FFPE block, two serial dissections were carried out. Tissue sections on unlabeled slides were stained with hematoxylin and eosin (H&E) and subjected to immunohistochemical (IHC) staining using anti-TRPV4 antibodies (Abcam 39260; 1:100 dilution). Whole-slide images were captured using the Olympus VS200 whole-slide imaging system. To validate the binding specificity of the antibody, both negative and positive control samples were prepared. IHC staining was performed by VitroVivo Biotech (Rockville, MD).

Two sequential annotations: A breast surgical pathologist annotated the pathology stages within specified regions of interest (ROIs) on wholeslide H&E images, based on cell morphological features. Two pathologists independently evaluated the same ROIs in the corresponding IHC images. They annotated the protein distribution using a three-tier classification:

Case 1: Absence of TRPV4.

Case 2: Intracellular TRPV4 localization.

Case 3: Presence of TRPV4 in the plasma membrane, with or without intracellular TRPV4.

The independent annotations concerning protein localization by the two pathologists were then subjected to statistical tests for selectivity and specificity. Any IHC ROIs with classification disagreements between the pathologists were designated as equivocal cases.

Calcium reporter assay

This assay was completed using the Fluo-4 Direct™ Calcium Assay Kit (Thermo Fisher F10471), including the assay buffer (Thermo Fisher), 2X Fluo-4 calcium assay reagent (Thermo Fisher), and preweighed, water-soluble probenecid (Thermo Fisher P36400; 2.5 mM). The 2X Fluo-4 calcium assay reagent was first prepared by adding 10 mL of the Fluo-4 calcium assay buffer and 200 μL of 250 mM probenecid to the desiccated calcium assay reagent at room temperature (21 °C). This mixture was then vortexed and allowed to sit for 15 min. While waiting for the 2X Fluo-4 calcium assay reagent to finish being prepared, the media from the well of a LabTek II Chambered Coverglass with Cover #1.5 Borosilicate Sterile 8-well plate (Thermo Fisher 155409) was removed and replaced with 200 μL of complete media. Once the 2X Fluo-4 calcium assay reagent was fully prepared, 200 μL of the reagent was added to the well with 200 μL of complete media. The 8-well plate was then placed at 37 °C and 5% CO₂ for 30 min. Following 30 min at 37 °C and 5% CO₂, the 8-well plate was removed from the incubator and placed at room temperature for 30 min. After incubation in the reagent for 30 min at room temperature, the 8-well plate was imaged using the confocal microscope with a 488 nm laser for 30 min with 30-second intervals between images and an exposure time of 200 ms. If the sample was treated, the sample was imaged for 35 min with 30-second intervals, with the treatment being added after 5 minhad elapsed. Added treatments all had a final volume of 200 μL and were composed of: 1 nM GSK 2 in 1:1 media:2X Fluo-4 calcium assay reagent, 0.2 nM GSK 2 in 1:1 media:2X Fluo-4 calcium assay reagent, 0.2 pM GSK 1 in 1:1 media:2X Fluo-4 calcium assay reagent, 1:1:1 water:media:2X Fluo-4 calcium assay reagent, and 0.3 Osm/L PEG 300 in 1:1 media:2X Fluo-4 calcium assay reagent.

Cell motility assay

Cells were first seeded in MatTek 35 mm dishes with No. 1.5 coverslip, 14 mm glass diameter, and Poly-D-Lysine coated glass bottoms (MatTek P35GC-1.5-14-C). Once the cells adhered, they were stained with 10 μg/mL WGA-488 (Thermo Fisher W11261) for 10 min at room temperature. Following the staining, the cells were washed with complete media twice and then incubated at 37 °C and 5% CO₂ for 1 h. During this incubation, the cells received the following treatments: 0.2 pM GSK 101 in complete media for 1 h, 0.05 pM GSK 101 in complete media for 1 h, 1 nM GSK 219 in complete media for 1 h, 0.2 nM GSK 219 in complete media for 1 h, 74.4 mOsm/Kg PEG 300 in complete media for 15 min, and 1:1 hypoosmotic solution in complete media for 15 min. After the incubation, the sample was placed in an OkoLab chamber at 37 °C and 5% CO₂ for imaging. Imaging was completed using a confocal microscope with a 488 nm laser for 3 h with 1-minute intervals.

Cell viability assay

MCF10DCIS.com cells were plated in a LabTek II 8-well Chambered Coverglass dish (Thermo Fisher 155409) at a seeding density of 10,000 cells per well. Once the cells in the well reached the desired number of days post-confluence, they were stained using Propidium Iodide Ready Flow Reagent (Thermo Fisher R37169). Specifically, 1 drop of propidium iodide (PI) was added to 500 µL of cell growth medium per well and incubated for 15 min at room temperature (22°C). After staining, the cells were washed twice with complete growth medium. Next, the cells were stained with 1 µg/mL WGA-488 (Thermo Fisher w11261) in complete growth medium for 15 min at 37°C in a 5% CO₂ atmosphere. Following the incubation with WGA-488, the cells were washed three times with complete growth medium and then incubated for 2 h at 37°C in 5% CO₂ to allow WGA migration into the nuclei. Finally, the cells were imaged using confocal microscopy at 4X magnification, with laser excitation at 488 nm for WGA-488 and 560 nm for propidium iodide.